---
title: "Intro to R"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**CCDL 2019**

### This [R Markdown](http://rmarkdown.rstudio.com) Notebook takes you through the basics of R  

*Resources for learning R* 

  - [Swirl, an interactive tutorial](https://swirlstats.com/)  
  - [R for Data Science](https://r4ds.had.co.nz/)  
  - [Tutorial on R, RStudio and R Markdown](https://ismayc.github.io/rbasics-book/)  
  - [Handy R cheatsheets](https://www.rstudio.com/resources/cheatsheets/)  
  - [R Notebooks tutorial](https://bookdown.org/yihui/rmarkdown/)  

#### Inserting and running code chunks in R Markdown

When you execute code within the notebook, the results appear beneath the code. 
Try executing this chunk by clicking the *Run* button within the chunk or by 
placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

```{r}
print("You ran a code chunk!")
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by 
pressing *Cmd+Option+I* (Windows *Ctrl + Alt + I*).

Try to put the following line into its own code chunk and run it:

print("You made a code chunk!")

## The Basics  

R remembers what you tell it and stores assigned values in it's *environment*. 
You can think of an R *environment* as a box that you can put objects in for R to 
store. 
In RStudio, you will see what R has in it's *environment* in the *environment* 
tab in the upper right corner.
Going with this analogy then, you can only references things in R that are in the 
*environment*. 
Here let's assign the value of `1` to `x` using the assignment operator: `<-`.
Let's subsequently print out `x` and see what output we get. 

```{r Basic assignment}
x <- 1
# Print out x:
x
```

As you may notice from above, R will ignore what you say (allowing you to make 
comments on your code) if you put hashtags/pound signs at the start of the line.

```{r Comments}
# This is important because you should always have comments to explain your code
```

What do you suppose will happen if we ask for R to print `y`? 
(Hint: see what's listed in the *environment* tab)

```{r error=TRUE}
y
```

This is R's way of telling us that it doesn't know what y is because it's not 
in it's *environment*. 
  
In summary: If there is no `<-` in your statement, this means it will **not** 
show up in your environment. 

Base R (R without any extra frills from *packages*) knows how to do arithmetic. 
Feel free to try out division, exponents, and multiplication.   
*Hint:* `/`  `^`  `*` 

```{r Arithmetic}
# Addition:
x + 5
```

## Data Types 

*Why are data types important?*  
Data *types* relate to the kind of data in an object: What is it made up of?
Not all questions or transformations make sense, nor do all R commands. 
*Functions* are things that take input, do something with this input, and then
give you output. 
But often these *functions* are built to work with specific *data types*. 
You can just search for a function by name in the *help bar* in RStudio in the
lower right corner.
A function's documentation in the *help* tab should tell you what data *type* it
will expect as input.

Let's explore different data types in R: 
To find out what data type `x` is, use the function `str`.
Note you can also see some information about `x` in the *environment* tab in the
upper right. 

```{r Data structures}
# Ask R to tell you what type of data structure "x" is
str(x)
```

So `x` is currently numeric (abbreviated `num`). 
*Numeric* data types refer to numbers like `1.3333` or `1`.  
Now let's turn `x` into something else to explore our next data type.
*Integer* data, as you would expect, is whole numbers that aren't decimals.

```{r Characters}
# Let's make x something else
x <- "x"

# What do you suppose this will say now? 
str(x)
```

By using quotes, we've made `x` a *character* variable. 
*Character* data types have quotes around them; usually strings of characters.
Note that quotes make it a character even if it's a number in the quote: 

```{r Characters continued}
x <- "1"
str(x)
```

Let's say we change our minds and we don't want `x` to be known as `1` in our 
environment anymore, we can use the function `rm`.

```{r Removal}
# Remove x from the environment
rm(x)
```

After running the above step, take a look at our environment tab. Is x there? 
Let's bring x back so we can try to get rid of it's assignment a different way:

```{r}
# Reassign x
x <- "1"

# Make x NULL
x <- NULL
```

Take a look at our environment. It's technically still there, but it's says
`NULL`. This means it's there, but empty.

Let's store a bunch of things together in different types of *vectors*: 
*Vectors* are one dimensional and contain multiple pieces of datum (more on that 
later).

Let's make a numeric *vector*.   
Also sidenote that the best object names are ones that intuitively make sense to 
people.
This way people who review your code can more easily understand what's going on
(more on this later).
Colons tell R to take all the numbers in between. 

```{r Vectors}
# We'll name this vector "nums" so it's more clear what this object is. 
nums <- 1:5

# What do you think this will give us now?: 
str(nums)
```

Let's make a *character* vector.
The `c` function allows us to *concatenate* ie. put multiple things together in 
one vector.
We'll make a character vector and call this object `words`.

```{r Non-numeric vector}
# Let's store these words as a vector
words <- c("apple", "boy", "cat", "dog", "elephant")

# What do you think this will give us now?:
str(words)
```

Let's make a *factor* vector. 
If we want R to treat the strings that are the same as the same value such as if
we had the vector below where `blue` and `red` are repeated, we would need to 
make this into a `factor.` 

```{r}
# Let's make a color vector
color_vector <- c("red", "blue", "red", "yellow", "blue")
```
  
Let's take a look at our vector `color_vector` and what it looks like.

```{r}
str(color_vector)
```
Oh no! We wanted to make a factor vector but R thinks that we want a character
vector. That's because R will, by default, assume this is a character string. 

But what's the difference anyway?...
  
Let's take a look at our object another way using the function `summary`.
Notice that if we use `summary` on a this character vector, it doesn't give us
terribly meaningful information about how many times each color is repeated 
because R sees this as a character string and doesn't know to look at the data
like we want it to. 

```{r Summary}
summary(color_vector)
```

If we want `color_vector` to be a factor we need to tell R to treat it as such. 
If you want to try to tell R to reinterpret something as a `factor`, you need to 
use the function `as.factor`
*Note* that every data type has it's own `as.___` function (eg. `as.numeric`, 
`as.character`, etc.).

```{r Coerce Factor}
# Let's turn our character vector into a factor
color_vector <- as.factor(color_vector)

# Character strings that are the same will be included in the same level
# What do you think this will give us now that we've turned this into a factor?
summary(color_vector)
```

*Logical* vectors are ones that can help you find out what data meets 
different criteria.
You can create logical vectors from using *logical statements* on other vectors.
`0` means `FALSE` and `1` means `TRUE`
  
We've already learned   `<-` is for assigning objects, here are other R operators
that can be used for logical expressions.  
  `==` : Is this equal? Note that this is different interpretation than a single
         `=` which is generally used for arguments in functions or also R will 
         interpret as an assignment(`<-`).  
  `!=` : Is this not equal? An exclamation point reverses the previous statement.
  `>` : Is this greater than?  
  `<` : Is this lesser than?  
  `>=` : Is this greater than or equal to?  
  `<=` : Is this lesser than or equal to?  
  
For example, if we want to see which of the words in our `words` vector are 
the words `"dog"`, we'll construct a statement: 

```{r Logical color_vector}
# First let's take another look at what `words` looks like: 
words
```

```{r}
# Now let's determine which of the items in `words` are equal to "dog"
words == "dog"
```

This has printed out a vector of the same length as "words" but with `TRUE` or 
`FALSE` at each position in the vector. 
For numeric values, we can use greater than or equal signs. 

```{r Logical nums}
# First let's take another look at what `nums` looks like: 
nums
```

```{r}
# Now let's determine which of the items in `nums` are greater than or equal to 2
nums >= 2
```

Note that when it comes to logical statements, factors can be referenced by their
character strings:

```{r Logical factor 2}
# Let's remind ourselves of what `color_vector` looks like: 
color_vector
```

```{r}
# Let's determine which of these colors are 'red'
color_vector != "red"
```

...or by their numeric level if you coerce the factor to numeric with 
`as.numeric`:

```{r Logical factor 1}
# Let's check out what levels are assigned to our colors with `levels` function
levels(color_vector)
```

Notice that levels are automatically assigned in alphabetical order. 
Let's turn color_vector into numeric data and see what numbers are assigned
to each level.

```{r Make factor into num}
# Let's transform 
color_vector_num <- as.numeric(color_vector)

# Let's take a look at what numbers were given to each level by printing the number and factor 
# versions together:
color_vector
color_vector_num
```

Based on what we've seen with the level order above, what does this give us?:

```{r}
# Now we can use greater than or lesser than with this object:
color_vector_num > 1
```

*To review, here are the R data types we learned*:  
  - *Numeric* : Numbers like 1.3333 or 1  
  - *Integer* : Numbers without decimals   
  - *Character* : Things with quotes around them; strings of characters  
  - *Factor* : Sets of variables that R will assign an factor levels to  
  - *Logical* : An object where 0 means FALSE and 1 means TRUE  

## Data Structures  

Data *structures* refer to the sizes and dimensions of the objects. 

*Example:* You may have already noticed, but `str()` reports the dimensions of 
an object so, for a vector that has 10 pieces dimensions will be shown as 
`[1:10]`.

```{r Vectors again}
one_through_ten <- 1:10
str(one_through_ten)
```

A matrix is a two dimensional collection of data. Like in math, commas separate
rows and columns indices. 
We will generally be using matrices when we are working with only the numbers of
a gene expression matrix.
eg. for a vector that has 10 rows and 10 columns `str()` will report dimensions 
of `[1:10, 1:10]`

```{r Matrix}
# Note that we need to use arguments for the matrix function to tell R how many 
# rows "nrow = 10" and how many columns "ncol = 10" we want. 
sq_matrix <- matrix(1:100, nrow = 10, ncol = 10)
str(sq_matrix)
```

```{r}
# In RStudio, you can also view an object by clicking on it in the Environment 
# tab OR by using `View()`
View(sq_matrix)
```

What dimensions do you suppose this matrix will have by default if we don't 
specify the number of rows and columns we want? 
To check out this dimensions of this object, let's use the `dim()` function.

```{r Matrix cont.}
not_sq_matrix <- matrix(1:100)
dim(not_sq_matrix)
```

Matrices however, cannot store different types of data. 
Notice that in the above examples we are only using integers. 
If you try to mix data types together in a matrix, R will coerce them to be all 
one type, which will probably be bad for you because your information will be 
distorted.   
  
Hence the `data.frame` is probably what you will be using for a lot of your data. 
Working with `data.frame`s also is handy because Hadley Wickham's `tidyverse` tools 
are mostly built around using `data.frame`s or its' closely related sibling the 
`tibble`. The next Rmd will show you all about manipulating `data.frame`s with the 
tidyverse tools. 

```{r Data frames}
# Let's make a data frame from our previously created vectors!
df <- data.frame(nums, words, color_vector)
```

Note that if these vectors were not conveniently all the same length, we could 
not put them in a `data.frame` together.  

```{r}
str(df)
# Also note that "obs" are rows and "variables" are the columns in a data.frame
```

Also note that the `data.frame` function makes strings a factor by default,
unless you tell it not to. 
So let's remake this `data.frame`, but tell R not to do that with one of the 
arguments for the `data.frame` function. 

```{r Data frames cont. }
# Let's make a data frame from our previously created vectors but tell R not to
# turn strings into factors!
df <- data.frame(nums, words, color_vector, stringsAsFactors = FALSE)
str(df)
# Note that with the argument `stringsAsFactors` now our words variable remains 
# a character vector as it was before. 
```

Now that we have a nice `data.frame` and other objects to play with in our 
environment; let's play around with it and extract pieces of it. 

```{r Subsetting Indices}
# Let's say we only want to see what's in the first row and first column of this
# data.frame. We can subset data structures by using brackets. 
df[1, 1]
```

We can do the same sort of thing to a matrix:

```{r Subsetting Indices 2}
# Note that you can do this same thing to matrices: 
sq_matrix[3, 3]
```

We can also select multiple columns at once. Either by using a colon:  

```{r}
sq_matrix[3, 1:3]
```

Or by using `c()`

```{r}
sq_matrix[c(1, 3), ]
```

Also note that by not putting a column index above, all the columns were selected. 

There are also built in objects to preview matrices or data.frames.   
-`head()` will give you the first 6 rows of data, by default.   
-`tail()` will give you the last 6 rows of data.  

```{r}
head(sq_matrix)
```

If you want more or fewer rows, you can specify this by putting an `n = ` argument.

```{r}
tail(df, n = 3)
```

Let's subset a vector we created previously.

```{r}
one_through_ten[8]
```
Why did we not use a comma to subset this object? 
Hint: Check it's structure with `str()` if you forgot what this object looks like. 

What if we want to subset the vector to everything except a certain index? 
That's where negative indexing comes in handy. 

```{r}
# Let's get rid of the item in the second spot using a negative sign
words[-2]

# Now let's print out the original for comparison
words
```

Data frames have column and row names. You can also specially reference 
`data.frame`s' variables by their column names. 

```{r Special data.frame things}
# data.frames store column names and row names 
df$nums
# We'll do fancier data.frame manipulations in the next Rmd so hold tight!
```

What if we want to store all of our objects in one big object? 
We can't do that with data frames because not all of these are the same length 
or dimensions. 
This is where `lists` come in handy.
Lists do not care about data type or data structure, but let you store whatever
you want in one big object.

```{r}
# Let's remind ourselves of what objects we've made so far using the `ls` function 
# Note you can also look in your environment tab panel to see this info.
ls()
```

Now let's make one crazy list with various objects of different dimensions and
types using the function `list`!

```{r Lists}
variety_list <- list(df, sq_matrix, color_vector, x)

# Let's take a look at this list 
str(variety_list)
```

How do we extract things from this weird list? We need to use brackets again. 

```{r Single bracket}
# This will show us our data.frame in the first spot as a singular item
first_item <- variety_list[1]

# Let's take a look at this, note that we've just made this a list of one now
str(first_item)
```

You'll find that the single bracket is probably not giving you what you want 
here. 
To explain this, let's look at a couple other things about how lists work. 
The structure of lists are kind of analogous to those 
[Russian nesting dolls](https://en.wikipedia.org/wiki/Matryoshka_doll) in that 
you can have lists inside of lists and you can think of them as layers.
In this analogy the single bracket is not actually extracting the `data.frame`
that is in this list. 

```{r Double Bracket}
# This however will give us the data.frame in this spot in the list. 
variety_list[[1]]
```

Let's take a look at the structure of this:

```{r Double Bracket str}
# Let's look at the structure of this after we've 'extracted' the data.frame
# from the list with the double brackets. 
str(variety_list[[1]])
```

To further clarify the difference between the double and single bracket, note 
that with a single bracket we can't use the `data.frame` as usual.
R tells us `NULL` if we try to reference the column `$color_vector` because it 
doesn't recognize it as a `data.frame`, but instead thinks of it as a `list` item.
But, with the double bracket we can play with our `data.frame` the same way as before: 

```{r Subset list and df}
# Single bracket: can't use the data.frame
variety_list[1]$color_vector
```

Notice the above doesn't give us the column we were looking for. 
Instead we need to use the double brackets like below:

```{r Subset list and df 2}
# Double bracket: can use the data.frame
variety_list[[1]]$color_vector
```

*To review, here are the data structures we learned:*  
  - A *Vector* is one dimensional collection of data that is all of the same type  
  - A *Matrix* a two dimensional storage of data that is all of the same type  
  - A *Dataframe* is a two dimensional storage of data that can be a mix of data
    types  
  - A *List* is multidimensional storage of data that can be a mix of data types 
    and structures (more explanation to come with )

## About good scientific coding practices

#### 1) Style guides help people read your code

Just like how incorrect punctuation and grammar can be distracting for a reader
to grasp your message in your writing, code that doesn't follow a style is 
difficult for others to understand and be able to use.
Your code is not useful if it isn't easily readable, which is why naming 
conventions, code style, and consistency are important. 

We suggest following a style guide like one of these:  
- [Hadley Wickham's R Style Guide](http://adv-r.had.co.nz/Style.html)  
- [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml).   
  
#### 2) `set.seed` helps people reproduce your results

When performing any kind of analyses that use random sampling, or something that
may vary each time you re-run it.

```{r}
# How to set the seed: 
# 1) Put any number as your argument for the function and run set.seed like below
# 2) Run your analyses like normal
# 3) If you run your analyses again, they give exactly the same numbers. 
set.seed(54321)
```

Setting the seed makes it so what ever results you get, are reproducible *exactly*.

#### 3) R Notebooks are helpful for documenting your science

As we've seen in this notebook, R Markdown notebooks are helpful for scientific 
code by allowing you to keep detailed notes, code, and output in one place. 

They also have the added benefit of having HTML file output that is styled and 
easy to read. 
Saving your R Markdown notebook will create an HTML file containing the code and 
output to be saved alongside it (click the *Preview* button or press 
*Cmd+Shift+K* to preview the HTML file). 
The preview shows you a rendered HTML copy of the contents of the editor. 
Consequently, unlike *Knit*, *Preview* does not run any R code chunks. 
Instead, the output of the chunk when it was last run in the editor is displayed.

#### 4) `sessionInfo` tells people what packages you used

The `sessionInfo` function prints out what packages and versions you used in your 
analysis.
This way, when others try to reproduce your research, they know what packages 
you have loaded and how things were set for the code you ran. 

```{r}
# Print out the versions and packages we are using in this session
sessionInfo()
```
