---
title: 'Exercise: Clean Metadata and Prepare Gene Expression Matrix for Downstream
  Analysis'
output:
  html_notebook:
    toc: yes
    toc_float: yes
---
The goal of these exercises is to help you get familiar with using tidyverse 
functions to manipulate data. 

In this example, we will use a [medulloblastoma dataset](https://www.refine.bio/experiments/GSE49243/gene-expression-data-from-medulloblastoma-tumor-samples) 
that has been processed on [refine.bio](http://docs.refine.bio/en/latest/main_text.html#processing-information). 
In short, this study collected medulloblastoma microarray profiles from patients
of a variety of ages and includes information about their demographics and 
histology of the medulloblastoma samples [(Kool et al. _Cancer Cell_ . 2014.)](https://www.ncbi.nlm.nih.gov/pubmed/24651015).

### Setup 

Before we start the exercises, we need to create the necessary directories and 
load the necessary libraries.

#### Creating necessary directories

We need to create a directory to store the results of the exercises. 
We have done this for you. 
Run the code chunk below and take a look at your files pane. 
You should see the directory `exercise-results` created. 

```{r}
results_dir <- "exercise-results"
# creating results directory if it doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```

#### Loading libraries

For these exercises, we will need the `tidyverse` libraries. 
Go ahead and load the library in the code chunk below.
Don't forget to run the code chunk once you write the command! 

```{r}
# load tidyverse library

```

#### Importing files

Let's import all the TSV files we will need for this exercise. 
The metadata is present in `exercise-data/metadata_GSE49243.tsv` and gene 
expression matrix is in `exercise-data/GSE49243.tsv`.
Go ahead and read the metadata and gene expression matrix into variables called
`metadata` and `gene_expression`.
Remember to use `file.path`. 

```{r}
# reading metadata
metadata <- 

# reading gene expression matrix
gene_expression <- 
```

### Cleaning metadata

First, we are going to clean the metadata using filtering, renaming, reordering 
and we will write our newly cleaned metadata into a new file. 

#### Inspect the metadata

Now that you have read the metadata into the variable `metadata`, explore what 
the metadata looks like. 

```{r}
# explore the metadata

```

#### Filtering and trimming

Next, we're going to do some manipulations on the metadata to retain only the 
information we are interested in. 
Below we have done some manipulations on the metadata. 
Take a look the steps we performed and then perform the same manipulations but 
by using pipes `%>%` and storing the result in `metadata_clean`. 

```{r}
# manipulations w/o pipes
metadata_nopipe <- metadata
metadata_nopipe <- filter(metadata_nopipe, refinebio_age < 18)
metadata_nopipe <- select(metadata_nopipe, refinebio_accession_code, 
                          refinebio_age, refinebio_sex, histology)
metadata_nopipe <- arrange(metadata_nopipe, refinebio_age)
```

It's your turn! 

```{r}
# now you try it, with pipes %>%
metadata_clean <- 
  
```

Run the code chunk below to check if you got the same results. 

```{r}
all.equal(metadata_nopipe, metadata_clean)
```

#### Rename columns

The column names `refinebio_accession_code`, `refinebio_age`, `refinebio_sex` 
are too unwieldy. 
Let's rename them to something more manageable like `sample_id`, `age`, and 
`sex` respectively. 

To do this, tidyverse has a function `dplyr::rename`. 
Use the help window to understand how you can use this function. 
Rename the variables in `metadata_clean` and store those changes as the same object name. 
(Feel free to google for solutions as well!)

```{r}
# use the help window to learn how you can use the rename function
metadata_clean <- 

```

Use the `head` function check if the column names changed.

```{r}

```

### Summarize metadata

Now, let's calculate the mean age by sex, mean age by histology, and counts of 
each tissue by sex. 

If you get stuck at any point, feel free to refer to the help window or google 
for solutions.

```{r}
# calculate mean age by sex

```

```{r}
# calculate mean age by histology

```

```{r}
# calculate how many of each histology type exists for each sex

```

### Select the samples in the gene expression matrix

Now, select the samples in the gene expression matrix so that it has the same 
samples as `metadata_clean`. 
(Remember we read the gene expression matrix into `gene_expression`)

```{r}
# select the gene expression matrix by samples in the cleaned metadata
cleaned_gene_expression <- 
```

### Write to files

#### Write to cleaned metadata and gene expression matrix to files

Last step, write the cleaned metadata into `metadata_GSE49243_cleaned.tsv` and 
cleaned gene expression 
matrix into `GSE49243_cleaned.tsv` the results directory (which we saved as 
`results_dir`).

```{r}
# write cleaned metadata to file


# write cleaned gene expression matrix to file

```

To check if you've completed these steps successfully, you should check for two 
files called `metadata_GSE49243_cleaned.tsv` and `GSE49243_cleaned.tsv` in your 
`exercise-results` directory. 
Open the files and check to see if they contain the 
information you would expect. 

### Session Info

Print out the session info.
```{r}

```
