---
title: 'Exercise: Getting comfortable with R'
output:
  html_notebook:
    toc: yes
    toc_float: yes
---

The goal of these exercises is to help you get comfortable with using R and R 
notebooks. 

We will create a play data set for a imaginary mouse experiment where you have a 
control and a treatment group of mice. 

### Create a imaginary mouse experiment

**Characters**

Use the empty chunk below to create a character vector with made up sample IDs 
for your mice in your imaginary experiment.
Call this new vector `mouse_id` and make sure each id is distinct.  

```{r}
mouse_id <-
```

Check your environment panel. Is your `mouse_id` there? 

Use the empty code chunk below to evaluate the structure and type of your object.

```{r}

```

Now evaluate the `length` of `mouse_id`. 
Remember you can look up information about a function using `?`

```{r}

```

Now use the `unique` function to check if your `mouse_id` vector has all unique
ids. 
Print out the original `mouse_id` vector and the unique `mouse_ids`s and compare 
them. 

```{r}

```

**Factors**

Now make a factor that is the same length as `mouse_id` but contains the 
experiment group information for each mouse. 
- Call this vector `exp_group`. 
- Stick to 2 or 3 experimental groups. 

To make this, start with a character vector that has values repeated and then use the
`as.factor` function. 

```{r}

```

Use the function `summary` to look at your factor vector. 
How many experimental groups do you have?  

```{r}

```

Now use the `length` function and the `==` operator to check if your `mouse_id` 
and `exp_group` vectors are the same length. 

```{r}

```

Use `levels` to explore the order of the levels of `exp_group`. 

```{r}

```

Do the levels make sense? Is your control the first level? 
If not, create a a new code chunk here and re-order your levels. 
Look up the `levels` function using `?` to do this. 

**Numeric** 

Let's create some fake gene expression data. 
We will be using a random number generator to do this, so use this code chunk to 
set the seed in preparation for that. 

```{r}

```

We've set up a template for the first fake gene's data for you. 
We will generate normally distributed data for a gene we will call `gene_a`.
Look up the `rnorm` function to get more info on what this does and it's 
associated functions. 

```{r}
gene_a <- rnorm(<SAMPLE_SIZE>, mean = <SET_A_MEAN>, sd = <SET_A_SD>)
```

Use the `summary` function to explore `gene_a`.

```{r}

```

Use the template from above to make `gene_b`. 
Feel free to alter the mean and sd arguments

```{r}

```

Explore your `gene_b`

```{r}

```

## Combine our information into a data.frame

Now that we have created `mouse_id`s, `exp_group`s and gene data, let's combine 
this all in one data.frame.

```{r}
mouse_data <- 
```

Use this chuink to explore this data.frame

```{r}

```


```{r}

```

### Session Info

Print out the session info.
```{r}

```
