---
title: 'Exercise: Getting comfortable with R'
output:
  html_notebook:
    toc: yes
    toc_float: yes
---

The goal of these exercises is to help you get comfortable with using R and R 
notebooks. 

We will create a play data set for an imaginary mouse experiment where you have 
control and treatment groups of mice. 

### Create a imaginary mouse experiment

Use the empty chunk below to create a *character* vector with made up sample IDs 
for your mice in your imaginary experiment.
Call this new vector `mouse_id` and make sure each ID is distinct.  

```{r}
mouse_id <-
```

Check your environment panel. Is `mouse_id` there? 

Use the empty code chunk below to evaluate the structure and type of your object.

```{r}

```

Now evaluate the `length` of `mouse_id`. 
Remember you can look up information about a function using `?`

```{r}

```

Now use the `unique` function to check if your `mouse_id` vector has all unique
ids. 
Print out the original `mouse_id` vector and the unique `mouse_ids`s and compare 
them. 

```{r}

```

Now make a *factor* vector that is the same length as `mouse_id` but contains 
the experiment group information for each mouse.  
- Call this vector `exp_group`.  
- Stick to 2 or 3 experimental groups.  

To make this, start with a character vector that has values repeated and then 
use the `as.factor` function. 

```{r}

```

Use the function `summary` to look at your factor vector. 
How many experimental groups do you have?  

```{r}

```

Now use the `length` function and the `==` operator to check if your `mouse_id` 
and `exp_group` vectors are the same length. 

```{r}

```

Use `levels` to explore the order of the factor levels of `exp_group`. 

```{r}

```

Do the levels make sense? Is your control the first level? 
If not, create a a new code chunk here and re-order your levels. 
Look up the `levels` function using `?` to do this. 

Let's create some fake gene expression data. 
We will be using a random number generator to do this, so use this code chunk to 
set the seed in preparation for that. 

```{r}

```

We've set up a template for the first fake gene's data for you. 
We will generate normally distributed data for a gene we will call `gene_a`.
Look up the `rnorm` function to get more info on what this does and it's 
associated functions. 

```{r}
gene_a <- rnorm(<SAMPLE_SIZE>, mean = <SET_A_MEAN>, sd = <SET_A_SD>)
```

Use the `summary` function to explore `gene_a`.

```{r}

```

Use the template from above to make a `gene_b` and a `gene_c`. 
Feel free to alter the `mean` and `sd` arguments as you please. 

```{r}

```

Explore your `gene_b` and `gene_c`.

```{r}

```

## Combine our information into a data.frame

Now that we have created `mouse_id`s, `exp_group`s and gene data, let's combine 
this all in one data.frame.

```{r}
mouse_data <- 
```

Use this chunk to explore this data.frame.

```{r}

```

Write these data to their own TSV file using the `readr::write_tsv` function
Use `file.path` to save this file to the `results` folder we created in 
`02-intro-to-tidyverse.Rmd`. 

```{r}

```

Use the `file.exists` function to check if you successfully created this TSV file.

```{r}

```

## Get summary statistics

Use indexing to isolate only the gene data. 

```{r}
gene_data <- 
```

Use `apply` to obtain total of gene a, b, and c expression for each mouse sample. 
Call this object `total_expression`.
Remember 1 = rows and 2 = columns. 

```{r}

```

Explore the `total_expression` object.

```{r}

```

Use `apply` to obtain the means of gene a, b, and c. 
Call this `gene_means`.

```{r}

```

Explore the `gene_means` object.

```{r}

```

Use the `==` operator to check the `gene_means` individually using `[ ]` and
checking them against means calculated by subsetting each gene using `mouse_data$`.  

```{r}

```

### Session Info

Print out the session info.
```{r}

```
