---
title: "Introduction to R - Exercises"
author: "Candace Savonen"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives
+ Navigate the RStudio environment
+ Use R for simple calculations, both mathematical and logical
+ Understand and apply base R functions
+ Understand, define, and use R datatypes, including vector manipulation and indexing
+ Understand the anatomy of a data frame

### Set Up 

Use this chunk to load the `tidyverse` library. 

```{r}
library(tidyverse)
```

## Read in the gene results file

Use `readr::read_tsv` to read in the file "gene_results_GSE44971.tsv" and assign 
it the name `stats_df`. 

```{r}
stats_df <- 
```

Use this chunk to explore what your data frame, `stats_df` looks like. 

```{r}

```

```{r}

```


### Defining and using variables 

```{r}

```

+ `mean()`: The average value in the vector
+ `min()`: The minumum value in the vector
+ `max()`: The maximum value in the vector
+ `sum()`: The sum of all values in the vector

Use subsetting to take a look at the `gene_symbol` column. 

```{r}
stats_df$gene_symbol

## OR ##

stats_df[, 2]
```

Determine what kind of data type the `gene_symbol` column is. 

```{r}
str(stats_df$gene_symbol)
```

Using the `dplyr::mutate` function, make a new variable that turns `gene_symbol` to a `factor` using `as.factor`. 

```{r}
summary(as.factor(stats_df$gene_symbol))
```



### Session Info

```{r}
sessionInfo()
```

