---
title: "Introduction to R - Exercises"
author: "Candace Savonen"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

The goal of these exercises is to help you get comfortable with using R and R 
notebooks by continuing to play with the gene results dataset we used in the [01-intro_to_base_R](./01-intro_to_base_R.Rmd) and [02-intro_to_ggplot2](./01-intro_to_ggplot2.Rmd) notebooks. 
It is a pre-processed [astrocytoma microarray dataset](https://www.refine.bio/experiments/GSE44971/gene-expression-data-from-pilocytic-astrocytoma-tumour-samples-and-normal-cerebellum-controls) 
that we performed a set of [differential expression analyses on](./scripts/00-setup-intro-to-R.R).

### Set Up 

Use this chunk to load the `tidyverse` library. 

```{r}

```

## Read in the gene results file

Use `readr::read_tsv` to read in the file "gene_results_GSE44971.tsv" and assign 
it the name `stats_df`. 

```{r}
stats_df <-
```

Use this chunk to explore what your data frame, `stats_df` looks like. 

```{r}

```

## Read in the metadata

Use `readr::read_tsv` to read in the file "cleaned_metadata_GSE44971.tsv" and assign 
it the name `metadata`. 

```{r}
metadata <- 
```

Use this chunk to explore what your data frame, `metadata` looks like. 

```{r}

```

### Subsetting variables

Use subsetting to take a look at the `avg_expression` column. 
Hint: `$`

```{r}

```

Use the `min()` argument to find what the minimum average expression in this dataset is. 
Remember you can use `?min` or the help panel to find out more about a function. 

```{r}
# Find the minimum average expression value

```

Find the `log`, using base 2, of the average expression values. 

```{r}
# Find the log of base 2 of the average expression

```

## Using logical arguments

Use subsetting to take a look at the `adj_p_value` column. 

```{r}

```

Find out which p-values in total are below a `0.05` cutoff using a logical 
statement. 

```{r}

```

Name the logical vector you created above as `significant_vector`.

```{r}

```

Use `sum` to count how many p values in the total set are below this cutoff. 

```{r}

```

## Filter the dataset

Subset the column `contrasts` from `stats_df`. 

```{r}

```

Construct a logical vector around the contrasts variable you subset above that 
indicates which rows of data in `stats_df` are from the `astrocytoma_normal` 
contrast test. 

```{r}

```

Use `dplyr::filter` to keep only the data for the `astrocytoma_normal` contrast
in `stats_df`. 

```{r}
astrocytoma_normal_df <- dplyr::filter(<DATA_FRAME>, <LOGICAL_VECTOR>)
```

Use the `nrow` argument on `astrocytoma_normal_df` to see if your filter worked. 
You should have `2268` rows. 

```{r}
nrow(astrocytoma_normal_df)
```

Do the same filtering as you did above, but this time, use a `%>%` by removing 
`stats_df` from the first argument position and piping it into the filter. 

```{r}
astrocytoma_normal_df <- <DATA_FRAME> %>% 
  dplyr::filter(<LOGICAL_VECTOR>)
```

Save your filtered data to a TSV file using `readr::write_tsv`. 
Call it `astrocytoma_normal_contrast_results.tsv` and save it to the `results`
directory. 

```{r}

```

### Create a density plot 

Set up a ggplot object for `astrocytoma_normal_df` and set `x` as the average 
expression variable. 
Use the `+` to add on a layer called `geom_density()`

```{r}
ggplot(<DATA_FRAME>, aes(x = <VARIABLE_NAME>)) + 
  geom_density()
```

Use the plot you started above and add a `ggtheme` layer to play with its aesthetics (e.g. `theme_classic()`)
See the [ggthemes vignette](https://mran.microsoft.com/snapshot/2017-02-04/web/packages/ggthemes/vignettes/ggthemes.html) 
to see a list of ggtheme options. 

```{r}

```

Feel free to make other customizations to this plot by adding more layers with `+`. 
You could add a `ylab()` and `xlab()` and/or by getting inspiration from this [handy cheatsheet for ggplot2](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf). 

```{r}
# Customize your plot!
```

Save your plot as a `PNG`. 

```{r}
ggsave(
  plot = last_plot(), # This will retrieve the last plot you printed. 
  filename = file.path("plots", "average_expression_density_plot.png")
)
```

### Session Info

```{r}
sessionInfo()
```

