---
title: "Introduction to Tidyverse - Exercises Part 1"
author: "Candace Savonen"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

The goal of these exercises is to practice the visualization techniques and 
data frame manipulations we used in [02-intro_to_ggplot2](./01-intro_to_ggplot2.Rmd) 
and [03-intro_to_tidyverse](./01-intro_to_tidyverse.Rmd) notebooks. 

Here we will use another [microrarray dataset](https://www.refine.bio/experiments/GSE19578/integrated-molecular-genetic-profiling-of-pediatric-high-grade-gliomas-reveals-key-differences-with-adult-disease) that is [pre-processed for us by refine.bio](http://docs.refine.bio/en/latest/main_text.html#processing-information).
This study from [Paugh et al, 2010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2903336/) has glioblastoma samples from a range of patient ages. 

### Set Up 

In this notebook, we will be using tidyverse packages. 

```{r}

```

It's convenient to declare our directory names at the beginning so we can change
it once and have it change throughout the notebook. 

```{r}
data_dir <- "data"
results_dir <- "results"
plots_dir <- "plots"
```

Let's set up an `if` statement that will create our plots directory with 
`dir.create()` if it does not currently exist. 

```{r}

```

## Read in the gene results file

Use `readr::read_tsv()` to read in the gene expression matrix TSV, "GSE19578.tsv" 
and assign it the name `genes_df`. 

```{r}
genes_df <-
```

Use this chunk to explore what your data frame, `genes_df` looks like. 

```{r}

```

## Read in the metadata

Use the same function to read in the file "metadata_GSE19578.tsv" and assign 
it to a variable named `metadata`. 

```{r}
metadata <-
```

Use this chunk to explore what your data frame, `metadata` looks like. 

```{r}

```

## Cleaning the metadata

As is common with metadata, we will need to do some cleaning. 
Although refine.bio and GEO give us a lot of information, a lot of these fields 
are not informative for us at this stage. 

**Step 1 of cleaning**: keep only informative columns. 

Use `dplyr::select()` to keep the 3 columns with the following information: 
+ The sample accession codes
+ age as a numeric variable
+ disease information

Note that all the columns we want start with `refinebio_`. 
Assign the `dplyr::select()`ed metadata as `metadata_clean`. 

```{r}

```

Look up details on `dplyr::select()` and figure out how we can (in the same step!) 
rename these columns to be shorter and more to the point. 
Rename the three columns `sample_id`, `age`, and `disease`, respectively.

```{r}

```

**Step 2 of cleaning**: format variables to the type they should be.

Currently, our `disease` variable from `metadata_clean` has repeated values, 
but using `summary()` wouldn't show us that: 

```{r}

```

To avoid handling issues of this variable, let's coerce it to a `factor` using
`dplyr::mutate()`. 
We'll need to use `metadata_clean` and reassign it the same name. 

```{r}

```

Use `summary()` on the re-factored `disease` column to see how many samples of each
disease there are. 

```{r}

```

**Step 3 of cleaning**: Create new grouping variables. 

Using the age variable, let's create a new grouping variable that indicates a 
`TRUE/FALSE` of whether the patient is above `18`. 

```{r}

```

**Step 4 of cleaning**: Arrange metadata in a sensible order

Use `dplyr::arrange()` to arrange the rows from youngest to oldest `age`. 

```{r}

```

**Step 5 of cleaning**: Write the cleaned metadata to TSV

Now that we've cleaned our metadata, let's save it to a file using 
`readr::write_tsv()`.

```{r}

```

Now that we've determined all the steps that need to be taken to clean our data, 
we can assemble these steps into one set of piped (`%>%`) transformations. 
Assemble steps 1-5 into one set where each step pipes (`%>%`) into the next. 
You should also make sure to have comments (`#`) on each step so we know why we
are doing it when we look back at this later!

```{r}

```

## Set up gene data

#### Order sample data

Commonly, the sample order for our metadata and gene expression matrix will not 
automatically match. 
This should be the first thing we check before we try to use them in tandem. 

Use `all.equal()` to see if the sample ids of columns of `genes_df` are in the 
same order as our `sample_id` column in `metadata_clean`. 
Hint: you will want to ignore the `Gene` column when checking this, try a negative 
index

```{r}

```

Before we reorder our samples, we should also check if all the samples list in 
`metadata_clean` also exist in `genes_df`. 
Use `all()` and `%in%` to write a logical expression to check this.

```{r}

```

Now let's use `dplyr::select()` to reorder the columns in `genes_df` to the order 
of `sample_id`s in `metadata_clean`. 
Hint: Don't forget about keepting the `Gene` column! 

```{r}

```

Re-test the order of the samples the same way you did before, using `all.equal()`

```{r}

```

#### Filter out low value genes

For microarray data, lower expression value genes may be an indication of 
unreliable measurement.

Let's calculate the median expression per gene using `apply()`. 
Don't forget to skip the `Gene` column (you can use `genes_df[, -1]`). 
Store this vector as `gene_medians`. 

```{r}

```

Use `summary()` to get a overall view of the `gene_medians` distribution. 

```{r}

```

In order to plot these data, we need to make `gene_medians` into a data frame. 
Use the function `as.data.frame()` to do so.

```{r}
gene_medians_df <- 
```

Now that our data is in a data.frame format, we can use `ggplot2::ggplot()` to 
plot it and add layers with `+`.
Add a layer with `geom_density()` to make a density plot of the medians. 

```{r}

```

If you haven't, don't forget to add `xlab()` and `ylab()` to your plot so your 
plot is appropriately labeled and easier to read. 

When you have your plot finalized, save it to a PNG to our `results_dir` using 
`ggplot2::ggsave()`.

```{r}

```

Now that we've seen what our distribution looks like, let's apply a filter 
cutoff to `genes_df`. 
Choose a minimum gene expression cutoff and use `dplyr::filter()` to filter out 
the low expression genes based on that cutoff. 
Recall that `dplyr::filter()` will *keep* rows where your logical expression is 
`TRUE`. 
We'll call this newly filtered data frame `filtered_genes_df`. 

```{r}

```

Let's write our newly filtered and arranged dataset to TSV file. 
Save this TSV file back to our `data_dir` and call it "filtered_GSE19578.tsv".

```{r}

```

### Session Info

```{r}
sessionInfo()
```

