---
title: "Introduction to Tidyverse - Exercises Part 1"
author: "Candace Savonen"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

tidyverse exercises - part 1

The goal of these exercises is to practice the visualization techniques and 
data frame manipulations we used in [02-intro_to_ggplot2](./01-intro_to_ggplot2.Rmd) 
and [03-intro_to_tidyverse](./01-intro_to_tidyverse.Rmd) notebooks. 

Here will use another [microrarray dataset](https://www.refine.bio/experiments/GSE19578/integrated-molecular-genetic-profiling-of-pediatric-high-grade-gliomas-reveals-key-differences-with-adult-disease) that is [pre-processed for us by refine.bio](http://docs.refine.bio/en/latest/main_text.html#processing-information).
This study from [Paugh et al, 2010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2903336/) has glioblastoma samples from a range of patient ages. 

### Set Up 

In this notebook, we will be using tidyverse packages. 
Load the library here, or you can use the `::` strategy.

```{r}

```

It's convenient to declare our directory names at the beginning so we can change
it once and have it change throughout the notebook. 

```{r}
data_dir <- "data"
results_dir <- "results"
plots_dir <- "plots"
```

## Read in the gene results file

Use `readr::read_tsv()` to read in the gene expression matrix TSV, "GSE19578.tsv" 
and assign it the name `genes_df`. 

```{r}
genes_df <- 
```

Use this chunk to explore what your data frame, `genes_df` looks like. 

```{r}

```

## Read in the metadata

Use the same function to read in the file "metadata_GSE19578.tsv" and assign 
it the name `metadata`. 

```{r}
metadata <- 
```

Use this chunk to explore what your data frame, `metadata` looks like. 

```{r}

```

## Cleaning the metadata

As is common with metadata, we will need to do some cleaning. 
Although GEO gives us a lot of information, a lot of these fields are not informative for us at this stage. 

**Step 1 of cleaning**: keep only informative columns. 

Use `dplyr::select` to keep the 3 columns with the following information: 
+ The sample accession codes
+ age as a numeric variable
+ disease information

Assign the `select()`ed metadata as `metadata_clean`. 

```{r}

```

Look up details on `select()` and figure out how we can (in the same step!) 
rename these columns to be shorter and more to the point like `sample_id`, 
`age` and `disease` for example. 

```{r}

```

**Step 2 of cleaning**: format variables to the type they should be.

Currently, our `disease` variable has repeated values, but using `summary()` 
wouldn't show us that: 

```{r}

```

To avoid handling issues of this variable, let's try to coerce it to a `factor`
using `dplyr::mutate()`. 
We'll need to use `metadata_clean` and reassign it the same name. 

```{r}

```

Use `summary()` on the re-factored `disease` column to see how many samples of each
there are. 

```{r}

```

**Step 3 of cleaning**: Create new grouping variables. 

Using the age variable, let's create a new grouping variable that indicates a 
`TRUE/FALSE` of whether the patient is above `18`. 

```{r}

```

**Step 4 of cleaning**: Arrange metdata in a sensible order

Use `dplyr::arrange()` to arrange the rows from youngest to oldest `age`. 

```{r}

```

**Step 5 of cleaning**: Write the cleaned metadata to TSV

Now that we've cleaned our metadata, let's save it to a file called "cleaned_metadata_GSE19578.tsv" using `readr::write_tsv()`.

```{r}

```

Now that we've determined all the steps that need to be taken to clean our data, 
we can assemble these steps into one set of piped (`%>%`) transformations. 
Assemble steps 1-5 into one set where each step pipes (`%>%`) into the next. 
You should also make sure to have comments (`#`) on each step so we know why we
are doing it when we look back at this later!

```{r}
# Reconstruct `metadata_clean` in one set of piped steps!
```

## Set up gene data

#### Order sample data

Commonly, the sample order for our metdata and gene expression matrix will not 
automatically match. 
This should be the first thing we check before we try to use them in tandem. 

Use `all.equal()` to see if the sample ids of columns of `gene_df` are in the 
same order as our `sample_id` column in `metadata_clean`. 
Hint: you will want to ignore the `Gene` column when checking this, try a negative 
index

```{r}

```

Before we reorder our samples, we should also check if all the samples list in 
`metadata_clean` also exist in `gene_df`. 
Use `all()` and `%>%` to write a logical expression to check this.

```{r}

```

Now let's use `dplyr::select()` to reorder the columns in `gene_df` to the order 
of `sample_id`s in `metadata_clean`. 
We've gotten you part of the way there by setting up a vector (`cols_to_keep`) you will use in 
`dplyr::select()`.

```{r}
# Use this vector inside of a `dplyr::select()` to select the columns we want from
# gene_df 
cols_to_keep <- c("Gene", metadata_clean$sample_id)


```

Re-test the order of the samples the same way you did before, using `all.equal()`

```{r}

```

#### Filter out low value genes

For microarray data, lower expression value genes may be an indication of 
unreliable measurement.

Let's calculate the median expression per gene using `apply()`. 
Don't forget to move `Gene` column out of the way. 
Store this vector as `gene_medians`. 

```{r}

```

Use `summary()` to get a overall view of the `gene_medians` distribution. 

```{r}

```

In order to plot these data, we need to make `gene_medians` into a data frame. 
Use the function `as.data.frame()` to do so.

```{r}
gene_medians_df <- 
```

Now that our data is in a data.frame format, we can use `ggplot2::ggplot` to 
plot it and add layers with `+`.
Add a layer with `geom_density()` to make a density plot of the medians. 

```{r}
density_plot <- ggplot2::ggplot(gene_medians_df, 
                               ggplot2::aes(x = <VARIABLE_NAME>)) + 

# Print a preview of the plot
density_plot
```

If you haven't, don't forget to add `xlab()` and `ylab()` to your plot so your 
plot is appropriately labeled and easier to read. 

When you have your plot finalized, save it to a PNG to our `plots_dir` using 
`ggplot2::ggsave()`.

```{r}
ggplot2::ggsave(
  plot = density_plot,
  filename = file.path(plots_dir, "gene_median_density_plot.png")
)
```

Now that we've seen what our distribution looks like, let's apply it as filter
cutoff to `genes_df`. 
Use `dplyr::filter()` to filter out the low expression genes based on your chosen 
cutoff. 
We'll call this newly filtered data frame `filtered_genes_df`. 

```{r}
filtered_genes_df <- 
```

Let's write our newly filtered and arraged dataset to TSV file. 
Save this TSV file back to our `data_dir` and call it "filtered_GSE19578.tsv".

```{r}

```

### Session Info

```{r}
sessionInfo()
```

