---
title: "Introduction to base R - Exercises"
author: "CCDL for ALSF"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

This notebook contains problems designed for you to practice what you have learned in the first day of instruction, in particular in the [`01-intro_to_base_R`](01-intro_to_base_R-live.Rmd) notebook.
We will review basics of calculating with R, using variables, functions and types.

## Calculations and variables

We'll start with some basic manipulations of values and variables.

In the chunk below calculate the product of 1.563 and 4.678.

```{r multiply, solution = TRUE}

```

Now assign the value 5.89 to a variable `numerator` and 2.2x10¹² to `denominator` and calculate their quotient. 
Note that you will probably want to use engineering notation to enter the denominator value.

```{r divide, solution = TRUE}

```

Create a variable called `name` that contains your name as text.

```{r name, solution = TRUE}

```

Run the chunk below to use `paste0()` to say hello to yourself, and to check if you have correctly entered your name above!

```{r hello}
cat(paste0("Hello ", name, "!!"))
```


### Vectors

Create a vector named `tri` with the following sequence of numbers: 1, 3, 6, 10, 15, 21. 

```{r vector, solution = TRUE}

```

Verify that the "type" of this vector is `numeric`

```{r type-check, solution = TRUE}

```

Extend `tri` with the next two numbers in the series, 28 and 36, and store the result back in `tri`.

```{r extend-vector, solution = TRUE}

```


Calculate the square root of each number in `tri`. 
You should be able to do this with a single call to the function `sqrt()`

```{r sqroot, solution = TRUE}

```

Add the word "triangle" as the first element of `tri`.
What happens to the type of `tri`?

```{r auto-coerce, solution = TRUE}

```

## Data frames

### Reading a file

To start working with data frames, we will read in the refine.bio metadata from an [astrocytoma microarray dataset](https://www.refine.bio/experiments/GSE44971/gene-expression-data-from-pilocytic-astrocytoma-tumour-samples-and-normal-cerebellum-controls).
This is a tab separated file named `metadata_GSE44971.tsv`, which can be found in the `data` subfolder of the `intro-to-r-tidyverse` folder. 
You can use the file browser pane to navigate to the file location, but don't try to open it from there!
Recalling that notebooks consider file locations relative to the folder they are located in, use `file.path()` to create the path to the file and store it in the variable `metadata_file`.

```{r metadata-path, solution = TRUE}

```

If you have done this correctly, the following block should return `TRUE`.

```{r}
file.exists(metadata_file)
```

We can then use the function `read_tsv()` from the `readr` package to read in the file.
First, we will load the `readr` package using `library()`:

```{r}
library(readr)
```

Now read the file, assigning the data frame in a variable named `metadata`.

```{r read-meta, solution = TRUE}

```

### Working with a data frame

How many rows and columns are there in the `metadata` data frame?

```{r meta-dim, solution = TRUE}

```

What type of data is in the `data_row_count` column?
What about the `refinebio_accession_code` column?
Use `$` notation to select these columns and check their types.

```{r column-type, solution = TRUE}

```

Select the first four sample accession codes:

```{r select-samples, solution = TRUE}

```



Some of the potentially interesting columns in this table (and others you will encounter) have spaces and other characters in their column names. 
Unfortunately, because a column like `brain location` has spaces in its name, it is a bit annoying to access.
If we didn't do anything special and tried to type `metadata$brain location`, R would think the space was the end of the variable, and we would most likely get an error.
One way to get around this is to use the following notation, adding backticks around the column name so the spaces are treated as part of the variable name.

```{r}
metadata$`brain location`
```

Since this can be annoying to have to type over and over, pull out that column into its own variable, named `brain_loc`. 
What type of variable is this?
```{r brain-loc, solution = TRUE}

```

Convert `brain_loc` to a `factor` (saving back to the `brain_loc` variable) and summarize the brain locations surveyed in the dataset.
What is the second most common brain location?

```{r brain-summary, solution = TRUE}

```

Which samples are from the cerebral hemisphere? 
Create a boolean vector named `cerebral` that is `TRUE` for cerebral hemisphere samples and `FALSE` otherwise.

```{r cerebral, solution = TRUE}

```

If you have done this correctly, the following test should return `TRUE`
```{r}
all.equal(metadata$refinebio_accession_code[cerebral], 
          c("GSM1094824", "GSM1094819", "GSM1094816", "GSM1094843")
)
```



### Session Info

```{r}
sessionInfo()
```

