---
title: "Introduction to Tidyverse - Exercises Part 2"
author: "CCDL for ALSF"
date: "2020"
output:
  html_notebook:
    toc: true
    toc_float: true
---

The goal of these exercises is to practice the visualization techniques and
data frame manipulations we used in [`02-intro_to_ggplot2`](02-intro_to_ggplot2-live.Rmd) and [`03-intro_to_tidyverse`](03-intro_to_tidyverse-live.Rmd) notebooks.

Here we will use another [microarray dataset](https://www.refine.bio/experiments/GSE19578/integrated-molecular-genetic-profiling-of-pediatric-high-grade-gliomas-reveals-key-differences-with-adult-disease) that is [pre-processed for us by refine.bio](http://docs.refine.bio/en/latest/main_text.html#processing-information).
This study from [Paugh et al, 2010](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2903336/) has glioblastoma samples from a range of patient ages.

**This notebook takes advantage of some of the work that you have done in the `exercise_03a-intro-to-tidyverse` notebook, so be sure to complete that notebook first!**

## Set Up

In this notebook, we will be using tidyverse packages.
You can load them with `library()` or use the `::` strategy.

```{r libraries, solution = TRUE}

```

It's convenient to declare our directory names at the beginning so we can change it once and have it change throughout the notebook.
Remember we need to be consistent with what we used in the previous notebook.

```{r}
data_dir <- "data"
plots_dir <- "plots"
```

## Read in the cleaned metadata

Use `readr::read_tsv` to read in the filtered gene expression matrix TSV, `filtered_GSE19578.tsv`, that you created in Part 1 of these exercises.
Save it in a variable called `genes_df`.

```{r read-filtered-genes, solution = TRUE}

```

Use `readr::read_tsv()` to read in the cleaned metadata TSV, "cleaned_metadata_GSE19578.tsv" and assign it the name `metadata`.

```{r read-cleaned-meta, solution = TRUE}

```

## Grouped summary statistics

Let's calculate some fold changes for our groups.
In the chunk below, we've set up a data frame that is transposed so you can continue using `tidyverse` functions.

```{r transform-genes}
transposed_genes_df <- genes_df |>
  # We want the gene information to be stored as rownames for the next set of
  # steps we are doing
  tibble::column_to_rownames("Gene") |>
  # Transposing the data frame so it is sample x genes
  t() |>
  # t() function turns our data frame into a matrix so let's turn it back to
  # a data.frame so we can use it with tidyverse functions
  as.data.frame() |>
  # By default our sample ids will be stored as rownames, let's bring it out as
  # its own columns
  tibble::rownames_to_column("sample_id")
```

Use `dplyr::select()` and an `%in%` statement to choose ~2-5 genes of interest from `transposed_genes_df` and save their data as a separate data frame object.
You can use [GeneCard](https://www.genecards.org/) to find the Ensembl IDs that correspond to your favorite genes.
If you don't have genes in particular you are interested in, feel free to use these:
`ENSG00000134323` (MYCN), `ENSG00000124762`(CDKN1A), and
`ENSG00000115414` (FN1).

```{r select-genes, solution = TRUE}

```

Use `dplyr::inner_join()` to join the `metadata` columns to `transposed_genes_df` by the `sample_id`.
Call this new data frame `genes_of_interest_df`.

```{r join-meta, solution = TRUE}

```

Calculate the age group means of your `dplyr::select()`ed genes.
(Recall that you made a column that indicates whether each sample is from an adult or not).

```{r group-means, solution = TRUE}

  # This solution we do a summary of each gene

# A more advanced, but succinct solution:

  # Here we are automatically calculating means for each variable that
  # starts with `ENSG`

```

### Plot the data

Now that we have our data set up, we will make some plots!

#### Plot 1) Scatterplot

Set up a ggplot scatterplot with either one or two of the selected genes in `genes_of_interest_df` and/or with `age`.

```{r scatterplot, solution = TRUE}

  # Make this a scatterplot

  # Add a ggplot2::ggtheme

  # Add a sensible y-axis label

```

#### Plot 2) Facet wrapped jitter

Create a jitter plot of each gene where each `disease` group has its own plot using `facet_wrap()`.

We've done the first step for you.
Your data will need to be in a long format.
Explore the `long_form_genes_oi_df` data frame to see how we've transformed it with `tidyr::pivot_longer()`.

```{r tidyr-pivot}
# Reformat to long format data frame
long_form_genes_oi_df <- genes_of_interest_df |>
  tidyr::pivot_longer(cols = tidyr::starts_with("ENSG"),
                      names_to = "gene",
                      values_to = "expression_value")
# Print preview
long_form_genes_oi_df
```

Make a `facet_wrap()`ped plot of the genes' expression levels, such as a boxplot or violin plot.

```{r gene-plot, solution = TRUE}

  # Make this a boxplot for now

  # Turn the x axis labels 90 degrees so they are easier to read

  # Facet wrap by gene
  # let the facets have different y axis scales

```

Make other customizations to this plot by adding more layers with `+` or change
the plot completely to something else!

**ggplot2 resources to get you going:**

- [Top 50 ggplots](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
- [Handy cheatsheet for ggplot2 (PDF)](https://github.com/rstudio/cheatsheets/raw/main/data-visualization.pdf)
- [ggplot2 courses and books from Tidyverse](https://ggplot2.tidyverse.org/)
- [ggplot2 data viz chapter of R for Data Science](https://r4ds.hadley.nz/data-visualize.html)
- [ggplot2 online tutorial](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html)

```{r plot-playground}
# Play with your plot and data!
# Ideas to get your juices flowing:
# - Add a jitter layer over the boxplot
# - Play with the color of the points!
# - Use ggplot2::ggthemes (e.g. theme_classic()) to play with the aesthetics
# - facet_wrap() by another variable
```

When you are happy with your plot, save your plot to the `plots_dir` as a `PNG` using `ggplot2::ggsave()`.

```{r save-plot, solution = TRUE}

```

### Session Info

```{r}
sessionInfo()
```

