---
title: "Exercise: Over-representation analysis of medulloblastoma PDX results"
output: 
  html_notebook:
    toc: true
    toc_float: true
author: CCDL for ALSF
date: 2021
---

In this exercise notebook, we will perform over-representation analysis (ORA) on differential gene expression analysis results from a PDX model of medulloblastoma.

## Introduction to the dataset

We will use an RNA-seq experiment comprised of medulloblastoma orthotopic patient-derived xenograft (PDX) models from [Huang *et al.* (2018)](https://doi.org/10.1126/scitranslmed.aat0150) to practice pathway analysis.

Here's an excerpt of [the experiment description](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA475298):

> RNA-seq data of Group 3 and 4 medulloblastoma with digoxin treatment.

Group 3 and Group 4 are subgroups of medulloblastoma. 
You can read more about these subgroups of medulloblastoma in [Menyhárt *et al.* (2019)](https://doi.org/10.1186/s13045-019-0712-y).

[Digoxin](https://en.wikipedia.org/wiki/Digoxin) is a treatment for various cardiovascular diseases.

*You may remember this dataset from the bulk RNA-seq exercise for exploratory data analysis!*

### DGE results

We've performed the differential gene expression analysis steps for you (you can see what we did [here](https://github.com/AlexsLemonade/training-modules/blob/master/pathway-analysis/setup/exercises/01-pdx_MB_DGE.Rmd)).

**Some things to note:**

* For the treatment comparison, `none` was the reference. 
  Genes with higher expression in `digoxin` will have positive fold change values.
* The design formula when we set up the `DESeqDataSet` was `~ Group + treatment`.
  The differential gene expression results for `treatment` in our example should be the differences between digoxin vs. none, controlling for differences due to molecular subtype (`Group`).
  
## Set up

### Libraries

Load in `magrittr`, `clusterProfiler`, `msigdbr`, and, `org.Hs.eg.db`.

```{r libraries, solution = TRUE}

```

### Directories and files

The DGE results are available at `data/medulloblastoma/pdx_medulloblastoma_treatment_dge.tsv`.
Use the next chunk to assign the path to the results file to a variable called `dge_results_file`.

```{r input_file, solution = TRUE}

```

Let's save the results of your over-representation analysis to `results/medulloblastoma`.
This directory might not exist yet, so you'll need to use the next chunk to assign the output directory to a variable called `output_dir` and create it if it doesn't exist yet.

```{r output_directory, solution = TRUE}

```

## Gene Sets

We can retrieve _all_ of the gene sets available for human in the `msigdbr` package with the command in the following chunk.

```{r msigdbr}
hs_msigdb_df <- msigdbr(species = "Homo sapiens")
head(hs_msigdb_df)
```

Take a look at the collections here: https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
The `gs_cat` and `gs_subcat` columns of `hs_msigdb_df` contain the collection information.
Once you've decided on what collection you want to use for your pathway analysis, use the next chunk to filter `hs_msigdb_df` and assign to a new data frame called `gene_set_df`. 
(Hint: We haven't loaded in the `dplyr` package.)

_Just a heads up, the solved version will use KEGG gene sets, but that doesn't mean you need to!_

```{r filter_gene_sets, solution = TRUE}
# Save the new filtered data frame as gene_set_df

```

## Read in data and convert gene identifiers

Use the next chunk to read in the DGE results and assign it to `dge_results_df`.

```{r read_in_dge, solution = TRUE}

```

> What gene identifiers are used in the DGE results?
  What gene identifiers are available in `gene_set_df`? 
  
To answer the first question, we can look at the `Gene` column that contains the gene identifiers.
The column name `Gene` doesn't tell us what kind of identifier we have!
When your instructors encounter an unfamiliar identifier, they often copy and paste one into Dr. Google.
All joking aside, for human datasets, you can search [GeneCards](https://www.genecards.org/).

For the first item in the list we searched the identifier in the `Gene` column: https://www.genecards.org/cgi-bin/carddisp.pl?gene=TSPAN6&keywords=ENSG00000000003
Notice under `External Ids for TSPAN6 Gene` that it indicates that this is an Ensembl identifier.
We can tell it's a _gene_ identifier because the prefix ends in `G`, rather than `T` which would indicate a transcript.

We'll need to convert the gene identifiers we have in the DGE results to something we can use for pathway analysis.

Let's remind ourselves what different gene identifiers are called in the `org.Hs.eg.db` package.

```{r keytypes}
keytypes(org.Hs.eg.db)
```

Use the next chunk to convert the identifiers in `Gene` column in the data frame of DGE results to Entrez IDs, which are identifiers you can use with `gene_set_df`.
Store the Entrez IDs in a new column (hint: use `dplyr::mutate()`) called `entrez_id`.

```{r identifier_conversion, solution = TRUE}
# Map from the identifiers in the DGE results data frame to one of the types of
# identifiers in gene_set_df and save the new identifier

```

## ORA

Now that we have our DGE results and have converted the gene identifiers, we're ready to perform ORA.
ORA requires a list of genes of interest, so we'll need to filter to some set of genes we'll use in our analysis.

Use the next chunk to get a list of genes you'll use for ORA.
Let's focus on genes that increase with treatment with digoxin.
Given what we know about how the results were obtained (see: [_DGE results_](#dge-results)), consider:

* Which columns of the DGE results do we need to use to filter?
* What thresholds should we use?

```{r genes_of_interest, solution = TRUE}

  # Filter to significant results & to genes with higher expression in digoxin
  # treated samples

  # Use dplyr::pull() to get a vector of gene identifiers

```

We want to remove any `NA` values (`!is.na()`) from the vector of genes and also make sure we don't include duplicates (`unique()`).

```{r unique_not_na}
significant_digoxin_genes <- unique(significant_digoxin_genes[!is.na(significant_digoxin_genes)])
```

In the next chunk, we generate the background set for our ORA. 
We are testing a list of _differentially expressed genes_, but not every gene in our list will be able to be **detected** by DGE.
For example, DESeq2 includes a step called independent filtering, where genes with low mean normalized counts will have adjusted p-values that are set to `NA` ([docs](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#independent-filtering-of-results)).
So to get our background set, we'll filter to remove genes (rows) that have no `NA` values associated with them with `tidyr::drop_na()`.
(This will also remove any genes without Entrez IDs.)

```{r background_set}
detected_genes <- unique(  # Only unique gene identifiers
  dge_results_df %>%
    # Remove any rows with NA
    tidyr::drop_na() %>%
    # Return a vector of Entrez IDs
    dplyr::pull(entrez_id)
)
```

Use the next chunk for the ORA step itself.

```{r run_ora, solution = TRUE}

```

`enricher()` returns a specialized object that has _slots_.
Slots can be accessed with `@` the same way you can get a column of a data frame with `$` and the column name.
`@result` will return the statistics you are likely most interested in in a tabular format.

```{r explore_ora, solution = TRUE, eval = FALSE}
# Use this chunk to explore the ORA results

```

Write the ORA results to file.

```{r write_ora, solution = TRUE}
# Save the file name to a variable - you'll probably want to customize the name
# of this file based on the analysis you choose to do (e.g., what gene sets)

# Save the results slot as a data frame

# Write to file with readr::write_tsv()

```

## Session Info

```{r}
sessionInfo()
```
