---
title: "CellMarker Genes for Over-Representation Analysis"
author: Jaclyn Taroni for the CCDL
date: 2021-06
output: 
  html_notebook: 
    toc: true
    toc_float: true
---

Adapted from [the `clusterProfiler` book section on Cell Markers](http://yulab-smu.top/clusterProfiler-book/chapter3.html#cell-marker), this notebook preps [CellMarker](http://bio-bigdata.hrbmu.edu.cn/CellMarker/) data for use in the ORA notebook.

```{r}
library(tidyverse)
```

```{r}
cell_markers <- read_tsv('http://bio-bigdata.hrbmu.edu.cn/CellMarker/download/Human_cell_markers.txt') %>% 
  # Remove undefined tissue types
  filter(tissueType != "Undefined") %>%
  # For the gene set "name" we're going to create a single column that has
  # all of this information
  unite("CellMarker", tissueType, cancerType, cellName, sep = ", ") %>% 
  # Composite gene set name + gene symbols, which are comma separated
  select(CellMarker, geneSymbol)
```

This dataset isn't quite ready for use with `clusterProfiler`:

```{r}
head(cell_markers, 10)
```

Some of the gene symbols have `[` or `]` to indicate when the mapping from a cell marker (e.g., `CDXX`) to a gene identifier is ambiguous.
We'll include any of the genes, so we'll simply remove the square brackets and then drop any instances where there are no gene symbols at all.

```{r}
cell_markers <- cell_markers %>%
  mutate(gene_symbol = str_remove_all(geneSymbol, "[\\[\\]]")) %>%
  drop_na(gene_symbol) %>%
  select(-geneSymbol)
```

The clusterProfiler functions generally want gene sets in a data frame where each row is a gene set, gene identifier pair. 
We'll need to split the comma-separated gene symbols and then create a new data frame.

```{r}
cell_markers_df <- cell_markers %>% 
  separate_rows(gene_symbol, sep = ", ") %>%
  rename(cell_marker = CellMarker)
```

And now write to file.

```{r}
output_dir <- file.path("..", "gene-sets")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
write_tsv(cell_markers_df,
          file.path(output_dir, "CellMarker_cleaned_human_markers.tsv"))
```

Session info.

```{r}
sessionInfo()
```
