---
title: "scRNA-seq Dimension Reduction"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**CCDL 2019**

In this notebook, we'll use dimension reduction techniques on single-cell 
RNA-seq data. 

As compared to bulk RNA-seq data, single cell RNA-seq data generally has more 
dimensions, (since we have data points for each individual cell level rather 
than the tissue as a whole). 
Because of this, its common to use dimension reduction techniques so all of the
data is in a more easily manageable form for plotting, clustering, or other 
downstream analyses. 

## Set Up 

For these analyses, we will need Seurat as well as the dplyr pipe. 

```{r}
# Load libraries
library(scran)
library(scater)

# Magrittr pipe
`%>%` <- dplyr::`%>%`
```

For this module, we are going to use the data that we processed part of in 
`02-tag-based_pre-processing_scRNA-seq.md`

```{r Load in Tabula Muris Data}
# Read in gene matrix
tab_mur_data <- readr::read_tsv(file.path("data", "tab_mur_scran_norm.tsv")) %>%
  dplyr::select(-'genes')

# Read in associated metadata
tab_mur_meta <- readr::read_tsv(file.path("data", "tab_mur_metadata.tsv"))
```

## Principal Components Analysis

Just like we did with bulk RNA-seq, we will also run PCA with our single-cell 
data. 
Previously, we identified the high variance genes in this dataset and now we 
will use these genes for PCA.

```{r PCA}
# Run PCA
pca_tab_mur <- prcomp(t(tab_mur_data))

# Make a dataframe with these PCA scores
pca_tab_mur <- data.frame(pca_tab_mur$x[, 1:2],
                          tab_mur_meta)
```

Let's plot our first two principal components and label our cells with what 
tissue they are from.

```{r Plot PCA}
# Plot this with ggplot2 and label the points with cell types
ggplot(pca_tab_mur, aes(x = PC1, y = PC2, color = tissue)) + 
  geom_point() +
  # Add this so our color palette is colorblindness friendly
  colorblindr::scale_color_OkabeIto()
```

## t-SNE experiment

t-SNE is a popular dimension reduction technique that has become even more 
popular with single-cell RNA-seq data analysis. 
It would be a good use of our time to explore t-SNE parameters and how they 
affect the outcomes.
First, let's run our data through t-SNE on default parameters and see what our 
data looks like with a `ggplot2` scatterplot.

```{r tSNE}
# Do tsne plots for each perplexity parameter, but this time use color argument
tsne_tab_mur <- Rtsn::Rtsne(tab_mur_data)

# Make this a data.frame so ggplot will like it
tsne <- data.frame(seurat_norm@dr$tsne@cell.embeddings)
  
# Plot this with ggplot2
ggplot(tsne, aes(x = tSNE_1, y = tSNE_2)) + 
  geom_point()
```

#### Apply metadata labels to our tSNE

Let's re-plot this with labels using some of our metadata to see 
how the cells are clustering (or not clustering) by different variables. 
Remember that previously we set up `sc_meta` to have the corresponding
metadata for each of our cells. 
Below we have set up the cells to be colored by the `cell.type` variable. 
Run this, but also feel free to look at what other information we have for these
cells in `sc_meta` by using `str()` and change this to a different variable to 
graph it by.

```{r}
# Plot this with ggplot2, but add the color argument
ggplot(tsne, aes(x = tSNE_1, y = tSNE_2, color = sc_meta$cell.type.ch1)) + 
  geom_point() +
  colorblindr::scale_color_OkabeIto()
```

Now that we have an idea of what the default parameters look like, let's try
experimenting with the perplexity parameter. 
For us to determine a good number range to use for testing perplexity, let's 
first look up the documentation for `Seurat::RunTSNE` and find out what the 
default for the `perplexity` argument is. 

In order to prepare for this experiment, we will create a *function* that allows
use to rerun this same code chunk easily, but create an argument that allows us
to change one variable: our perplexity variable. 

```{r Make tSNE plot function}
tSNE_plot_wrapper <- function(perplexity_param = 30) {
  # Purpose: Run t-SNE and plot the output
  # Args: perplexity_param: a single numeric argument that will change the 
  #                         perplexity variable in the RunTSNE function. 
  # Output: a ggplot scatterplot with the two t-SNE coordinates plotted and 
  #         cell-types labeled with data point colors. 

  # Run tSNE and plot it
  seurat_norm <- RunTSNE(object = seurat_norm, dims.use = 1:10, seed.use = 1234, 
                         perplexity = perplexity_param, do.fast = TRUE)

  # Make this a data.frame so ggplot will like it
  tsne <- data.frame(seurat_norm@dr$tsne@cell.embeddings)
  
  # Plot this with ggplot2
  ggplot(tsne, aes(x = tSNE_1, y = tSNE_2, color = sc_meta$cell.type.ch1)) +
    geom_point() +
    colorblindr::scale_color_OkabeIto()
}
```

Fill in the next five code chunks with the function and the perplexity argument 
you would like to use for each. 
Then run the chunks and compare your output graphs.

```{r Run tSNE 1}
# Use the tSNE_plot_wrapper function but alter the perplexity argument
tSNE_plot_wrapper(perplexity = 3)
```

```{r Run tSNE 2}
# Run tSNE_plot_wrapper function for a second time, but use a different perplexity
# argument
```

```{r Run tSNE 3}
# Run tSNE_plot_wrapper function for a third time, but use a different perplexity
# argument
```

```{r Run tSNE 4}
# Run tSNE_plot_wrapper function for a fourth time, but use a different perplexity
# argument
```

```{r Run tSNE 5}
# Run tSNE_plot_wrapper function for a fifth time, but use a different perplexity
# argument
```

You'll notice that for some settings of tSNE parameters, it's hard to distinguish
distinct clusters of cells.

#### Some 'big picture' thoughts to take from this experiment: 

1) Different analyses (such as t-SNE) have various limitations for interpretaility.
   We found in our experiment that the coordinates of tSNE output for any given 
   cell might change drastically, 
   This probably means that you shouldn't rely too heavily on the exact values of
   t-SNE's output. 
2) Different analyses also have their strengths.   
   Using cell-type labeling, our experiment illstrated that t-SNE does 
   appear to give some biologically relevant output information for this dataset. 
3) Playing with parameters so you can fine-tune them is a good way to give you more
   information about a particular analysis as well as the data itself. 
   
In summary, if the results of an analysis can be completely changed by changing it's
parameters, you should be more cautious when it comes to the conclusions you
draw from it as well as having good rationale for he parameters you choose. 

Read [this article](https://distill.pub/2016/misread-tsne/) by Wattenberg et al
to learn more about using t-SNE properly. 
  
t-SNE may be okay for some things, but note that the following dimension reduction 
methods may be worth considering: 

- [PCA](http://setosa.io/ev/principal-component-analysis/)  
- [UMAP](https://github.com/lmcinnes/umap)  
- [VASC](https://www.sciencedirect.com/science/article/pii/S167202291830439X)  
- [scVI](https://scvi.readthedocs.io/en/master/readme.html)  

#### Print session info:

```{r}
sessionInfo()
```
