---
title: "Intro to Tidyverse"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**CCDL 2019**

In this notebook, we will take you through the basics of the tidyverse: 
a collection of packages that make R use easier and more convenient.  

There are a lot of tutorials and free books Hadley Wickham and friends have on 
how to use the [tidyverse packages](https://r4ds.had.co.nz/index.html). 
[Datacamp](https://www.datacamp.com/) also has a bunch of good tutorials on this. 

## Set Up
R packages are collections of functions. 
Installing R on your computer installs `base` functions like statistical tests 
and some plotting functionality.
The `tidyverse` is a collection of packages that are handy for general data 
wrangling, analysis, and visualization. 
Other packages that are specifically handy for different biological analyses are 
found on [Bioconductor](https://www.bioconductor.org/).
If we want to use a package's functions we first need to install them.
In our Docker container, we already have `tidyverse` and other packages we will
use in this workshop installed for you. 
But if you needed to install it or other packages available on CRAN, you 
do it using the `install.packages` function like this: `install.packages("tidyverse")`
```{r Load tidyverse}
library(tidyverse)
```

"metadata.tsv" is a file we will use for practicing the tidyverse.
This metadata is from GSE30074 [refine.bio](https://www.refine.bio), a gene 
expression dataset from medulloblastoma.
  
First let's read in the dataset using the tidyverse package: `readr`.
Although base R has read-in functions, the `readr` functions are faster and more
straight forward to use so we are going to use those here. 
Because the file we are reading in is a .tsv file we will be using the 
`read_tsv` function. 
But note that there are analogous functions for .csv files (`read_csv`) and 
other files types.
```{r}
# The file we are reading in is our argument 
metadata <- read_tsv("metadata.tsv")
```

## dplyr pipes
A nifty thing with the tidyverse is the pipes: `%>%`
This handy thing allows you to funnel the result of one expression to the next
making your code a little more streamlined.

Let's look at an example of how this works by doing the same thing with and 
without a pipe:

The output from this:  
```
filter(metadata, sex == "F")
```  
...is the same as the output from this:  
```
metadata %>% filter(sex == "F")
```  
  
This makes it so your code is cleaner and easier to read. 
Let's look at an example with our metadata of of how the same 
functions looks with or without pipes:

*Example 1:* without pipes:   
```{r}
metadata.edit <- metadata
metadata.edit <- arrange(metadata.edit, histology)
metadata.edit <- filter(metadata.edit, follow_up_mths > 50)
metadata.edit <- select(metadata.edit, sample_id, sex, organism)
```
UGH, we have to repeat and assign `special.mtcars` so many times here! 
It's annoying and makes it harder for people to read. 

*Example 2:* Same result as 1 but with pipes!
```{r}
# Example of the same modifications as above but with pipes!
metadata.edit <- metadata %>%
                 arrange(histology) %>%
                 filter(follow_up_mths > 50) %>%
                 select(sample_id, sex, organism)
```

Now that hopefully you are convinced that tidyverse helps your code to be neater and
easier to use and read, let's go through some of the popular tidyverse functions
and so we can create pipelines like this. 

## Common tidyverse functions

Let's say we wanted to filter this gene expression dataset to particular sample
groups.
In order to do this, we would use the function `filter` as well as a logic statement.
```{r}
# Here let's filter the data to have only females
metadata %>% filter(sex == "F")
```

We can use filter similarly for numeric statements.  
```{r}
# Here let's filter the data to have people only older than 
metadata %>% filter(age_mths > 100)
```

What if we wanted filter this dataset to males who are younger than 100 months? 
Let's make a longer pipe.
```{r }
metadata %>% filter(age_mths > 100, sex == "M")
```

Let's build on to this pipe and keep only some of these variables with the 
`select` function. 
```{r}
metadata %>% filter(age_mths > 100) %>%
             filter(sex == "M") %>% 
             select(sample_id, histology, metastasis)
```

Let's say we wanted to arrange this dataset so that the patients were arranged
youngest to oldest.
In order to do this, we would use the function `arrange` as well as the variable
we would like to sort by (in this case `age_mths`).
```{r}
metadata %>% arrange(age_mths) 
```

What if we want the the oldest patients at the top of this dataset instead? 
We can use the same function, but instead use the `desc` function too. 
```{r}
metadata %>% arrange(desc(age_mths))
```

What if we want to obtain summary statistics based on groups? 
Conveniently named, there's a function called `group_by` that seamlessly allows
us to do this. 
Also note that you can group by multiple variables at a time if you want to.
```{r}
metadata %>% group_by(sex) %>% 
             summarize(mean(age_mths))
```
What if we would like to create new variables based on other variables? 
For that we use `mutate` function. 
Let's convert the ages that are in months to years and make that into a new variable.
Here we are also using the `round` function from base R to limit the number of 
decimal points in our new variable to 2.
```{r}
metadata.edit <- metadata %>% 
             mutate("age_yrs" = round(age_mths/12, 2)) %>%
             select(sample_id, age_mths, age_yrs)
```

Let's write this data to a file using write_tsv. Note that the second argument 
needs to be a character string that ends in the appropriate file extension ".tsv"
and it is what your file you are creating will be named. 
```{r}
write_tsv(metadata, "metadata_clean.tsv")
```

Session Info: 
```{r}
sessionInfo()
```
