---
title: "Intro to Tidyverse"
author: "CCDL for ALSF"
date: "1/28/2019"
output: html_document
---

This Rmd is taking you through the basics of the tidyverse: a collection of packages
that make R use easier and more convenient.  

There are a lot of tutorials and free books Hadley Wickham and friends have on 
how to use the [tidyverse packages](https://r4ds.had.co.nz/index.html). 
[Datacamp](https://www.datacamp.com/) also has a bunch of good tutorials on this. 

## Set Up
R packages are collections of functions. But your R as it is will not have these
functions available unless you tell it to.  
In our docker container, we already have tidyverse installed for you. 
But if you needed to install it or other pacakges available on CRAN, you 
do it using the `install.packages` function like this: `install.packages("tidyverse")`
```{r Load tidyverse}
library(tidyverse)
```

"GSE30074.tsv" is a file we will use for practicing the tidyverse.
This data is from `refine.bio` and is gene expression data from medulloblastoma.
  
First let's read in the dataset using the tidyverse package: `readr`.
Although base R has read-in functions, the `readr` functions are faster and more
straight forward to use so we are going to use those here. 
Because the file we are reading in is a .tsv file we will be using the 
`read_tsv` function. 
But note that there are analogous functions for `csv` files and other files types.
```{r}
# The file we are reading in, relative to our current directory, is in the "data"
# folder, so we need to put that as the argument. 
gene.data <- read_tsv("data/GSE30074.tsv")

# Let's also read in the metadata for this gene expression data
metadata <- read_tsv("data/metadata_clean.tsv")
```

## Dplyr pipes
A nifty thing with the tidyverse is the pipes: `%>%`
This handy thing allows you to funnel the result of one expression to the next
making your code a little more streamlined.
This makes it so your code is cleaner and easier to read. 
Let's look at an example with the built in data of `mtcars` of how the same 
functions looks with or without pipes:

*Example 1:* without pipes: 
```{r}
special.mtcars <- mtcars
special.mtcars$car_model <- rownames(special.mtcars)
special.mtcars <- arrange(special.mtcars, mpg)
special.mtcars <- filter(special.mtcars, mpg > 18)
special.mtcars <- select(special.mtcars, "mpg", "cyl", "gear")
```
UGH, we have to repeat and assign `special.mtcars` so many times here! 
It's annoying and makes it harder for people to read. 

*Example 2:* Same result as 1 but with pipes!
```{r}
# Example of the same modifications as above but with pipes!
special.mtcars <- mtcars %>% 
                  rownames_to_column("car_model") %>% 
                  arrange(mpg) %>%
                  filter(mpg > 18) %>% 
                  select("mpg", "cyl", "gear")
```

Now that hopefully you are convinced that tidyverse helps your code to be neater and
easier to use and read, let's go through some of the popular tidyverse functions
and so we can create pipelines like this. 

## Some Tidyverse Functions:

Let's say we wanted to filter this gene expression dataset to particular sample
groups.
In order to do this, we would use the function `filter` as well as a logic statement.
```{r}
# Here let's filter the data to have only females
metadata %>% filter(sex == "F")
```

We can use filter similarly for numeric statements.
```{r}
# Here let's filter the data to have people only older than 
metadata %>% filter(age_mths > 100)
```

What if we wanted filter this dataset to males who are younger than 100 months? 
Let's make a longer pipe.
```{r }
metadata %>% filter(age_mths > 100) %>%
             filter(sex == "M")
```

Let's build on to this pipe and keep only some of these variables with the 
`select` function. 
```{r}
metadata %>% filter(age_mths > 100) %>%
             filter(sex == "M") %>% 
             select(sample_id, histology, metastasis)
```

Let's say we wanted to arrange this dataset so that the patients were arranged
youngest to oldest.
In order to do this, we would use the function `arrange` as well as the variable
we would like to sort by (in this case `age_mths`).
```{r}
metadata %>% arrange(age_mths) 
```

What if we want the the oldest patients at the top of this dataset instead? 
We can use the same function, but instead use the `desc` function too. 
```{r}
metadata %>% arrange(desc(age_mths))
```

What if we want to obtain summary statistics based on groups? 
Conveniently named, there's a function called `group_by` that seamlessly allows
us to do this. 
Also note that you can group by multiple variables at a time if you want to.
```{r}
metadata %>% group_by(sex) %>% 
             summarize(mean(age_mths))
```
What if we would like to create new variables based on other variables? 
For that we use `mutate` function. 
Let's convert the ages that are in months to years and make that into a new variable.
Here we are also using the `round` function from base R to limit the number of 
decimal points in our new variable to 2.
```{r}
metadata %>% mutate("age_yrs" = round(age_mths/12, 2)) %>%
             select(sample_id, age_mths, age_yrs)
```

Let's do some work with the gene dataset now. 
First, let's rename the column that is clearly our ensembl gene ids, to a column 
name that reflects that. 
```{r}
gene.data <- gene.data %>% rename("genes" = X1)
```

Now let's obtain the average expression by gene. 
```{r}
gene.data <- gene.data %>% 
             mutate("gene_mean" = select(., contains("GSM")) %>% apply(., 1, mean))
```

Let's obtain the standard deviation value for each gene.
```{r}
gene.data <- gene.data %>% 
             mutate("gene_sd" = select(., contains("GSM")) %>% apply(., 1, sd))
```

Let's write this data to a file using write_tsv. Note that the second argument 
needs to be a character string that ends in the appropriate file extension ".tsv"
and it is what your file you are creating will be named. 
```{r}
write_tsv(gene.data, "gene_clean.tsv")
```

