---
title: "Prepare OpenPBTA medulloblastoma data"
output:
  html_notebook:
   toc: true
   toc_float: true
author: CCDL for ALSF
date: 2020
---

In this notebook, we'll prepare medulloblastoma data from the Open Pediatric Brain Tumor Atlas (OpenPBTA) project.

**Note that this notebook relies on the setup of the `machine-learning` module, where the OpenPBTA data is transformed (`../../machine-learning/setup/01-transform-rnaseq.Rmd`),
and the collapse step in the module itself  (`../../machine-learning/03-openpbta_PLIER.Rmd`). Therefore it is not intended for processing during training, but as part of development.**

## Set up

### Libraries

```{r}
library(tidyverse)
```

### Directories and files

#### Directories

We transformed these data in the `machine-learning` module set up.

```{r}
input_dir <- file.path("..", "..", "machine-learning", "data",
                       "open-pbta", "processed")

# We'll store the output of this notebook in a data directory in the pathway
# analysis module
output_dir <- file.path("..", "data", "open-pbta")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

#### Input files

```{r}
histologies_file <- file.path(input_dir, "pbta-histologies-stranded-rnaseq.tsv")
rnaseq_file <- file.path(input_dir, "pbta-vst-stranded-collapsed.tsv")
```

#### Output files

```{r}
output_metadata_file <- file.path(output_dir, "medulloblastoma_histologies.tsv")
output_rnaseq_file <- file.path(output_dir, "medulloblastoma_vst_collapsed.tsv")
```

## Read in and prepare data

We're going to simply limit the two files to medulloblastoma samples.

```{r}
histologies_df <- read_tsv(histologies_file)
rnaseq_df <- read_tsv(rnaseq_file)
```

Filter both data frames to only the medulloblastoma samples.

```{r}
mb_histologies_df <- histologies_df |>
  filter(short_histology == "Medulloblastoma") |>
  # Remove columns that are all NA -- typically columns used for info about
  # other histologies
  discard(~ all(is.na(.)))

# Use the biospecimen identifiers (Kids_First_Biospecimen_ID) from metadata
# data frame to filter columns of the RNA-seq data
mb_rnaseq_df <- rnaseq_df |>
  select(gene_symbol, mb_histologies_df$Kids_First_Biospecimen_ID)
```

Write to file.

```{r}
write_tsv(mb_histologies_df, output_metadata_file)
write_tsv(mb_rnaseq_df, output_rnaseq_file)
```

## Session Info

```{r}
sessionInfo()
```

