---
title: "Reading and filtering scRNA-seq data"
author: Data Lab for ALSF
date: 2022
output:
  html_notebook: 
    toc: true
    toc_float: true
---

## Objectives

This notebook will demonstrate how to:

- Read Cell Ranger data into R, including CITE-seq data
- Filter to cells using `emptyDropsCellRanger()`
- Calculate quality control measures
- Remove likely compromised cells with `miQC()`

---


In this notebook, we will



## Set Up

```{r setup}
# Load libraries
library(magrittr)
library(SingleCellExperiment)


# Setting the seed for reproducibility
set.seed(12345)
```

### Directories and files

The data we will be using for this module comes from a a 10X Genomics data set of expression data from  human peripheral blood mononuclear cells collection from a healthy volunteer.
The cells were labeled with a panel of a panel of nine [TotalSeq™-B antibodies (BioLegend)](https://www.biolegend.com/en-us/products/totalseq-b-human-tbnk-cocktail-19043).

The gene expression data and and antibody tag counts were processed with Cell Ranger 6.0.
We have provided the raw data folder 


```{r filepaths}
# main data directory
data_dir <- file.path("data", "PBMC-TotalSeqB")

# Path to the Cell Ranger matrix
raw_matrix_dir <- file.path(data_dir, "raw_feature_bc_matrix")

# reference data directory
ref_dir <- file.path("data", "reference")
# Path to mitochondrial genes table
mito_file <- file.path(ref_dir, "hs_mitochondrial_genes.tsv")

# Directory and file to save output
normalized_dir <- file.path(data_dir, "normalized")
if (!dir.exists(normalized_dir)) {
  dir.create(normalized_dir, recursive = TRUE)
}

output_sce_file <- file.path(normalized_dir, "pbmc_totalseq_sce.rds")

```


## Reading Cell Ranger data

```{r}
raw_sce <- DropletUtils::read10xCounts(raw_matrix_dir) 
```

Look at the object:
```{r}
raw_sce
```

36611 features, 2421702 droplet barcodes.

### Separating gene expression from antibody tags

```{r}
head(rowData(raw_sce))
```
```{r}
table(rowData(raw_sce)$Type)
```


```{r}
# Split out the ADT into an altExp
raw_sce <- splitAltExps(raw_sce, 
                        # "Gene Expression" or "Antibody Capture"
                        rowData(raw_sce)$Type, 
                        # The Type which should be the main experiment
                        "Gene Expression")
```

```{r}
raw_sce
```
```{r}
altExpNames(raw_sce) <- "ADT"
```

```{r}
dim(raw_sce)
```
```{r}
dim(altExp(raw_sce, "ADT"))
```


### Filtering empty droplets

Most droplets are empty

```{r}
droplet_df <- DropletUtils::emptyDropsCellRanger(
  counts(raw_sce), 
  BPPARAM = BiocParallel::MulticoreParam(4)
)


cells_to_retain <- which(droplet_df$FDR <= 0.01)
filtered_sce <- raw_sce[, cells_to_retain]
```



