---
title: "Reading and filtering scRNA-seq data"
author: Data Lab for ALSF
date: 2022
output:
  html_notebook: 
    toc: true
    toc_float: true
---

## Objectives

This notebook will demonstrate how to:

- Read Cell Ranger data into R, including CITE-seq data
- Filter to cells using `emptyDropsCellRanger()`
- Calculate quality control measures
- Remove likely compromised cells with `miQC()`

---


In this notebook, we will



## Set Up

```{r setup}
# Load libraries
library(magrittr)
library(ggplot2)
library(SingleCellExperiment)


# Setting the seed for reproducibility
set.seed(12345)
```

### Directories and files

The data we will be using for this module comes from a a 10X Genomics data set of expression data from  human peripheral blood mononuclear cells collection from a healthy volunteer.
The cells were labeled with a panel of a panel of nine [TotalSeq™-B antibodies (BioLegend)](https://www.biolegend.com/en-us/products/totalseq-b-human-tbnk-cocktail-19043).

The gene expression data and and antibody tag counts were processed with Cell Ranger 6.0.
We have provided the raw data folder 


```{r filepaths}
# main data directory
data_dir <- file.path("data", "PBMC-TotalSeqB")

# Path to the Cell Ranger matrix
raw_matrix_dir <- file.path(data_dir, "raw_feature_bc_matrix")

# reference data directory
ref_dir <- file.path("data", "reference")
# Path to mitochondrial genes table
mito_file <- file.path(ref_dir, "hs_mitochondrial_genes.tsv")

# Directory and file to save output
normalized_dir <- file.path(data_dir, "normalized")
if (!dir.exists(normalized_dir)) {
  dir.create(normalized_dir, recursive = TRUE)
}

output_sce_file <- file.path(normalized_dir, "pbmc_totalseq_sce.rds")

```


## Reading Cell Ranger data

```{r}
raw_sce <- DropletUtils::read10xCounts(raw_matrix_dir) 
```

Look at the object:
```{r}
raw_sce
```

36611 features, 2421702 droplet barcodes.

### Separating gene expression from antibody tags

```{r}
head(rowData(raw_sce))
```
```{r}
table(rowData(raw_sce)$Type)
```


```{r}
# Split out the ADT into an altExp
raw_sce <- splitAltExps(raw_sce, 
                        # "Gene Expression" or "Antibody Capture"
                        f = rowData(raw_sce)$Type, 
                        # The Type which should be the main experiment
                        ref = "Gene Expression")
```

```{r}
raw_sce
```
```{r}
altExpNames(raw_sce) <- "ADT"
```

```{r}
dim(raw_sce)
```
```{r}
dim(altExp(raw_sce, "ADT"))
```


### Filtering empty droplets

Most droplets are empty

```{r}
droplet_df <- DropletUtils::emptyDropsCellRanger(
  counts(raw_sce), 
  BPPARAM = BiocParallel::MulticoreParam(4)
)

cells_to_retain <- which(droplet_df$FDR <= 0.01)
filtered_sce <- raw_sce[, cells_to_retain]
```

```{r}
filtered_sce
```
### Calculating QC statistics

Get the mitochondrial genes
```{r}
mito_genes <- readr::read_tsv(mito_file) %>%
  dplyr::filter(gene_id %in% rownames(filtered_sce)) %>%
  dplyr::pull(gene_id)
```


```{r}
filtered_sce <- scuttle::addPerCellQC(filtered_sce,
                                      subsets = list(mito = mito_genes))
```

```{r}
cell_stats <- colData(filtered_sce) %>%
  as.data.frame()
cell_stats
```

### Filtering compromised cells


```{r}
ggplot(cell_stats, aes(x = detected, y = subsets_mito_percent)) +
  geom_point(alpha = 0.2)
```

Fit a mixture model 
```{r}
miqc_model <- miQC::mixtureModel(filtered_sce) 
```

Plot the model results
```{r}
miQC::plotModel(filtered_sce, miqc_model)
```


```{r}
miQC::plotFiltering(filtered_sce, miqc_model)
```
```{r}
qcfiltered_sce <- miQC::filterCells(filtered_sce, model = miqc_model)
```


### Filtering by ADT results

We're not going to talk about this in depth, since it's still an active area of study! 
But, be aware that RNA and ADTs are treated differently. 
As always we recommend [Orchestrating](http://bioconductor.org/books/3.16/OSCA.advanced/integrating-with-protein-abundance.html).

### Removing negative controls




One quick option is the `DropletUtils::cleanTagCounts()` function, which will flag cells for removal if there is no detectable ambient concentration of tags.
Cells with "unusually" high values of an ambient scaling factor also get flagged for removal since the those droplets are more likely to be affected by protein aggregates. 

In this case, we are going to take advantage of the fact that we still have the raw data that contains "empty" droplets with no cells to estimate the expected ambient profile, and then use the relative values to identify outliers.

If empty droplets are not available, you can leave off the `ambient` argument and `DropletUtils::cleanTagCounts()` will estimate an ambient profile based on the assumption that cells either do or do not express each surface protein, which should result in a bimodal distribution of tag counts.

```{r}
adt_ambient <- DropletUtils::ambientProfileEmpty(altExp(raw_sce))
```


```{r}
# Run QC on the tags
adt_qc_df <- DropletUtils::cleanTagCounts(
  altExp(qcfiltered_sce),
  ambient = adt_ambient
)
adt_qc_df
```

We can look at how many cells will be removed by each of these two measures.
```{r}
sum(adt_qc_df$discard)
sum(adt_qc_df$zero.ambient)
sum(adt_qc_df$high.ambient)
```

We definitely want to remove any cells with zero counts, as those could cause trouble later.
Whether you believe the upper outliers is probably worth more exploration, but we will remove them as well for simplicity.

```{r}
qcfiltered_sce <- qcfiltered_sce[, which(!adt_qc_df$discard)]
```


## Normalization and Dimension Reduction

### RNA-seq normalization

We will start with normalizing the RNA-seq data.
The goal here is to convert the raw counts into a measure that accounts for differences in sequencing depth between cells, and to convert the distribution of expression values from the skewed distribution we expect to see in raw counts to one that is more normally distributed.  

```{r}
# Normalize RNA
qclust <- scran::quickCluster(qcfiltered_sce)
qcfiltered_sce <- scran::computeSumFactors(qcfiltered_sce, clusters = qclust)
normalized_sce <- scuttle::logNormCounts(qcfiltered_sce)
```

This creates a new "assay" in the `normalized_sce` object, `logcounts`, which contains the normalized count values.

```{r}
normalized_sce

```

### ADT Normalization

```{r}
adt_sf <- scuttle::medianSizeFactors(
  altExp(normalized_sce), 
  reference = adt_ambient )
sum(adt_sf == 0)

```
```{r}
altExp(normalized_sce) <- scater::logNormCounts(
  altExp(normalized_sce), 
  size.factors = adt_sf
)
```


### Dimension Reduction

HVGs:
```{r}
num_genes <- 2000

gene_variance <- scran::modelGeneVar(normalized_sce)
hv_genes <- scran::getTopHVGs(gene_variance,
                              n = num_genes)
```

PCA:
```{r}
# Do we want to set the number of PCs?
normalized_sce <- scater::runPCA(normalized_sce,
                                 subset_row = hv_genes)
```

UMAP:
```{r}
normalized_sce <- scater::runUMAP(normalized_sce,
                                  dimred = "PCA")
```



## Unsupervised clustering
```{r}
nn_k <- 20
nn_clusters <- bluster::clusterRows(
  reducedDim(normalized_sce, "PCA"), 
  bluster::NNGraphParam(
    k = nn_k, 
    type = "jaccard",
    cluster.fun = "louvain",
    BPPARAM = BiocParallel::MulticoreParam(4)
  )
)

normalized_sce$nn_cluster <- nn_clusters
```

```{r}
scater::plotUMAP(normalized_sce,
                 colour_by = "nn_cluster")
```

## Save normalized results

```{r}
saveRDS(normalized_sce, file = output_sce_file)
```



## Session Info

```{r session info}
sessionInfo()
```

