---
title: "Northcott Medulloblastoma: Data preparation"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni and C. Greene 2018**

In this notebook, we'll cover how to do conversion between different gene
identifiers.
Different tools use different ids.
[`PLIER`](https://github.com/wgmao/PLIER) 
([Mao, et al. _bioRxiv._ 2017.](https://doi.org/10.1101/116061)), 
which we'll use shortly, 
requires that we use gene symbols and the refine.bio processed data uses
Ensembl gene ids.

For this part of the workshop, we'll be using [`GSE37382`](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE37382),
primary medulloblastoma sample data from 
[Northcott, et al. _Nature._ 2012.](https://www.ncbi.nlm.nih.gov/pubmed/22832581).

## Libraries and functions

We'll need something that maps between different types of identifiers to
do the conversion.
Here, we'll use the `org.Hs.eg.db` package from [Bioconductor]().

For more information about using Bioconductor annotation packages, see 
[this vignette](http://bioconductor.org/packages/release/bioc/vignettes/AnnotationDbi/inst/doc/IntroToAnnotationPackages.pdf) about AnnotationDbi.

```{r}
library(org.Hs.eg.db)
```

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%` 
```


## Files and directories

```{r}
# the directory where the refine.bio processed data lives, it's also where we 
# will write the output after the conversion
data_dir <- file.path("data", "GSE37382")
# refine.bio processed expression file
exprs_file <- file.path(data_dir, "GSE37382_SCAN.pcl")
# the file used for mapping between gene ids that is output
mapping_file <- file.path(data_dir, 
                          "genesymbol_ensembl_mappings_used_for_GSE37382.tsv")
# the prepped expression data output fule
output_file <- file.path(data_dir, "GSE37382_SCAN_symbol_mean_agg.pcl")
```

## Read in data

We'll read in the data with a [`data.table`](https://cran.r-project.org/web/packages/data.table/index.html) function because it is much faster than base R.

```{r}
exprs_df <- data.table::fread(exprs_file, data.table = FALSE)
```

Let's take a look at the expression `data.frame`.

```{r}
exprs_df[1:5, 1:5]
```

We can see that the first column, called `Gene`, contains the Ensembl gene ids.
Let's take a look at the dimensions of `exprs_df`.

```{r}
dim(exprs_df)
```

This means there are 21661 rows, which correspond to genes, and 285 samples
because the first of the 286 columns contain gene identifers.

## Annotation

### Conversion

`PLIER` uses gene symbols, but `org.Hs.eg.db` can be used for a number of 
different conversions. 
Let's take a look at what it contains.

```{r}
keytypes(org.Hs.eg.db)
```

If we had data that used Entrez gene ids, we could use `ENTREZID` to prep the
data for use with `PLIER`.

Now for the conversion bit! 
We can use [`mapIds`](https://www.rdocumentation.org/packages/AnnotationDbi/versions/1.34.4/topics/AnnotationDb-objects) 
to do the conversion.

Sometimes the gene id we're converting from will map to many ids of the type
we are converting to or vice versa.
`mapIds` will pick the first one it comes across when it's using the default
arguments or options (setting `multiVals = "first"`).
We don't necessarily want that behavior, so we can make a list.

```{r}
# don't replace the identifiers in exprs -- we're concerned about 1:many mappings
mapped_list <- mapIds(org.Hs.eg.db, keys = exprs_df$Gene, column = "SYMBOL", 
                      keytype = "ENSEMBL", multiVals = "list")
```

We can see that this 1:many mapping is the case.
`mapped_list` is a `list` where the names are Ensembl gene ids and the elements
are the gene symbols they map to.

```{r}
head(mapped_list)
```

Where are the 1:many mappings? 
Let's see what one of those looks like.

```{r}
list_index <- which(lapply(mapped_list, length) > 1)[2]
mapped_list[list_index]
```

How many of the mappings look like this?

```{r}
sum(lapply(mapped_list, length) > 1)
```

If we wanted to, we could manually resolve these by looking at other resources.

Let's get the data into a `data.frame` for joining.

```{r}
mapped_df <- reshape2::melt(mapped_list)
head(mapped_df)
```

Let's rename those columns.

```{r}
colnames(mapped_df) <- c("SYMBOL", "ENSEMBL")
head(mapped_df)
```

What happens to those 1:many mapping cases?

```{r}
mapped_df %>%
  dplyr::filter(ENSEMBL == "ENSG00000011454")
```

Let's write this to file for posterity.
If we find one gene we're really interested in, we'll want to go back to this
mapping file and make sure it's not ambiguous by using other resources.

```{r}
readr::write_tsv(mapped_df, path = mapping_file) 
```

Now we'll want to join the mappings to our expression data.

```{r}
# let's use the mappings to reannotate our data
annot_exprs_df <- mapped_df %>%
  # removing anything in the SYMBOL column that is NA
  dplyr::filter(!is.na(SYMBOL)) %>%
  # join by Ensembl gene ids, retaining only Ensembl gene ids in both
  dplyr::inner_join(y = exprs_df, by = c("ENSEMBL" = "Gene"))
```

What does this new `data.frame` look like?

```{r}
annot_exprs_df[1:5, 1:5]
```

We'll want to drop the `ENSEMBL` column.

```{r}
annot_exprs_df <- annot_exprs_df %>%
  dplyr::select(-ENSEMBL)
annot_exprs_df[1:5, 1:5]
```

Note that if Ensembl gene id mapped to more than one gene symbol, 
the expression value in that ENSG row was copied to each of the rows using
that gene symbol.
This should be okay for our purposes.

```{r}
annot_exprs_df %>%
  dplyr::filter(SYMBOL %in% c("GPR21", "RABGAP1"))
```

### Summarizing to mean

We'll want to check if any of the gene symbols are duplicated, because this 
will cause issues downstream.

If there are multiples of a gene symbol, let's take the average expression
value. 
Again, this should be okay for our particular use case.

```{r}
# how many of the symbols are duplicated?
sum(duplicated(annot_exprs_df$SYMBOL))
```

Let's take a look at an example.

```{r}
dup_index <- which(duplicated(annot_exprs_df$SYMBOL))[1]
dup_symbol <- annot_exprs_df$SYMBOL[dup_index]
annot_exprs_df %>%
  dplyr::filter(SYMBOL == dup_symbol)
```

```{r}
# checking for missing values -- we'll want to deal with these when we summarize
any(is.na(annot_exprs_df))
```

```{r}
# can collapse to mean values for duplicated genes without removing NAs, then
agg_exprs_df <- annot_exprs_df %>%
  # group by gene identifier
  dplyr::group_by(SYMBOL) %>%
  # for each column -- take the mean expression value
  dplyr::summarise_all(mean)
```

Let's take another look at the `r dup_symbol` example.

```{r}
agg_exprs_df %>%
  dplyr::filter(SYMBOL == dup_symbol)
```

Is this consistent with what we expect?

```{r}
# GSM917011
mean(c(-0.04392194, -0.05420797))
```

```{r}
# GSM917014
mean(c(0.08537644, 0.10227144))
```

Let's write our prepped data to file.

```{r}
# write to file!
readr::write_tsv(agg_exprs_df, path = output_file)
```

## Session Info

Record session info for reproducibility & provenence purposes.

```{r}
sessionInfo()
```


