---
title: "OpenPBTA: Create a heatmap"
output:   
  html_notebook: 
    toc: true
    toc_float: true
author: CCDL for ALSF
date: 2020
---

In this notebook, we cluster RNA-seq data from the Open Pediatric Brain Tumor Atlas (OpenPBTA) project and create a heatmap.
OpenPBTA is a collaborative project organized by the CCDL and the Center for Data-Driven Discovery in Biomedicine (D3b) at the Children's Hospital of Philadelphia conducted openly on GitHub.

You can read more about the project [here](https://github.com/alexslemonade/openpbta-analysis/#openpbta-analysis).

We've downloaded the publicly available data from the project and performed a variance stabilizing transformation on the stranded RNA-seq dataset.
(That code is available in `setup`!)

We'll use a package called [`ComplexHeatmap`](https://www.bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) to make our heatmap.
This package also will perform clustering as part of the heatmap.
It is highly flexible and opinionated - the data structures we pass `ComplexHeatmap` functions often need to be _just right_.
See the [`ComplexHeatmap` Complete Reference](https://jokergoo.github.io/ComplexHeatmap-reference/book/) for more information.

## Set up

### Libraries

```{r load_library}
# We'll be doing a bit of data wrangling with the Tidyverse at the outset
library(tidyverse)
# ComplexHeatmap is the package we'll use for making a heatmap
# It will do the hierarchical clustering for us as well
library(ComplexHeatmap)
```

### Directories and files

We have stored the data we'll use in this notebook in `data/open-pbta/processed`. 

```{r directories, live = TRUE}
data_dir <- file.path("data", "open-pbta", "processed")

# We'll store the heatmap in plots - create directory if it doesn't exist yet
plots_dir <- "plots"
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
```

#### Input files

```{r input_files}
histologies_file <- file.path(data_dir, "pbta-histologies-stranded-rnaseq.tsv")
rnaseq_file <- file.path(data_dir, "pbta-vst-stranded.tsv") 
```

#### Output files

```{r png_output, live = TRUE}
heatmap_file <- file.path(plots_dir, 
                          "common_histologies_high_variance_heatmap.png")
```

## Read in and prepare data

### Metadata 

Let's read in the metadata file and take a look.

```{r read_in_metadata, live = TRUE}
histologies_df <- read_tsv(histologies_file,
                           # this specifies how many lines should be used to
                           # guess column types - this file has well under
                           # 10000 lines
                           guess_max = 10000)   
```

Use the chunk below to explore the metadata data frame.

```{r explore_histologies_df, live = TRUE, eval = FALSE}
View(histologies_df)
```

We'll use the disease labels in a column called `short_histologies` when we label the heatmap.
Let's count how many samples are assigned each `short_histologies` label using the Tidyverse.

```{r count_histologies}
histology_count_df <- histologies_df %>%
  group_by(short_histology) %>%
  # Count how many samples are in each short_histology and name the column 
  # with that number n
  tally() %>%
  # Sort from largest number of samples to smallest number of samples in a 
  # histology
  arrange(desc(n))

histology_count_df
```

```{r filter_histologies, live = TRUE}
# Only include histologies with over 50 samples
histologies_to_keep <- histology_count_df %>%
  filter(n > 50) %>%
  pull(short_histology)

# Filter the full metadata file to only the histologies with n > 50
filtered_histologies_df <- histologies_df %>%
  filter(short_histology %in% histologies_to_keep)
```

### RNA-seq data

```{r read_in_rnaseq, live = TRUE}
# Read in and examine the RNA-seq data
rnaseq_df <- read_tsv(rnaseq_file)
head(rnaseq_df)
```

Samples are columns, rows are genes.
Gene identifiers are in the first column.
To subset the expression data to only the histologies we're interested in, we need to select the columns that correspond to the sample IDs in the now filtered metadata file.

The sample identifiers we need are in a column called `Kids_First_Biospecimen_ID`.

```{r look_at_sample_ids, live = TRUE}
head(filtered_histologies_df$Kids_First_Biospecimen_ID)
```

```{r create_rnaseq_mat}
# Filter to the samples in the histologies we're interested in
# and create a matrix where the rownames are the gene identifiers
rnaseq_mat <- rnaseq_df %>%
  select(gene_id,
         # In addition to *selecting* the samples in histologies that are in 
         # the filtered histologies data frame, this step also makes it such 
         # that the samples are in the same order in both data frames
         filtered_histologies_df$Kids_First_Biospecimen_ID) %>%
  tibble::column_to_rownames("gene_id") %>%
  as.matrix()
```

What are the dimensions of this RNA-seq data?

```{r rnaseq_dim, live = TRUE}
dim(rnaseq_mat)
```

Over 50k genes would be hard to put it the same heatmap!
If we are making a heatmap to get an idea of the structure in our data, it can be helpful to subset to high variance genes.
This is because the genes that *don't* vary are not likely to contribute much to the overall patterns we are interested in.

First, we'll calculate the variance for each gene using `rowVars()` from the `matrixStats` package and then take the genes in the top 10%.

```{r high_var_genes}
# Calculate variance
gene_variance <- matrixStats::rowVars(rnaseq_mat)
# Find the value that we'll use as a threshold to filter the top 10%
variance_threshold <- quantile(gene_variance, 0.9)
# Row indices of high variance genes
high_variance_index <- which(gene_variance > variance_threshold)
# What does a row index look like?
head(high_variance_index)
```

```{r high_var_mat, live = TRUE}
# Get a matrix that is subset to just the high variance genes
high_var_mat <- rnaseq_mat[high_variance_index, ]
```

## Heatmap

### Annotation

First, we'll set up the metadata that we want to use to label samples in the heatmap.
In `ComplexHeatmap` terminology, this is called annotation, or `HeatmapAnnotation`, specifically.

```{r sample_annotation_df}
sample_annotation_df <- filtered_histologies_df %>%
  # Select only the columns that we'll use
  select(Kids_First_Biospecimen_ID,
         short_histology,
         composition,
         tumor_descriptor)

# Let's examine these columns
sample_annotation_df
```

Right now, `sample_annotation_df` is a tibble because it's a filtered version of something we read in with `readr::read_tsv()`.
Usually this is pretty much equivalent to a data frame.

```{r sample_annotation_class}
class(sample_annotation_df)
```

BUT `ComplexHeatmap` expects a data frame, so sometimes it will behave in unexpected ways when you give it a tibble! 
Let's make a data frame and make the sample identifiers the rownames.

```{r annotation_as_df, live = TRUE}
sample_annotation_df <- sample_annotation_df %>%
  as.data.frame() %>%
  tibble::column_to_rownames("Kids_First_Biospecimen_ID")
```

To specify the colors in our annotation bar, we need to create a list of named vectors.
The names of the list need to *exactly* match the column names in `sample_annotation_df` and the names in each vector need to *exactly* match the values in those columns.

```{r color_palettes}
# Palette recommended for those with color vision deficencies - we'll use that 
# for histologies
histology_colors <- colorblindr::palette_OkabeIto[2:5]
sample_annotation_colors <- list(
  short_histology = c(
    "LGAT" = histology_colors[1],
    "Ependymoma" = histology_colors[2],
    "HGAT" = histology_colors[3],
    "Medulloblastoma" = histology_colors[4]
  ),
  composition = c(
    "Solid Tissue" = "#A0A0A0",  # light grey
    "Derived Cell Line" = "#000000"  # black
  ),
  tumor_descriptor = c(
    "Initial CNS Tumor" = "#3333FF",
    "Progressive" = "#FFFF99",
    "Recurrence" = "#CCCCFF",
    "Second Malignancy" = "#000033",
    "Unavailable" = "#FFFFFF"  # white for missing data
  )
)
```

We need to create a special type of object with `HeatmapAnnotation` using the annotation data frame and the list of color vectors.

```{r create_annotation, live = TRUE}
column_annotation <- HeatmapAnnotation(
  df = sample_annotation_df,
  col = sample_annotation_colors
)
```

### Values for display

We will z-score the expression values for display.
This is sometimes called a standardized score.

```{r row_normalize}
# PLIER::rowNorm z-scores the gene (row) values
zscores_mat <- PLIER::rowNorm(high_var_mat)
```

```{r}
zscores_alt <- (high_var_mat - rowMeans(high_var_mat)) / matrixStats::rowSds(high_var_mat)
```


Let's look at the mean and standard deviation of values for a randomly selected gene prior to standardization.

```{r random_row_mean}
mean(high_var_mat[25, ])
```

```{r random_row_sd}
sd(high_var_mat[25, ])
```

And now those values after standardization.

```{r random_z_mean, live = TRUE}
mean(zscores_mat[25, ])
```

This value isn't exactly zero but it's very close!

```{r random_z_sd, live = TRUE}
sd(zscores_mat[25, ])
```

The standard deviation is now 1.
That means all the variance values will be equal to one, too.

```{r summary_row_variance}
summary(matrixStats::rowVars(zscores_mat))
```

If you're filtering to high variance values, it's important to do that _prior_ to standardizing.

### Heatmap itself!

Okay, now we're ready to make a heatmap complete with annotation bars.

```{r make_heatmap}
# Assign the heatmap object
mb_heatmap <- Heatmap(zscores_mat,
                      # The distance metric used for clustering the rows
                      # This is different from the default (Euclidean)
                      clustering_distance_rows = "pearson",
                      # Linkage method for row clustering
                      # This is different from the default (complete)
                      clustering_method_rows = "average",
                      # Distance metric for columns
                      clustering_distance_columns = "pearson",
                      # Linkage for columns
                      clustering_method_columns = "average",
                      show_row_names = FALSE,
                      show_column_names = FALSE,
                      # Add annotation bars to the top of the 
                      top_annotation = column_annotation,
                      # This will be used as the label for the color bar
                      # of the cells of the heatmap itself
                      name = "z-score")

mb_heatmap
```

We have to display the heatmap by calling `mb_heatmap` _twice_ -- once for display in the notebook and once to save to a PNG file -- because we can't easily assign 

```{r save_heatmap_png, live = TRUE}
# Open PNG plot device
png(filename = heatmap_file, 
    width = 11, 
    height = 7, 
    units = "in", 
    res = 300)
# Heatmap object itself
mb_heatmap
# Shut down current graphics device
dev.off()
```

## Spotlight on medulloblastoma

The heatmap we made with all of the histologies under consideration suggested that there's some structure _within_ histologies.
Let's take a closer look at the medulloblastoma samples.

```{r mb_sample_ids, live = TRUE}
medulloblastoma_samples <- filtered_histologies_df %>%
  filter(short_histology == "Medulloblastoma") %>%
  pull(Kids_First_Biospecimen_ID)
```

Use all genes to create a matrix that only contains medulloblastoma samples.

```{r mb_mat, live = TRUE}
medulloblastoma_mat <- rnaseq_mat[, medulloblastoma_samples]
```

We'll use Pearson correlation for our distance metric here again, but let's break down what goes into hierarchical clustering a bit here.

```{r calculate_cor}
# cor() will take the correlation between columns - we're interested in the 
# relationship between medulloblastoma samples, so we don't need to transpose
# the matrix before using cor()
mb_sample_correlation <- cor(medulloblastoma_mat,
                             use = "pairwise.complete.obs",
                             method = "pearson")
```

Hierarchical clustering is performed on dissimilarities, so we'll subtract the correlation values from 1 -- now we have distance information.
To use `hclust()` -- the built-in method for hierarchical clustering in R -- we need a dissimilarity structure.
`as.dist()` returns a special object of class `dist`, a distance matrix used by `hclust()`.

```{r cor_as_dist, live = TRUE}
mb_sample_dist <- as.dist(1 - mb_sample_correlation)
```

Now we're ready to perform [hierarchical clustering](https://en.wikipedia.org/wiki/Hierarchical_clustering) with `hclust()`.
This is an _agglomerative_ method of clustering - each sample starts in its own cluster and then at each step of the algorithm the two most similar clusters are joined until there's only a single cluster left.

```{r mb_hclust, live = TRUE}
mb_sample_hclust <- hclust(mb_sample_dist, 
                           method = "average")
```

Here, `method = "average"` refers to the agglomeration or linkage method used during clustering. 
The method we're using is also known as [UPGMA (unweighted pair group method with arithmetic mean)](https://en.wikipedia.org/wiki/UPGMA) and is how we specify what formula to use for calculating inter-cluster distances during hierarchical clustering.
The distance between two clusters is the average distance between all pairs of points in the two clusters, where each pair is comprised of one point from each of the two clusters.

To look at the dendrogram produced by `hclust()`, we can use built-in graphics by calling `plot()`.

```{r mb_dendrogram}
plot(mb_sample_hclust, 
     # Put the labels at the same height
     hang = -1, 
     # Control the size of the text labels
     cex = 0.5) 
```

How many groups of medulloblastoma samples do you think there are?

## Session Info

```{r session_info}
sessionInfo()
```
