---
title: "Northcott Medulloblastoma: PLIER"
output:   
  html_notebook: 
    toc: true
    toc_float: true
---

**J. Taroni and C. Greene 2018**

In this notebook, we'll use a method called Pathway-Level Information Extractor
(PLIER) ([Mao, et al. _bioRxiv._ 2017.](https://doi.org/10.1101/116061)).

We like PLIER for a few reasons:

* It is a matrix factorization approach. 
  That means we can get a low-dimensional representation of our data.
  Specifically, PLIER learns correlated patterns of expression in our data
  or latent variables (LVs).
* It includes penalties such that _some_ of the LVs will align with gene sets 
  that we give it, so it's excellent for biological discovery.
* The authors demonstrated that it performs favorably wrt estimating proportion
  of immune cells in a sample as compared to other methods.
* Because not _all_ LVs align with the gene sets we input, some of them capture
  unwanted technical variation. 
  In our experience with the method, it does this quite well.

## Libraries and functions

```{r}
# PLIER library
library(PLIER)
# heatmap 
library(ComplexHeatmap)
```

```{r}
# magrittr pipe
`%>%` <- dplyr::`%>%`
```

## Directories and files

```{r}
# directory for the data
data_dir <- file.path("data", "GSE37382")
results_dir <- "results"
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
# directory where we will save any plots
plots_dir <- "plots"
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
# directory where we'll put the PLIER model
model_dir <- "models"
if (!dir.exists(model_dir)) {
  dir.create(model_dir, recursive = TRUE)
}
# expression data that we prepped in the first notebook
exprs_file <- file.path(data_dir, "GSE37382_SCAN_symbol_mean_agg.pcl")
```

**Output**

```{r}
# plier model output
plier_file <- file.path(model_dir, "GSE37382_PLIER_model.RDS")
# B matrix heatmap
b_heatmap_file <- file.path(plots_dir, "GSE37382_B_heatmap_significant.pdf")
```

## Read in expression data and prep for use with PLIER

```{r}
exprs_df <- data.table::fread(exprs_file, data.table = FALSE)
```

We'll need a matrix of expression data where the `rownames` are the gene 
symbols.

```{r}
exprs_mat <- as.matrix(tibble::column_to_rownames(exprs_df, "SYMBOL"))
exprs_mat[1:5, 1:5]
```

## Run PLIER

We'll set a seed using `set.seed`.
This makes our results reproducible.

```{r}
seed <- 12345
set.seed(seed)
```

### Gene sets for PLIER

The PLIER package comes with gene sets. 
We'll use the ones assocted with immune cell types (`bloodCellMarkersIRISDMAP` 
and `svmMarkers`) and the `canonicalPathways` (e.g., KEGG, REACTOME).
You can also use your own custom gene sets if desired.

```{r}
# load PLIER pathway and cell type data
data(bloodCellMarkersIRISDMAP)
data(svmMarkers)
data(canonicalPathways)
```

We'll use PLIER built in functions to prepare the gene sets for use.

```{r}
# combine the pathway data from PLIER
all_paths <- PLIER::combinePaths(bloodCellMarkersIRISDMAP, svmMarkers, 
                                 canonicalPathways)

# what genes are common to the pathway data and the expression matrix
cm_genes <- PLIER::commonRows(all_paths, exprs_mat)
```

### Row normalize

And a built-in function to row normalize the expression data.
This is z-scoring the expression values for each gene.

```{r}
# row normalize -- z-scoring the genes
exprs_norm <- PLIER::rowNorm(exprs_mat)
```

Let's compare the expression values from the first row (gene) before and after
the normalization. 

_Before_

```{r}
summary(exprs_mat[1, ])
```

```{r}
var(exprs_mat[1, ])
```

_After_

```{r}
summary(exprs_norm[1, ])
```
```{r}
var(exprs_norm[1, ])
```


### Set k

To define the number of latent variables (`k`), we'll need to first determine
the number of "significant principal components" using a built-in PLIER 
function.

```{r}
# what should we set the minimum k parameter to in PLIER? estimate the number 
# of PC for the SVD decomposition 
set_k <- PLIER::num.pc(exprs_norm[cm_genes, ])
```

### PLIER

Okay, now we're ready to run PLIER!
The authors recommend using the number of significant PCs + 30% for the
`k` argument.

```{r}
# PLIER main function + return results
plier_results <- PLIER::PLIER(exprs_norm[cm_genes, ], all_paths[cm_genes, ], 
                              k = round((set_k + set_k * 0.3), 0), trace = TRUE)
```

Let's save this to our model directory

```{r}
readr::write_rds(plier_results, path = plier_file)
```

## Exploring the PLIER model

`plotU` will give us insight into what pathways or cell types were captured by
the latent variables from the model.

```{r}
PLIER::plotU(plier_results, fontsize_col = 6, fontsize_row = 5)
```

We can save this as a PDF in our plots directory

```{r}
u_plot_file <- file.path(plots_dir, "GSE37382_U_plot.pdf")
pdf(u_plot_file)
PLIER::plotU(plier_results, fontsize_col = 6, fontsize_row = 5)
dev.off()
```

### What LVs have pathways or cell types associated with them?

PLIER returns a summary `data.frame` that will give us a bit more information.

```{r}
plier_summary_df <- plier_results$summary
plier_summary_df
```

We can filter this `data.frame` based on significance of associations.
We'll use `FDR < 0.05`.

```{r}
plier_summary_df %>%
  dplyr::filter(FDR < 0.05)
```

We can also sort by AUC values after filtering.

```{r}
plier_summary_df %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::arrange(dplyr::desc(AUC))
```

We can see very strong associations between the ribosome and LV90, for example.

Let's make a heatmap of the LV expression values, which can be found in 
`plier_results$B`

```{r}
plier_results$B[1:10, 1:5]
```

To make visualization a bit easier, let's limit our heatmap to only those 
LVs with a significant association with a gene set using the same FDR cutoff
as above and `AUC > 0.75`

```{r}
# first, let's subset summary to only the significant associations like above
sig_summary_df <- plier_summary_df %>%
  dplyr::filter(FDR < 0.05,
                AUC > 0.75)
# the index of the LVs we want will now be in the `LV index` column of this 
# data.frame -- but if an LV is associated with multiple gene sets we'll get
# duplicate values. unique helps us with that.
sig_index <- unique(sig_summary_df$`LV index`)
```

Let's get the LV expression value matrix out of `plier_results`

```{r}
b_matrix <- plier_results$B
sig_b_matrix <- b_matrix[as.integer(sig_index), ]
```

Let's make the heatmap itself

```{r}
pheatmap::pheatmap(sig_b_matrix, 
                   color = colorRampPalette(c("blue", "white", "red"))(100), 
                   scale = "row",
                   clustering_distance_cols = "correlation",
                   clustering_distance_rows = "correlation",
                   clustering_method = "average",
                   show_colnames = FALSE,
                   fontsize_row = 6)
```

Save this plot

```{r}
pdf(b_heatmap_file)
pheatmap::pheatmap(sig_b_matrix, 
                   color = colorRampPalette(c("blue", "white", "red"))(100), 
                   scale = "row",
                   clustering_distance_cols = "correlation",
                   clustering_distance_rows = "correlation",
                   clustering_method = "average",
                   show_colnames = FALSE,
                   fontsize_row = 6)
dev.off()
```

## Session Info

Record session info for reproducibility & provenence purposes.

```{r}
sessionInfo()
```