---
title: "Northcott Medulloblastoma: Group differences in LV expression"
output: html_notebook
---

**J. Taroni and C. Greene 2018**

In this notebook, we'll plot the differences in PLIER latent variable 
expression between the three subgroups: SHH, Group 3, and Group 4.

Our objective is to learn about melting wide data, wrangling data,
and using `ggplot2` along the way.

## First, a note on tidy data

In order to use `ggplot2`, we'll need the data in "long" or "tidy" format.
`PLIER` outputs what we want to plot in what we call "wide" format.

Read more about tidy data [here](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html).

To quote from Hadley Wickham's [R for Data Science](r4ds.had.co.nz/tidy-data.html):

> There are three interrelated rules which make a dataset tidy:
> * Each variable must have its own column.
> * Each observation must have its own row.
> * Each value must have its own cell.

Let's look at a toy example.

```{r}
set.seed(12345)
toy_df <- data.frame( 
  cbind(c("GENEA", "GENEB", "GENEC"),
        matrix(rnorm(30), ncol = 10))
)
colnames(toy_df) <- c("Gene", paste0("Sample", 1:10))
```

`toy_df` is now in "wide" format.

```{r}
toy_df
```

Let's get it into "long" format.

```{r}
toy_long_df <- reshape2::melt(toy_df, id.vars = "Gene", 
                              variable.name = "Sample", 
                              value.name = "Expression")
toy_long_df
```

Let's remove these toy examples from the workspace.

```{r}
rm(toy_df, toy_long_df)
```

## Libraries and functions

```{r}
library(ggplot2)
library(dplyr)
library(ggsignif)
```

## Directories and files

```{r}
# directory for the data
data_dir <- file.path("data", "GSE37382")
results_dir <- "results"
# directory where we will save any plots
plots_dir <- "plots"
# directory where we'll put the PLIER model
model_dir <- "models"
# metadata file that has been cleaned
metadata_file <- file.path(data_dir, "GSE37382_cleaned_metadata.tsv")
# PLIER model trained in third notebook
plier_file <- file.path(model_dir, "GSE37382_PLIER_model.RDS")
```

**Output**

```{r}
lv_plot <- file.path(plots_dir, "LV94_boxplot.pdf")
```

## Read in data

### Sample metadata

We'll use the sample metadata, which contains the subgroup information, in our
plotting.

```{r}
sample_meta_df <- readr::read_tsv(metadata_file)
```

### PLIER model

```{r}
plier_results <- readr::read_rds(plier_file)
```

We'll focus on the significant (`FDR < 0.05` and `AUC > 0.75`) latent variables
for now.
Let's get the indices of those LVs.

```{r}
sig_summary_df <- plier_results$summary %>%
  dplyr::filter(FDR < 0.05,
                AUC > 0.75)
sig_index <- unique(sig_summary_df$`LV index`)
```

Let's get the B matrix out of `plier_results` and filter it to only the 
significant LVs.

```{r}
b_matrix <- plier_results$B
sig_b_matrix <- b_matrix[as.integer(sig_index), ]
# add LV column
sig_b_wide <- tibble::rownames_to_column(as.data.frame(sig_b_matrix), 
                                         var = "LV")
```

### Tidy

Let's look at what format this is in currently.

```{r}
sig_b_wide[1:5, 1:5]
```

We want this in long format. 
We'll use `reshape2::melt` to do this just like in the toy example above.

```{r}
sig_b_df <- reshape2::melt(sig_b_wide, id.vars = "LV", variable.name = "Sample",
                           value.name = "Expression_level")
head(sig_b_df)
```

Let's add the sample metadata to this `data.frame` with `inner_join`.

```{r}
b_df <- inner_join(x = sig_b_df, y = sample_meta_df, 
                   by = c("Sample" = "source_name"))
head(b_df)
```

## Plotting

For illustrative purposes, we'll pick an LV and go through using `ggplot2` to
make a plot.

```{r}
lv <- "94,IRIS_Monocyte-Day7"
```

First, let's filter the `data.frame` to only observations of LV 94.

```{r}
lv94_df <- b_df %>% 
  dplyr::filter(LV == lv) 
```

#### Boxplot

Let's make a boxplot, using the subgroups as groups. 
We'll use defaults and then tweak different things one by one, so you can see
how different functions and arguments work.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot()
```

Let's add individual points with `geom_jitter`.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter()
```

Let's make the points more transparent by setting `alpha` and narrow the
width of the jitter with `width`

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5)
```

Let's change the theme to something more suitable for publication.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic()
```

Let's add labels and a title with the `labs` function.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages")
```

Let's center the title and make it bold with `theme` and `element_text`.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Let's put the x-axis labels on a 45 degree angle with `theme`.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(hjust = 1, angle = 45))
```

Let's get rid of the legend since it's redundant.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = "none")
```

Let's use colors that are consistent with our clustering notebook with
`scale_color_manual`.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = "none") +
  scale_colour_manual(values = c("#FFC125", "#00CD66", "#00BFFF"))
```

### Stats

Let's do a pairwise Wilcoxon rank sum test and put that information onto the
plot.

```{r}
pairwise.wilcox.test(x = lv94_df$Expression_level, g = lv94_df$subgroup,
                     p.adjust.method = "bonferroni")
```

We'll use the package `ggsignif` to add this information. 
This will automatically do the tests and add the p-values to the plot, but 
we have more control over how the test is done if we do things manually.

```{r}
lv94_df %>%
  ggplot(aes(x = subgroup, y = Expression_level, 
             group = subgroup, colour = subgroup)) +
  geom_boxplot(outlier.shape = NA) +  # we remove outlier points because they
  # will be included in the jitter
  geom_jitter(width = 0.175, alpha = 0.5) +
  # add built in theme
  theme_classic() +
  labs(x = "Subgroup", y = "LV Expression", title = "LV94 Macrophages") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        axis.text.x = element_text(hjust = 1, angle = 45),
        legend.position = "none") +
  scale_colour_manual(values = c("#FFC125", "#00CD66", "#00BFFF")) +
  geom_signif(y_position = c(1.3, 1.6, 1.9), 
              xmin = c("Group 3", "Group 3", "Group 4"), 
              xmax = c("Group 4", "SHH", "SHH"),
              annotation = c("0.00041", "3.7e-05", "0.00047"), 
              tip_length = 0, color = "black") 
```

### Save to file

`ggplot2` has a function `ggsave` that we will use to save a pdf of the last
plot we ran.

```{r}
ggsave(lv_plot, plot = last_plot())
```

## Session Info

Record session info for reproducibility & provenence purposes.

```{r}
sessionInfo()
```

