---
title: "Exercise: Exploring drug exposure in the NCI-60 cell lines"
output: 
  html_notebook: 
    toc: true
    toc_float: true
---

## NCI-60 Cell Line Data

In this exercise, we'll work with datafrom GEO accession [`GSE116436`](https://www.refine.bio/experiments/GSE116436/drug-induced-change-in-gene-expression-across-nci-60-cell-lines-after-exposure-to-15-anticancer-agents-for-2-6-and-24h)
that was processed with refine.bio. 

Here's an excerpt of the description from the authors:

> To identify patterns of drug-induced gene modulation that occur across different cell types, we measured gene expression changes across NCI-60 cell lines after exposure to 15 anticancer agents.

**We've supplied the following to you:**

* The gene expression matrix (`data/GSE116436/GSE116436_filtered.tsv`)
* The sample metadata (`data/GSE116436/metadata_GSE116436_cleaned.tsv`)
* Sample-sample Pearson correlation matrix (`data/GSE116436/GSE116436_sample_correlation.RDS`)
* PCA (`prcomp` output) (`models/NCI60_prcomp_results.RDS`)
* A PLIER model (`models/NCI60_PLIER_model.RDS`)

_Note that we've only made a subset of samples available and we've already
cleaned the metadata for you.
If you want to learn more about how the metadata was cleaned, ask an
instructor._

We're going to look specifically at **pathways that change with treatment**, 
but we precomputed the other objects in case you would like to understand what 
samples tend to group together or what is driving the variation in the data.

We'll use `dplyr` and `ggplot2` quite a bit in this notebook, so use the chunk
below to load them.

```{r}

```

The metadata is available at `data/data/GSE116436/metadata_GSE116436_cleaned.tsv`.
Read it in with `readr` and save it as `metadata_df`

```{r}
metadata_df <- 
```

What metadata do we have for these samples ? 
Use this chunk to explore `metadata_df`.

```{r}

```

## What pathways are perturbed by the drugs in this experiment?

The PLIER model we've precomputed is available `models/NCI60_PLIER_model.RDS`.
Read it in with `readr` and save it as `plier_results`.

```{r}
plier_results <-
```

The `summary` element of what is returned by PLIER can tell us about
associations, you can access it with `$` and it is a `data.frame`.

Use this chunk to view the PLIER model `summary`.

```{r}

```

Save the summary information as it's own data.frame called `summary_df`. 
We'll use this object to find out more about our model. 

```{r}
summary_df <-
```

The latent variables from this model will be associated with a pathway if the
gene expression data supports that there is some variation in the expression
levels of genes in that pathway.
It could be due to treatment with a drug or variation could arise from tissue- 
or cell line-specific differences.

### Is there a latent variable associated with the proteasome pathway?

We can use a chain of `dplyr` functions to filter `summary_df`.
Let's call `FDR < 0.05` associated; this is the same cutoff we used in the
PLIER notebook earlier.

```{r}
summary_df %>%
  dplyr::filter(,  # significant associations only
                # grepl returns TRUE when the pattern "proteasome" is found
                # in the pathway column
                grepl("proteasome", pathway, ignore.case = TRUE))
```

The LV index column of `summary_df` tells us _which row_ in the LV by sample
matrix `plier_results$B` we should look at.
To make plotting easier downstream, let's make a data.frame that contains two 
columns, `geo_accession` and `proteasome_lv` that contain the sample identifiers
and the values for the latent variable associated with the proteasome gene set, 
respectively.
Recall that the column names of `plier_results$B` are the sample identifiers.
(Hint: `?colnames`)

Call this data.frame `proteasome_df`.

```{r}
proteasome_df <- data.frame(geo_accession = ,
                            proteasome_lv = )
```

Now that you have `proteasome_df`, let's join it with our metadata using
an `inner_join` from `dplyr` to form a new data.frame called `prot_meta_df`.
We'll join by the sample names, which is what `by = "geo_accession"` specifies.

```{r}
prot_meta_df <- dplyr::inner_join(x = , 
                                  y = , 
                                  by = "geo_accession")
```

Let's find out what samples have the highest expression values for the
proteasome latent variable.
We can sort `prot_meta_df` with `arrange`.
Use the `Help` panel or `?arrange` to learn what to expect from this function.

```{r}
prot_meta_df %>%
```

What do you notice about the samples with the highest values?
Look up the mechanism of action for the drugs in `treatment`.
Do these results make sense?

### Line graph

Let's make a plot that shows the change of the proteasome latent variable
expression level over time, when treated with the drug that comes to the top
of the list.
We'll look at a single tissue of your choosing.
This requires use to do the following:

1. Filter `prot_meta_df` to samples from your tissue of interest and the
   drug that we are interested in
2. Make a line graph that tracks the expression value of the proteasome 
   latent variable during different time points of the experiment.

Save this plot as `p`.

```{r}
p <- prot_meta_df %>%
  # use these lines to filter to only the samples that are treated with the drug
  # of interest and to the tissue of your choice
  %>% 
  ggplot(aes(x = ,  # what needs to be on the x-axis?
             y = ,  # what needs to be on the y-axis?
             color = cell_line,  # color by individual cell line
             # we need to use group to make our line plot
             group = cell_line)) +
  # use facets to separate dosages
  facet_wrap(~ dose_nM) +
  # these two geoms make a line graph connecting individual points
  geom_point() +
  geom_line() +
  # use this scale for accessibility
  colorblindr::scale_color_OkabeIto()
```

Let's see what the plot looks like with this chunk below.

```{r}
p
```

**Based on what you know about the mechanism of action of this drug, do these 
results look how you expect?
What did the PLIER model uncover and what does the plot mean?**

#### Customizing the plot

We can add to `p` or tweak things by using the `+` operator without having to
repeat the code that set it up.
In the chunk below, we're changing the theme of the plot to `theme_bw()`.

```{r}
p + theme_bw()
```

Using the new theme, label the x-axis Time (Hours) and the y-axis Proteasome 
LV Expression Level.
Make the title the name of the drug you are plotting data for.

```{r}

```

If we wanted to change the spacing of the x-axis, such that the time points are
evenly spaced, we'd have to make `time_hours` into a factor.
_This is not always an appropriate visualization choice, it depends on your
experiment!_

Let's check what class `time_hours` currently is with the function `class`.

```{r}
class(prot_meta_df$time_hours)
```

Use `mutate()` and `as.factor()` to make the `time_hours` column into a
factor. 

```{r}

```

Now make a new plot, `p_factor`, using the same theme, color scale, labels, etc.

```{r}

```

Display `p_factor`

```{r}

```

Did it display in the order that you wanted?
If not, you can use the `levels` argument of the `factor()` function
to specify the order.
See [this post](https://rstudio-pubs-static.s3.amazonaws.com/7433_4537ea5073dc4162950abb715f513469.html).

## Additional analyses

Now that you've made and customized a line graph, explore the data in another
way:

* Try another latent variable, tissue, drug combination
* Make a boxplot of PC1 by tissue or treatment
* Make a heatmap of the sample-sample correlation matrix and use annotation
bars to help identify patterns
