---
title: "Importing, processing, and exploring Visium data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Read Visium data into R
- Filter to spots overlapping tissue
- Calculate quality control measures on spatial transcriptomic data
- Remove likely low-quality spots with `SpotSweeper()`
- Normalize spatial expression data
- Visualize spatial transcriptomic data 

## Introduction

In this notebook, we'll learn some basics of import, processing, and visualization of a Visium data set.

We will be using an anaplastic Wilms Tumor sample from the the [Single-cell Pediatric Cancer Atlas](https://scpca.alexslemonade.org/projects/SCPCP000006).
This sample was processed using first-generation Visium technology, and it was quantified with `Space Ranger 1.3.1`.
We will use the outputs from `Space Ranger` as our starting point for analysis in this notebook.


## Set up

To begin, we'll load `SpatialExperiment`, the core package for spatial transcriptomic objects in Bioconductor, as well as some packages for plotting.

```{r libraries}
# The main class we use for spatial transcriptomic data
library(SpatialExperiment)

# To support plotting
library(patchwork)

# Set ggplot2 theme
ggplot2::theme_set(ggplot2::theme_bw())
```

### Directories and files

Next we'll define input paths and files that we'll use in this notebook.

There are a variety of files that you will often see as part of the standard `Space Ranger` output.
The specific files present will depend on which specific spatial technology you are using, but they are all described in detail in [10x Genomics documentation](https://www.10xgenomics.com/support/software/space-ranger/latest/analysis/outputs/output-overview).

We've placed the core outputs you'll need for reading this first-generation Visium data into R in the directory `data/wilms-tumor/SCPCS000190/outs`, organized as `Space Ranger` returns them.
We'll define that input path, along with other inputs we'll need, here.

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "SCPCS000190"

# main data directory for this sample
data_dir <- file.path("data/wilms-tumor", sample_id)

# Path to Space Ranger output
raw_spaceranger_dir <- file.path(data_dir, "outs")

# reference data directory
ref_dir <- file.path("data/reference")

# Path to mitochondrial genes table
mito_file <- file.path(ref_dir, "hs_mitochondrial_genes.tsv")
```


And now, we'll define the output paths and files we'll use for exporting.

```{r outputs, live = TRUE}
# Directory and file to save output
normalized_dir <- file.path(data_dir, "normalized")

# create the directory if it does not exist
fs::dir_create(normalized_dir)

# output RDS file for normalized Spatial Experiment (SPE) object
# we'll use the existing sample_id variable to write the file name
output_spe_file <- file.path(
  normalized_dir,
  glue::glue("{sample_id}_normalized.rds")
)
```


## Reading Space Ranger data

Space Ranger output includes count data,  spatial information, and images.
The count data is generally provided in two main formats, just as `Cell Ranger` would provide them for single-cell data: 1) A folder with a feature list, a barcode list, and a sparse matrix in ["Matrix Exchange" format](https://math.nist.gov/MatrixMarket/formats.html), and 2) an `HDF5` formatted file containing all this information.
Because the data we're working with comes from the ScPCA Portal which only provides the folder format, that's what we'll read in here!

`Space Ranger` also outputs both filtered and raw matrices; today we will start with the raw matrix and perform all of our own filtering.

Before importing the data, let's go ahead and look what files we have:


```{r dir-spaceranger, live = TRUE}
dir(raw_spaceranger_dir)
```

As described, the `raw_feature_bc_matrix` and `filtered_feature_bc_matrix` directories are analogous to we might get from `Cell Ranger` for single-cell data.
However, **unlike** single-cell data where these barcodes correspond to droplets (some of which are actual cells), the barcodes in Visium data correspond to _spots_.
Here, each spot corresponds to a 55 µm diameter area on the slide that contains roughly 1-10 cells, depending on the tissue type being analyzed.
In this particular Visium data set, spots are laid out on a 6.5x6.5 mm slide (although other experiments may be 11x11 mm for larger tissue sections).

While the `raw_feature_bc_matrix` contains data for all possible spots, the `filtered_feature_bc_matrix` contains data only for spots that overlap tissue. 

Parallel to these directories is the `spatial` directory which contains files that provide the full spatial context to go along with the feature-by-barcode count matrices. 
Let's take a closer look at what's in the `spatial` directory:

```{r dir-spatial, live = TRUE}
dir(
  file.path(raw_spaceranger_dir, "spatial")
)
```

We see several different files here, which we'll go ahead and look at directly by clicking on them in the Files Pane.
Below are brief descriptions of each file we see here, and others that would be represented by more recent Visium technologies.

##### `"aligned_fiducials.jpg"`

This image file shows the provided tissue image with a border of the _fiducials_, which are small dots surrounding the slide capture area.
`Space Ranger` uses fiducials to anchor the quantified gene expression onto the tissue image itself, and this image can help you see how successful that _fiducial alignment_ was. 
If something looks really off in this image (e.g., it looks distorted, or some fiducials are missing), that's a sign you may need to manually adjust the alignment and run `Space Ranger` again.

##### `"detected_tissue_image.jpg"`

Part of `Space Ranger`'s algorithm involves identifying which spots overlay tissue and which do not.
This image file shows the Visium grid of spots overlaid on the tissue, where spots that overlay tissue are shaded in.
It's important to filter out spots that do not overlay tissue before proceeding with analysis.
This is because those spots will have 0 gene expression, but that doesn't mean that those genes weren't actually expressed elsewhere.
Those values of 0 mean that there was simply nothing to sequence, and we don't want them to contribute noise. 

##### `"tissue_hires_image.png"` and `"tissue_lowres_image.png"`

These are two different versions of the provided tissue image that have each been down-scaled for size to, respectively, ~2000 pixels and ~600 pixels.

##### `"tissue_positions_list.csv"`

This CSV file provides the tissue coordinates for each spot, indexed by its barcode.

##### `"scalefactors_json.json"` 

This JSON file stores the specific scaling factors used to downsize the provided tissue image to the provided image resolutions.


##### Other files you might see

In other Visium datasets, you may see more files in the `spatial/` directory than you see here.
In particular, this data was prepared with the first-generation Visium technology, in which tissue was manually placed directly onto a Visium capture slide.
Most current Visium experiments use an upgraded technology called CytAssist, an instrument that assists in transferring tissue from a standard slide to the Visium capture slide and dramatically facilitates FFPE tissue processing.

In Visium CytAssist datasets, there will be two additional image files you should be aware of:

- `"cytassist_image.tif"`: This is the image captured by the CytAssist instrument itself and is the primary image file that `Space Ranger` uses when processing CytAssist data
- `"aligned_tissue_image.jpg"`: If you provided a microscope image, such as an H&E or fluorescence image, along with the CytAssist image to a CytAssist run, `Space Ranger` will also return the microscope image aligned to the CytAssist image's coordinates

There may also be additional files, depending on which version of `Space Ranger` was used, but these are the core files to be aware of for first-generation Visium data and for CytAssist data.

### Import with `VisiumIO`

We'll begin by importing the raw counts and spatial information into a `SpatialExperiment` (SPE) object.
This object is a derived version of the `SingleCellExperiment` (SCE) object that contains a couple more slots that augment the object with spatial information.
This means there are some functions we'll use to interact with the object that you might recognize from your adventures with SCE objects, but there will also be some new functions that are just used for SPE objects.


![Structure of a `SpatialExperiment` object. Image from [Orchestrating Spatial Transcriptomics Analysis](https://bioconductor.org/books/release/OSTA)](diagrams/SpatialExperiment.png)
<br><br>

To import Visium data with the `VisiumIO` package, we'll need to specify three items (the rest should be automatically detected):

- The path to the `Space Ranger` count output, which in this case is the `raw_feature_bc_matrix` folder
- The path to the `Space Ranger` spatial output, which corresponds to the `spatial` directory
- Which image(s) should be imported into the object

By default, this function will try to import several images types, including the CytAssist image which we don't have here! 
So, in this case we'll specify to import only the `lowres` version of the image.


```{r import-visium, live = TRUE}
# path to quantified RNA 10X output
rna_dir <- file.path(raw_spaceranger_dir, "raw_feature_bc_matrix")

# path to spatial 10X output
spatial_dir <- file.path(raw_spaceranger_dir, "spatial")

# import to an object called spe
spe <- VisiumIO::import(
  VisiumIO::TENxVisium(
    # path that contains the sequencing data
    resources = rna_dir,
    # path that contains the spatial data
    spatialResource = spatial_dir,
    # which image(s) to import - we'll use lowres to save some memory/space
    images = "lowres"
  )
)
```

Let's have a look:

```{r print-spe, live = TRUE}
# print the spe
spe
```

First, have a look at the dimensions: We have 60664 genes (roughly a full transcriptome) and 4992 spots.
This 4992 is not a random value, but it's the exact the number of spots on a Visium 6.5x6.5 mm slide. 
In this raw object, some of these spots, but not all, overlap tissue and contain meaningful expression data.

Like SCE objects, this SPE object has slots like `assays` for count matrices, `rowData` for feature metadata, and `colData` for spot metadata.

Let's take a tour:

```{r spe-rowdata}
# print the rowData
rowData(spe)
```

```{r spe-assays}
# print the top corner of the count data
counts(spe)[1:10, 1:10]
```

```{r spe-coldata, live = TRUE}
# print the colData
colData(spe)
```
The `colData` shows us spot coordinate information on the Visium slide, as well as an `in_tissue` column --- it's an important one!
An `in_tissue` value of 0 indicates spots that do _not_ overlap tissue, so we'll need to filter those rows out before processing or analyzing this data further. 
`Space Ranger` determines which spots overlay tissue using the fiducial alignment frame, which we saw in this `"aligned_fiducials.jpg"` image. 

**Unlike** SCE objects, SPE objects contain two additional slots with spatial information.

First is a dedicated `spatialCoords` slot which contains the x and y coordinates on the slide in pixel units.
These are the coordinates that will be used when we plot this object.

```{r spe-spatialcoords, live = TRUE}
# handy function from our code package SpatialExperiment
spatialCoords(spe) |> head()
```

There is also a slot called `imgData` with a `DataFrame` of the imported images themselves, although looking at it directly isn't very interesting (but it does allow us to make pretty pictures!).
This `DataFrame` will have a row for each image imported, and in this case we only imported the `lowres` image.
We can also see its associated `scaleFactor`, which `VisiumIO` parsed from the input file `"spatial/scalefactors_json.json"`.

```{r spe-imgdata, live = TRUE}
imgData(spe)
```

### Introduction to spatial data visualization

Before we get any further, let's go ahead and actually look at our data.
Because the spatial and image information is contained in the object, we can plot directly from this object using the package [`ggspavis`](https://bioconductor.posit.co/packages/release/bioc/vignettes/ggspavis/inst/doc/ggspavis_overview.html).

This package contains a couple key functions we'll use to plot:

- `ggspavis::plotVisium()`: Plot the tissue H&E image, with or without spots
- `ggspavis::plotCoords()`: Plot the spots without the tissue

Let's try it out!

```{r plot-visium, live = TRUE}
#| fig.width: 7

ggspavis::plotVisium(spe)
```

Here, we see the spots overlaid on the slide as well as slide boundaries.
You'll notice that every single spot is shown, including ones that don't overlap tissue directly --- indeed, all 4992 spots are present in the object right now!

By default, `ggspavis::plotVisium()` shows the spots, but we can hide them and zoom into the tissue on the slide.
This is a helpful way to pop up the H&E for side-by-side comparisons with other plots we might make (foreshadowing!).

```{r plot-visium-zoom, live = TRUE}
#| fig.width: 5

ggspavis::plotVisium(
  spe, 
  spots = FALSE, # don't show the spots
  zoom = TRUE    # zoom into the slide area
) 
```

Any histology experts see patterns in this Wilms tumor you want to point out? 
We'll talk more about this slide later in the notebook, but for now, we'll just note that there seem to be a couple different tissue compartments distinguished by different nuclear densities. 

We can also plot only values from the spots themselves with `ggspavis::plotCoords()`.

```{r plot-coords, live = TRUE}
#| fig.width: 7

# plot just the spots
ggspavis::plotCoords(spe) 
```

Hm, the spot layout looks a bit different.
By default, this function only plot spots that overlay tissue (i.e. where `in_tissue` is 1) and ignore spots that do not overlay tissue.
We can override this if we want, but again, what we really want to do is actually remove those spots entirely.


## Filtering empty spots

We only care about spots on top of tissue, so let's begin by removing the `in_tissue = 0` spots.
Note that if we had read in the `filtered_feature_bc_matrix` instead, the data would already be filtered to only `1` in this column, but we need to take this step because we read in the `raw`. 
We can make a rough analogy, that this analysis step is like filtering empty droplets to retain only droplets with cells in scRNA-seq, but here we're retaining only spots over tissue.

We can visualize which spots those are, and we'll do it over the H&E to clearly see the relationship.
We'll use the `annotate` argument, but `plotVisium` sees that this column is an integer and forces it to use a continuous color scale; we'll go ahead and make it a factor version of it to plot with.

```{r plot_in_tissue}
#| fig.width: 7

# make a factor version of this column to plot with, specifying "No" and "Yes" plot labels
spe$in_tissue_factor <- factor(spe$in_tissue, levels = c(0, 1), labels = c("No", "Yes"))

ggspavis::plotVisium(
  spe, 
  annotate = "in_tissue_factor", 
  # specify a custom palette
  pal = c("Yes" = "yellow", "No" = "red")
) +
  # use ggplot2::guides() to override legend titles in ggspavis
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      title = "Spot overlaps tissue", 
      override.aes = list(size = 2)
    )
  )
```

- The purple tissue overhang on the left isn't colored at all - indeed, there aren't spots at those coordinates outside the slide
- Red points are those to filter out - they are uninformative since they don't overlap tissue.

```{r filter_in_tissue, live = TRUE}
# keep only spots that are in the tissue and save to filtered_spe
filtered_spe <- spe[, spe$in_tissue == 1]

# print resulting filtered_spe
filtered_spe
```

Now, we're down to 4210 spots from the original 4992, but that's still a pretty good amount. 
But not all of these spots are necessarily good quality, so we'll want to do some additional QC filtering next.

## Filtering low-quality spots

### Exploring global QC thresholds

As a first step towards filtering, we can borrow some approaches from scRNA-seq and calculate some quality-control measures. 
Let's do that and have a look - we'll get our mitochondrial genes for QC calculations and we'll use `scran::addPerCellQC()` like we might for single-cell (except here it adds per spot).

```{r get mitochondrial genes}
# read in a table of mitochondrial genes and extract ids
mito_genes <- readr::read_tsv(mito_file) |>
  # filter to only the genes that are found in our dataset
  dplyr::filter(gene_id %in% rownames(filtered_spe)) |>
  # create a vector from the gene_id column
  dplyr::pull(gene_id)
```

```{r calculate qc, live = TRUE}
filtered_spe <- scuttle::addPerCellQC(
  filtered_spe,
  subsets = list(mito = mito_genes)
)

# print resulting colData to see QC stats
colData(filtered_spe) |> head()
```

`ggspavis` comes with a helpful QC plotter (makes histograms by default but has a couple more options!).
These are `ggplot2` objects, so we can use `ggplot2` code with them like add a title to each.

```{r global qc distributions}
#| fig.width: 12
#| fig.height: 4

# Extract the colData into a data frame for plotting
coldata_df <- colData(filtered_spe) |>
  as.data.frame() |>
  # relocate barcodes to an actual column instead of rownames
  tibble::rownames_to_column("barcode") 

# Define density plots for each statistic
qc_plot_sum <- ggplot2::ggplot(coldata_df) + 
  ggplot2::aes(x = sum) + 
  ggplot2::geom_density(fill = "lightblue") +
  ggplot2::labs(title = "Total unique UMIs")

qc_plot_detected <- ggplot2::ggplot(coldata_df) + 
  ggplot2::aes(x = detected) + 
  ggplot2::geom_density(fill = "steelblue") +
  ggplot2::labs(title = "Total detected genes")

qc_plot_mito <- ggplot2::ggplot(coldata_df) + 
  ggplot2::aes(x = subsets_mito_percent) + 
  ggplot2::geom_density(fill = "slateblue") +
  ggplot2::labs(title = "Mitochondrial %")


# Plot together with patchwork
qc_plot_sum + qc_plot_detected + qc_plot_mito
```

These distributions all look unimodal without too much skew, which is different from how distributions from single-cell data might look. 

Placeholder for discussion about differences/similarities in technical artifacts between data types:

- One of the reasons we filter on these metrics in single cell is due to potential differences in cell capture across droplets
- This isn't a factor in spatial where QC stats are calculated per spot aka for aggregates of cells, extremes end up getting smoothed out
  - Differences in capture aren't due to heterogeneity from droplet capture but because of tissue biology/prep:
  - We have distinct regions of tissue or groups of cells that have less genes detected than other ones (biology) or because the tissue didn't lay on the slide completely flat or wasn't completely permeabilized in all spots equally

For example, let's consider the mitochondria percent distribution:

- We do see a long right-tail for mitochondrial percentages, but the values are all really low which doesn't suggest any major quality issues.
- The lower values here vs in single-cell data make sense because cells weren't stressed in the same way they would be for single-cell library prep; more likely to tell you about the tissue quality itself (biology) and not spot quality (technical) 

Let's plot the same stats on the slide.
We'll include the H&E plot here as well for an immediate comparison.

```{r global qc coords}
#| fig.width: 10
#| message: false


he_plot <- ggspavis::plotVisium(filtered_spe, spots = FALSE, zoom = TRUE) +
  ggplot2::ggtitle("H&E") +
  # add styling to match other panels
  ggplot2::theme(
    plot.title = ggplot2::element_text(
      hjust = 0.5, 
      margin = ggplot2::margin(0, 0, 0.1, 0)
    ),
    panel.border = ggplot2::element_rect(color = "black", linewidth = 0.25)
  )

qc_spots_sum <- ggspavis::plotCoords(filtered_spe, annotate="sum", point_size = 1) + 
  # use distinct palette
  ggplot2::scale_color_distiller(palette = "Blues", direction = 1) +
  ggplot2::ggtitle("Total unique UMIs")

qc_spots_detected <- ggspavis::plotCoords(filtered_spe, annotate="detected", point_size = 1) + 
  # use distinct palette
  ggplot2::scale_color_distiller(palette = "YlOrRd", direction = 1) +
  ggplot2::ggtitle("Total detected genes")

qc_spots_mito <- ggspavis::plotCoords(filtered_spe, annotate="subsets_mito_percent", point_size = 1) + 
  # use distinct palette
  ggplot2::scale_color_distiller(palette = "Greens", direction = 1) +
  ggplot2::ggtitle("Mitochondrial %")

# wrap plots with patchwork into 2x2 grid
patchwork::wrap_plots(
  he_plot, 
  qc_spots_sum, 
  qc_spots_detected, 
  qc_spots_mito, 
  nrow = 2
)
```

Now, we start to see there's more to the story.
There is spatial heterogeneity in these QC stats:

- the likely stromal regions have distinctly higher mitochondrial percentages (although again, these values are pretty low all around!) and fewer detected UMIs/genes
- the likely blastema regions tend to have more detected UMIs/genes

This tells us that there is local structure in the data that we might like our QC approach to take into consideration.
Using these global thresholds, we can see that QC is confounded by biology. 
If we filter these spots, we'll be removing spots associated with a particular tissue region.
This shows that global thresholds may not be suitable for data with this kind of heterogeneity, so a different approach could be warranted.
Worth noting, this is not strictly a spatial transcriptomics issue; there can be quite a bit of heterogeneity in scRNA-seq data too depending on what was sequenced!
Always plot your data!

### Filtering with local QC thresholds

Rather than finding outliers based on overall QC stats derived across biologically distinct domains/tissue regions, we can instead look for _local_ outliers to filter, i.e. spots with properties that deviate from their neighborhood of nearby spots.
Placeholder text to introduce [`SpotSweeper`](https://www.bioconductor.org/packages/3.22/bioc/vignettes/SpotSweeper/inst/doc/getting_started.html):

- Evaluates spots relative to their local neighborhood which is defined with `kNN`
- Calculates local means and variances of standard QC metrics to flag outliers for removal within a local tissue context
- Can be used to find both individual outlying spots as well as so-called "hangnails" which are regional artifacts caused by e.g. issues in tissue prep or other technical issues
  - Note that hangnail detection is performed using variance in mitochondrial percentages. 
  Since our values for mitochondrial percentage are very low with minimal variance, hangnail detection won't be relevant for us here

We'll detect local outliers based using `SpotSweeper::localOutliers()` which takes these arguments:

- `metric`: name of column in `colData` with QC stat of interest
- `direction`: are lower value or higher value outliers bad?
  - e.g. we'd say "lower" for detected (we don't want low detected values), and we'd say "higher" for mitochondrial percentage
- `n_neighbors`: number of neighbors for building the `kNN` graph; Default is 36
- `log`: whether to consider the `log1p` of the metric; Default is `FALSE`
  - The method assumes a normal distribution, so logging a metric may be needed to meet assumptions
  - logging "detected" and "sum" is often recommended because these distributions are unbounded and span orders of magnitude
  - logging "mitochondrial" is not often recommended because it's bounded `[0,100]` and logging percentages can distort the distribution
- `cutoff`: Threshold for identifying outliers; Default is 3, meaning 3 standard deviations from the mean (think z-scores)
  - Can make this lower to be more strict (more spots will get removed), lower to be less strict (more tolerant of outliers), but don't go too high or low or you're not really detecting outliers anymore
- `workers`: Number of cores for parallel processing; Default is 1

This function returns an `SPE` object with added `colData` columns we can use to find all the local outliers.

Let's start off by looking for library size ("sum" column) outliers:

```{r spotsweeper-sum, live = TRUE}
# specify defaults for sum to be explicit
filtered_spe <- SpotSweeper::localOutliers(
  filtered_spe, 
  metric = "sum", 
  direction = "lower", 
  n_neighbors = 36, 
  log = TRUE,
  cutoff = 3
)
```


What do we have now?

```{r coldata-spotsweeper-sum, live = TRUE}
colData(filtered_spe)
```
New columns include:

- `sum_log`: The `log1p()` (i.e., `log(x+1)`) of the `sum` column
- `sum_outliers`: Logical whether or not the spot is an outlier
- `sum_z`: The _local_ z-score for `sum_log`, among spots in the local neighborhood.
  - 0: same as neighbors on average
  - high positive value: much higher `sum` than neighbors
  - low negative value: much lower `sum` than neighbors, which in this case would be the ones to filter
  
Let's see some of this relationship - we'll plot `sum_log` (log of a globally-calculated stat) against `sum_z` (stat for _local_ outlier), with a guiding line for the cutoff.

```{r sum_log_z_plot}
filtered_spe_coldata <- colData(filtered_spe) |>
  as.data.frame()

ggplot2::ggplot(filtered_spe_coldata) + 
  ggplot2::aes(
    x = sum_log,
    y = sum_z,
    color = sum_outliers
  ) +
  ggplot2::geom_point(alpha = 0.5)
```

Based on this metric, most spots are not outliers, and while most of the spots classified as outliers are fairly close to the 3 cutoff, several are are more extreme outliers!
We also gain some insight about how global vs local filtering would play out.
Let's imagine we did global filtering with a threshold of 500, whose log is 6.21.
Looking at the x-axis here, that means we would only have removed three spots.
But with local filtering, we are catching additional spots that deviate from their local neighborhood and filtering a few more out, but on the whole this dataset seems to be fairly good quality so it's not a ton.

Let's plot this next to the H&E:
We'll use a plotting function from `SpotSweeper`, `SpotSweeper::plotQCmetrics()`, which will allow us to both color by a statistic and highlight outlier points.
Well plot the `sum_log` side-by-side with `sum_z` to compare.

Tip!
When using this function, the point color is a "fill" aesthetic (which can be customized), and the point outline is a "color" aesthetic.
This insight will help us to customize the plot colors

```{r spotsweeper-plotqc-sum}
#| fig.width: 10
#| message: false

sum_log_plot <- SpotSweeper::plotQCmetrics(
  filtered_spe, 
  metric = "sum_log", # how to *fill* spots
  outliers = "sum_outliers", # which spots to *color* with an outline
  stroke = 0.75 # size of outline
) +
  # customize the palette for filling spots
  ggplot2::scale_fill_viridis_c(option = "magma") +
  # customize the outlier outline
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent")) +
  ggplot2::ggtitle("Log of total UMI")


sum_z_plot <- SpotSweeper::plotQCmetrics(
  filtered_spe, 
  metric = "sum_z", # how to *fill* spots
  outliers = "sum_outliers", # which spots to *color* with an outline
  stroke = 0.75 # size of outline
) +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent")) +
  ggplot2::ggtitle("Local z-score for total UMI")

sum_log_plot + sum_z_plot
```
Here, we can see how spots are distributed across the slide.
Are there spots you would expect to be flagged for filtering based on the `sum_log` plot?
Are there spots you might not have expected to be filtered based on the `sum_log` plot?

#### Effect of neighborhood size

**TODO: Decide final placement for this section. May move to an exercise, or may shift to a more open-ended individual exploration.**
One key parameter that this relies on is the "neighborhood size" (`n_neighbors` argument).
How would this have changed if we shifted the neighborhood size?
To not clutter up our existing object, we'll save to new objects and plot as a exploration to learn about neighborhood size.

```{r spotsweeper-compare-n-neighbors}
#| fig.width: 10
#| message: false

# Detect outliers with a large neighborhood and make associated plot
filtered_spe_n100 <- SpotSweeper::localOutliers(
  filtered_spe, 
  metric = "sum", 
  direction = "lower", 
  n_neighbors = 100
)
outlier_plot_n100 <- SpotSweeper::plotQCmetrics(
  filtered_spe_n100, 
  metric = "sum_log",
  outliers = "sum_outliers", 
  stroke = 0.75 
) +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent")) +
  ggplot2::ggtitle("neighborhood k = 100")


# Detect outliers with a small neighborhood and make associated plot
filtered_spe_n10 <- SpotSweeper::localOutliers(
  filtered_spe, 
  metric = "sum", 
  direction = "lower", 
  n_neighbors = 10
)
outlier_plot_n10 <- SpotSweeper::plotQCmetrics(
  filtered_spe_n10, 
  metric = "sum_log",
  outliers = "sum_outliers",
  stroke = 0.75 
) +
  ggplot2::scale_fill_viridis_c() +
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent")) +
  ggplot2::ggtitle("neighborhood k = 10")


# Plot outliers detected with different neighborhood sizes
## TODO: Include all tested neighborhood values
outlier_plot_n100 + outlier_plot_n10
```

There's definitely some overlap, but also spots that are only filtered out for one of these neighborhood sizes.
In particular, the spot on the top left: With a large neighborhood, its low value is an outlier, but with a small neighborhood it is not an outlier because the immediately surrounding spots are also lower.

We can also see this difference by plotting the `sum_z` for each neighborhood size against each other on a scatterplot, coloring the global `sum_log`. 

```{r sum-z-scatterplot}
# create data frame for plotting
compare_z_df <- data.frame(
    n100_sum_z = filtered_spe_n100$sum_z, 
    n10_sum_z = filtered_spe_n10$sum_z, 
    # both versions have same value here since it's a global stat 
    sum_log = filtered_spe_n10$sum_log
  )

ggplot2::ggplot(compare_z_df) + 
  ggplot2::aes(x = n100_sum_z, y = n10_sum_z, color = sum_log) +
  ggplot2::geom_point(alpha = 0.6) + 
  ggplot2::labs(
    x = "n_neighbors = 100", 
    y = "n_neighbors = 10", 
    title = "sum_z scores between neighorbhood sizes"
  ) + 
  ggplot2::geom_hline(yintercept = -3, color = "blue") +
  ggplot2::geom_vline(xintercept = -3, color = "blue") +
  ggplot2::scale_color_viridis_c(option = "magma") +
  ggplot2::theme(aspect.ratio = 1)
```
The top right quadrant shows spots that both neighborhood sizes retain, and the bottom left quadrant shows spots that both neighborhood sizes will filter out.
The top left shows spots that will only be filtered out for `n_neighbors = 100`, and the bottom right shows spots that will only be filtered out for `n_neighbors = 10`.
Note there's at least one pretty low `sum_log` value retained in both versions.


The point is, changing the cutoff or changing the neighborhood size will influence which spots are flagged for removal - plot your data!
If you have specific biological intuition about how much variation you expect in the tissue, you might want to use that to inform cutoffs and the neighborhood size, and it's ok to explore.


#### Detected gene and mitochondrial outliers

Let's get back on track: let's do the same outlier detection for `detected` and `subsets_mito_percent`, and we'll proceed with the default cutoff of 3 with the default neighborhood size of 36.


```{r spotsweeper-detected, live = TRUE}
#| fig.width: 7

# Calculate local outliers for the `detected` metric
filtered_spe <- SpotSweeper::localOutliers(
  filtered_spe, 
  metric = "detected", 
  direction = "lower"
)

# Plot the slide showing outliers
SpotSweeper::plotQCmetrics(
  filtered_spe, 
  metric = "detected_log",
  outliers = "detected_outliers", 
  stroke = 0.75
) +
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent")) +
  ggplot2::scale_fill_viridis_c()
```


```{r spotsweeper-mito, live = TRUE}
#| fig.width: 7

# Calculate local outliers for the `subsets_mito_percent` metric
filtered_spe <- SpotSweeper::localOutliers(
  filtered_spe, 
  metric = "subsets_mito_percent", 
  direction = "higher", # we want to remove HIGHER values 
  log = FALSE, # don't log
  cutoff = 3
)

# Plot the slide showing outliers
SpotSweeper::plotQCmetrics(
  filtered_spe, 
  metric = "subsets_mito_percent",
  outliers = "subsets_mito_percent_outliers", 
  stroke = 0.75
) +
  # switch palette direction to make outliers are darker colors, matching other plots
  ggplot2::scale_fill_viridis_c(direction = -1) +
  ggplot2::scale_color_manual(values = c(`TRUE` = "black", `FALSE` = "transparent"))

```
Even though mitochondrial percentages are fairly low across all spots, this approach helps us find spots that are much lower within their neighborhood, all of which we would miss with global filtering (where we might pick a threshold of 15-25% for single-cell data).


Now that we've calculated all this, how many of each kind of outlier are there?

```{r count-local-outliers}
# Prepare data frame to count outliers
outliers_df <- colData(filtered_spe) |>
  as.data.frame() |>
  dplyr::select(ends_with("_outliers"))

# Count total outliers for each metric: The total TRUE values in the `sum_` columns
outliers_df |>
  dplyr::summarize(
    sum_outlier_count = sum(sum_outliers),
    detected_outlier_count = sum(detected_outliers),
    mito_outlier_count = sum(subsets_mito_percent_outliers)
  )
```

Let's visualize how these outliers relate to one another with a proportional Venn Diagram, using the `eulerr` package to help us:

```{r outliers-venn-diagram}
eulerr_data <- outliers_df |>
  # reorder & rename columns columns for display
  dplyr::select(
    mito_percent = subsets_mito_percent_outliers, 
    detected = detected_outliers, 
    sum = sum_outliers
  ) |>
  # venn diagram calculations
  eulerr::euler()

plot(
  eulerr_data, 
  quantities = TRUE, 
  legend = TRUE,
  labels = TRUE,
  fills = c("#F0E442", "#56B4E9", "#009E73")
)
```

Let's go ahead and filter all these spots out:

```{r build-local-filter}
# combine all outliers into "local_outliers" column to filter on
filtered_spe$local_outliers <- filtered_spe$sum_outliers |
  filtered_spe$detected_outliers |
  filtered_spe$subsets_mito_percent_outliers

# how many total outliers?
sum(filtered_spe$local_outliers)
```


```{r perform-local-filter, live = TRUE}
# keep only spots where `local_outliers` is FALSE
filtered_spe <- filtered_spe[, filtered_spe$local_outliers == FALSE]

# print resulting SPE
filtered_spe
```

After removing these additional 61 spots, we are left with 4149 spots that both overlap tissue and don't have obvious quality issues.


## Normalization

We'll proceed to normalize counts.
In single-cell data, we often use a deconvolution approach which does a quick clustering, calculates a scaling factor for each cluster, and then performs normalization using the cell's associated scaling factor. 
But, this isn't quite what we want for spot data since Visium spots are heterogeneous compositions of 1-10 cells.
So, we'll instead use a "library size normalization approach" which treats each spot independently using the `scuttle` package. 
It corrects for sequencing depth without smoothing out any heterogeneous signal.

It's worth noting this approach doesn't take spatial information directly into account; there are some approaches out there which account for spatial information, but this aspect of analysis seems to be pretty in flux. 
Also, bear in mind the caveat that in some cases library size itself is [also confounded with spatial structure](https://link.springer.com/article/10.1186/s13059-024-03241-7); so, keep your eyes peeled for normalization developments, since this space is evolving rapidly!

```{r normalize-spe, live = TRUE}
# calculate size factors
filtered_spe <- scuttle::computeLibraryFactors(filtered_spe)

# perform normalization
norm_spe <- scuttle::logNormCounts(filtered_spe)

# view normalized SPE
norm_spe
```


## Exploring gene expression

Finally, the moment you've all been waiting for: pretty plots!
Even though spots aren't exactly cells, we can still color by expression of genes of interest to get a general sense of patterns in our sample.
As a reminder, here's the H&E:

```{r show-he}
he_plot
```
Wilms Tumors are often composed of a couple histologic compartments: blastemal, stromal (with mesenchymal biology), and sometimes epithelial components.
We can see some of this biology reflected in this tissue section that appears to have these two major components: bluer regions indicating ECM-poor areas with high cellular density consistent with blastemal biology, and paler pink regions consistent with stromal tissue that may have an endothelial presence.

Let's see if we can detect regions that might represent these compartments using marker gene expression.
For this, we'll plot expression using `ggpsavis::plotCoords()` for two genes:

* `WT1`: A transcription factor that regulates kidney development.
We expect this gene to be highly expressed in Wilms Tumor blastemal compartments
* `COL1A1`: A collagen protein that we expect to be highly expressed in stromal/mesenchymal compartments

We'll need to provide the gene's corresponding row name, which is its Ensembl gene id, to `ggpsavis::plotCoords()` so let's go ahead and find those first.

The `rowData` slot helpfully provides a gene symbol/Ensembl id mapping, if you haven't memorized all the ids yet:

```{r show-rowdata}
rowData(spe) |> head()
```



```{r get-ensembl, live = TRUE}
# define gene symbols
gene_symbols <- c("WT1", "COL1A1")

# filter rowData to genes of interest
rowData(spe) |>
  as.data.frame() |>
  dplyr::filter(Symbol %in% gene_symbols)
```

Great, let's plot!
First `WT1`, which we'll show next to the H&E for context:

```{r plot-wt1, live = TRUE}
#| message: false

# save plot of WT1 to a variable
wt1_plot <- ggspavis::plotCoords(
  norm_spe, 
  annotate = "ENSG00000184937", # name of variable to color by. here, an ensembl id
  assay = "logcounts", # counts are plotted by default, so we need to override this
  point_size = 1.25
) + 
  ggplot2::scale_color_viridis_c() + # custom color scale
  ggplot2::labs(title = "WT1")

# plot h&e with WT1 plot
he_plot + wt1_plot 
```



What patterns do you see here?
Do you think there is potentially spatial structure to `WT1` expression?
Notice that little yellow band inside the dark region on top that matches nicely with the nucleated bit of that tissue image?
It's very satisfying.

Let's compare this to `COL1A1` expression, whose Ensembl id we determined is `ENSG00000108821`. 
Again, we'll plot this alongside the H&E.

```{r plot-col1a1}
#| message: false

# create COL1A1 plot and save to variable
col1a1_plot <- ggspavis::plotCoords(
  norm_spe, 
  annotate = "ENSG00000108821",
  assay = "logcounts",
  point_size = 1.25
) + 
  ggplot2::scale_color_viridis_c() + 
  ggplot2::labs(title = "COL1A1")

# plot H&E with COL1A1 plot
he_plot + col1a1_plot 
```

What differences do you see in gene expression compared to `WT1`?
Did you expect this based on the H&E?

Based on these (very few, very cursory, and highly exploratory!) plots, do you think there is evidence for blastemal and stromal compartments in this sample?


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

