---
title: "Spatially-aware clustering"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Perform PCA, clustering, UMAP sans spatial information
- Use BANKSY to perform spatially-aware PCA and clustering
- Compare clusters calculated with and without spatial information

## Introduction

We will be using a sample from a mouse with osteosarcoma lung metastases e from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049).
This sample was processed using second-generation Visium technology, and it was quantified with `Space Ranger 2.1.1`.

We have already processed this dataset for you: filtered (note that `SpotSweeper` removed 20 low-quality spots during this step) and normalized it. 

## Set up


```{r libraries}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
})


# To support plotting
library(ggspavis)
library(ggplot2)
library(patchwork)

# Set ggplot2 theme
theme_set(theme_bw())
```


Set a seed for reproducibility:

```{r}
set.seed(2026)
```


### Directories and files

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "GSM8478586"

# data directory for this sample
data_dir <- file.path("data/osteo", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "osteo_normalized_spe.rds")
```

And the output file:

```{r}
clustered_rds <- file.path(data_dir, "osteo_clustered_spe.rds")
```


## Import

```{r import-rds, live = TRUE}
spe <- readr::read_rds(normalized_rds)
```

Look at it first:

```{r print-spe, live = TRUE}
# print the spe
spe
```


```{r plot-visium}
#| fig.width: 5

# define H&E plot for general use in notebook
he_plot <- plotVisium(spe, spots = FALSE, zoom = TRUE) +
  ggtitle("H&E") +
  # add styling to match other panels
  theme(
    plot.title = ggplot2::element_text(
      hjust = 0.5, 
      margin = ggplot2::margin(0, 0, 0.1, 0)
    ),
    panel.border = ggplot2::element_rect(color = "black", linewidth = 0.25)
  )

he_plot
```

We can clearly tumors amid the lung parenchyma, which itself is more sparse because of the alveolar spaces which is just how lungs work.
Let's get a sense of the RNA counts here by plotting some of the QC stats we calculated for you.
Note that these plots use right-censoring and set an upper limit on the color scale to be able to see lower value spots more clearly.

```{r show qc}
#| warning: false
#| message: false

total_plot <- plotCoords(spe, annotate = "sum", point_size = 1.25) + 
  scale_color_viridis_c(
    # cap scale at 20,000; all value above this will be shown as 20,000
    limits = c(0, 20000), 
    oob = scales::oob_squish, 
    # fun with color scales!
    option = "plasma"
  )
detected_plot <- plotCoords(spe, annotate = "detected",  point_size = 1.25) + 
  scale_color_viridis_c(
    # cap scale at 10,000
    limits = c(0, 10000), 
    oob = scales::oob_squish, 
    option = "plasma"
  )
total_plot + detected_plot
```
Takeaways:

- Clearly, more expression in the tumors in general vs parenchyma, which makes sense biologically!
- Because we have really different tissue compartments here, if we had used global thresholds we would have just ended up retaining almost exclusively tumor cells and removing their spatial context due to lower expression, but we would like to retain this spatial context because otherwise we've just wasted a ton of money on Visium


## Calculate clusters

We're going to see how standard single-cell post-processing would look and then compare to one approach for injecting spatial information into the clustering procedure.

### Spatial-free, aka expression only

First, we'll run PCA, clustering, UMAP as we might do for single-cell.
Therefore, in this section, clusters will be driven by expression only but keep in mind that we're working at the spot level with Visium data, not the cell level, and each spot contains 5-20 cells.
This means that we're clustering heterogeneous groups of cells as single unit regions of expression as each spot may not represent a single cell (if you'd like to know many & what cells each spot might have, please keep attending the workshop!). 

First, PCA:

- Model gene variance and pull out the 2000 HVGs for use in PCA
  - Note that this is not spatially aware and there is a distinct problem "spatially variable genes"
- Then we'll use these genes to run the PCA

```{r nonspatial pca, live = TRUE}
gene_variance <- scran::modelGeneVar(spe)
highvar_genes <- scran::getTopHVGs(spe, n = 2000) 

spe <- scater::runPCA(spe, subset_row = highvar_genes)
```

Now, we can cluster.
For this notebook, we'll generally use 20 nearest neighbors with the Louvain algorithm for graph-based clustering.
Worth noting that it's best practice to exploring parameter space though, see [the clustering exercise](../scRNA-seq-advanced/exercise_05-cluster_evaluation.Rmd) to learn more about this.

```{r nonspatial cluster, live = TRUE}
# define clustering params, and we can re-use this later too
cluster_params <- bluster::NNGraphParam(
  k = 20,
  cluster.fun = "louvain"
)

# perform clustering
colData(spe)$cluster_nonspatial <- scran::clusterCells(
  spe, 
  use.dimred = "PCA", 
  BLUSPARAM = cluster_params
)
```


We'll also calculate the UMAP for viz.

```{r nonspatial umap}
spe <- scater::runUMAP(spe)
```

  
Let's have a look, both at the UMAP and at the coordinates colored by cluster.
We'll use the same palette to assist in interpretation, and show the H&E here as well.


```{r plot nonspatial}
#| message: false
#| fig.width: 10

umap_nonspatial <- scater::plotUMAP(
  spe, 
  color_by = "cluster_nonspatial",
  point_size = 0.5
) + 
scale_color_brewer(palette = "Dark2")

coords_nonspatial <- plotCoords(
  spe, 
  annotate = "cluster_nonspatial", 
  point_size = 1.25
) + 
  scale_color_brewer(palette = "Dark2")

umap_nonspatial + coords_nonspatial + he_plot
```
We see four clusters which have some correspondence to patterns in the H&E for tumor vs parenchymal regions, and clusters seem to be based on expression patterns along a "gradient" from tumor -> parenchymal.
Interestingly we see in the UMAP that spatial regions tend to be adjacent; whether this is meaningful isn't clear but it's neat nonetheless.


### Spatially-aware clustering with BANKSY

The field is popping for approaches to incorporate spatial information, and we're going to go through one popular method used for clustering and/or domain identification in spatial data.

Text placeholder for BANKSY methods, briefly...
- calculates a matrix ("BANKSY matrix") that considers both expression and spatial information
  - there are two frameworks that BANKSY uses to consider spatial information: graph-based neighbor averaging, and optionally, AGF on top of that, where the neighborhood sizes are controlled by a `k_geom` parameter
  - Incorporating AGF may help to detect boundaries between domains and can resolve some finer structure
  - At some point we have to note a fun naming convention: H0 and M0 refer to matrices calculated sans AGF, and H1 and M1 with AGF. The H/M distinction is a bit of mathematics pedantry but just remember 0 --> no AGF and 1 --> with AGF.
- Then we can perform PCA on this matrix, using a parameter called `lambda` to tune how much the spatial information contributes
- Finally we can cluster/run UMAP on this PCA

See more details about parameters [from these docs](https://prabhakarlab.github.io/Banksy/articles/parameter-selection.html#parameters).

We're going to run this without and with AGF and then compare all around.

#### Without AGF

```{r banksy matrix agf off, live = TRUE}
# define parameters recommended for Visium
lambda <- 0.2
k_geom <- 18

# calcuate the BANKSY matrix without AGF
spe <- Banksy::computeBanksy(spe, assay_name = "logcounts", k_geom = k_geom)

# show spe
spe
```

Show and explain the H0 matrix (spatially-smoothed neighbor-averaged expression).

```{r show ho}
assay(spe, "H0")[1:5, 1:5]
```
Now we can calculate PCA, specifying `lambda` for how to weight the original expression (`logcounts`) and the BANKSY matrix (here, H0, 

```{r banksy pca, live = TRUE}
# calculate PCA within BANKSY
spe <- Banksy::runBanksyPCA(spe, lambda = lambda)

# show SPE
spe
```
We have a new PCA called `PCA_M0_lam0.2`.
BANKSY named this for us: `M0` implies that AGF is off, and `lam0.2` is our lambda parameter.

We can proceed to calculate UMAP and clusters now, and we'll use `scran`/`scater` functions for that.
But, it's worth point out that BANKSY does allow you to specify multiple `lambda` parameters at once and then offers wrapper functions to cluster/UMAP across those parameters in one function call.

```{r cluster banksy nonagf}
pca_name <- "PCA_M0_lam0.2"
umap_name <- "UMAP_banksy_nonagf"

spe <- scater::runUMAP(
  spe,
  dimred = pca_name,
  name = umap_name
)

# perform clustering
colData(spe)$cluster_banksy_nonagf <- scran::clusterCells(
  spe, 
  use.dimred = pca_name,
  # previously defined as 20 nearest neighbors using the Louvain algorithm and default resolution 1 
  BLUSPARAM = cluster_params 
)
```

And plot:

```{r plot banksy nonagf}
#| message: false
#| fig.width: 10

umap_banksky_nonagf <- scater::plotReducedDim(
  spe, 
  dimred = umap_name,
  color_by = "cluster_banksy_nonagf", 
  point_size = 0.5
) + 
scale_color_brewer(palette = "Set1")


coords_banksky_nonagf <- plotCoords(
  spe, 
  annotate = "cluster_banksy_nonagf", 
  point_size = 1.25
) + 
scale_color_brewer(palette = "Set1")

umap_banksky_nonagf + coords_banksky_nonagf + he_plot
```

How does this compare to non-spatial?
Kind of neat to see that there is still spatial arrangement in the UMAP, although it's not clear what it means if anything!


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

