---
title: "Clustering spatial transcriptomics data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Perform dimensionality reduction (PCA and UMAP) and graph-based clustering on Visium data
- Use BANKSY to perform spatially-aware dimensionality reduction and clustering
- Compare clusters calculated with and without spatial information

## Introduction

In this notebook, we'll cluster expression data from a mouse sample with osteosarcoma lung metastases from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049).
This sample was processed using second-generation Visium technology and quantified with `Space Ranger 2.1.1`.
We have already processed this dataset for you, including filtering out low quality spots (`SpotSweeper` removed 20 low-quality spots) and normalization, which you can see in this notebook: <https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/osteo/prepare-osteo-data.Rmd>.

As we have discussed, Visium spatial transcriptomics data does not quantify expression at the single-cell level, but rather each Visium spot captures the combined expression of 5-20 cells.
This means that when we perform clustering on Visium data, we are actually clustering spots with similar transcriptional profiles, not individual cells.

Also unlike single-cell data, we now have an additional consideration of spatial information.
In many cases (but not all!), it's reasonable to assume that nearby spots will tend to have more similar expression to one another than to distant spots. 
We can therefore incorporate each spot's physical position in the tissue alongside its expression profile during clustering, which may help to approximate spatial domains. 


In this notebook, we'll explore complementary approaches: first running expression-based dimensionality reduction and clustering just as we would for single-cell data, and then performing the same analyses using the [`BANKSY` package](https://bioconductor.org/packages/release/bioc/html/Banksy.html). 
`BANKSY` offers a spatially-aware dimensionality reduction and clustering approach that augments each spot's expression with information from its spatial neighborhood.

## Set up

To begin, we'll load the packages we'll use throughout this notebook.

```{r libraries}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
})


# To support plotting
library(ggspavis)
library(ggplot2)
library(patchwork)

# Set ggplot2 theme
theme_set(theme_bw())
```


We'll set a seed for reproducibility, since we'll be running UMAP and clustering.

```{r}
set.seed(2026)
```


### Directories and files

The pre-processed SPE object for this sample is stored in a sample-specific subdirectory of `data/osteo`, organized by GEO sample ID.

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "GSM8478586"

# data directory for this sample
data_dir <- file.path("data/osteo", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "osteo_normalized_spe.rds")
```

We'll also define the path for outputting clusters.
To save space, we'll conclude by export a TSV with the clusters we calculate in this notebook.

```{r}
cluster_tsv <- file.path(data_dir, "osteo_clusters.tsv")
```


## Import

Let's start by reading in the pre-processed SPE object:

```{r import-rds, live = TRUE}
spe <- readr::read_rds(normalized_rds)
```

Before working with this SPE, let's get familiar with it. 
We'll first print the SPE to review its contents, including the number of spots and genes and what assays are available.

```{r print-spe, live = TRUE}
# print the spe
spe
```

We'll also create a plot of the H&E image to orient ourselves to the tissue before diving into clustering, adjusting the styling of the plot to our preferences.
We'll refer back to it throughout the notebook to help interpret the spatial patterns we find.

```{r plot-visium}
#| fig.width: 3

# define H&E plot for general use in notebook
he_plot <- plotVisium(spe, spots = FALSE, zoom = TRUE) +
  # add styling to match other panels
  theme(
    plot.title = ggplot2::element_text(
      hjust = 0.5,
      margin = ggplot2::margin(0, 0, 0.1, 0)
    ),
    panel.border = ggplot2::element_rect(color = "black", linewidth = 0.3)
  )

he_plot
```

This image clearly shows regions of tumor cells amid the lung parenchyma, which itself is more sparse because of the lung's alveolar spaces.

Let's also get a sense of the RNA counts here by plotting some of the QC statistics that were calculated for you, including the total UMIs (`sum`) and number of unique detected genes (`detected`).
Note that these plots use set an upper limit on the color scale to be able to see lower value spots more clearly.

```{r show qc}
#| warning: false
#| message: false
#| fig.height: 3

total_plot <- plotCoords(spe, annotate = "sum", point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 20,000; all value above this will be shown as 20,000
    limits = c(0, 20000), 
    # tell the scales to "squish" to the limit specified above, where
    # oob stands for "out of bounds" handling
    oob = scales::oob_squish, 
    # fun with color scales!
    option = "plasma"
  )
detected_plot <- plotCoords(spe, annotate = "detected",  point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 10,000
    limits = c(0, 10000), 
    oob = scales::oob_squish, 
    option = "plasma"
  )
total_plot + detected_plot
```


There is clearly more RNA expression in the tumor regions than in the parenchyma, which makes sense biologically: tumor cells are highly proliferative and transcriptionally active, while the lower density of parenchymal tissue makes for fewer cells per spot, and less overall expression.

This plot also illustrates why local QC filtering thresholds are preferable to global ones for heterogeneous tissue samples like this.
Specifically, the parenchymal spots have lower UMI and gene counts not because they are low quality, but because of the alveolar architecture of lung tissue itself.
A global threshold would not distinguish between a genuinely low-quality spot and a spot that overlies sparse tissue, and would have ended up preferentially removing parenchymal spots.
For a Visium experiment, we want to retain much of the spatial context, including the contrast between tumor and parenchymal regions, and using local QC thresholds allows us to do so. 


## Dimension reduction and clustering

The first clustering analysis we'll perform will look analogous to single-cell post-processing. 
After that, we'll compare to clustering with the `BANKSY` approach to see how including spatial information influences cluster detection.

### Expression-only analysis

In this section, we'll apply the same dimensionality reduction and clustering workflow we would use for single-cell data, treating each spot's expression profile as the unit of analysis without any spatial information.
This serves as a useful baseline for comparison when we later apply the spatially-aware `BANKSY` approach.

Again, remember that we are still working at the spot level.
This means clusters will represent groups of cells, rather than similar cells specifically.

When calculating PCA, we often prefer to reduce the number of genes to only the most highly variable ones, which will both speed up calculations and ensure stronger signal. 
To do this, we'll start by modeling per-gene expression variance with `scran::modelGeneVar()` which separated each gene's variation into biological and technical components, under the assumption that genes with lower average expression tend to have lower variance due to technical reasons.
We'll then use these modeling results to select the top 2000 genes with the highest biological variation (these are our high-variance genes, aka HVGs), and finally we'll perform PCA that considers just these genes with `scater::runPCA()`.

```{r nonspatial pca, live = TRUE}
# model gene variance
gene_variance <- scran::modelGeneVar(spe)

# calculate the top 2000 HVGs
hvg_vector <- scran::getTopHVGs(spe, n = 2000)

# calculate PCE using the top HVGs, saving back to the SPE object
spe <- scater::runPCA(spe, subset_row = hvg_vector)
```

We can now perform graph-based clustering using `scran::clusterCells()`, which calculates clusters using the `bluster` package under the hood.

We'll use Jaccard weighting and the Louvain algorithm, specifying `k = 20` nearest neighbors.
We'll also save these clustering parameters as `cluster_params` so we can reuse the same settings when running spatially-aware `BANKSY` clustering later.

But first, one caveat: it's worth noting that the choice of `k` and clustering algorithm are important analytical decisions (as well as other parameters like resolution, which we are leaving at its default value of 1 here), but we are not going to evaluate those parameters here for time.
In your own research, we do recommend taking some time to explore parameter space when clustering, which you can learn more about in the [the `scRNA-seq-advanced` training module's clustering evaluation exercise](../scRNA-seq-advanced/exercise_05-cluster_evaluation.Rmd).

```{r nonspatial cluster, live = TRUE}
# define clustering parameters
cluster_params <- bluster::NNGraphParam(
 k = 20,
 cluster.fun = "louvain",
 type = "jaccard"
)

# perform clustering and save to SPE colData
colData(spe)$cluster_nonspatial <- scran::clusterCells(
  spe,
  use.dimred = "PCA",
  BLUSPARAM = cluster_params
)
```

We'll also calculate a UMAP for visualization using `scater::runUMAP()`.

```{r nonspatial umap}
spe <- scater::runUMAP(spe)
```

Let's visualize our results so far by plotting the UMAP and spatial coordinate plots, each colored by cluster with the same color palette.
We'll also include the H&E image for reference.

```{r plot nonspatial}
#| message: false
#| fig.width: 10
#| fig.height: 3

# define UMAP plot
umap_nonspatial <- scater::plotUMAP(
  spe,
  color_by = "cluster_nonspatial",
  point_size = 0.5
) +
  # set custom palette and legend title
  scale_color_brewer(
    name = "Non-spatial cluster", 
    palette = "Dark2"
  ) +
  # ensure more visible points in legend
  guides(
    color = guide_legend(override.aes = list(alpha = 1, size = 1.5))
  )

# define coordinates plot
coords_nonspatial <- plotCoords(
  spe,
  annotate = "cluster_nonspatial",
  point_size = 1.25
) +
  # set custom palette and legend title
  scale_color_brewer(
    name = "Non-spatial cluster", 
    palette = "Dark2"
  )

# combine plots with patchwork
umap_nonspatial + coords_nonspatial + he_plot
```

This has resulted in five clusters.
How do you think these compare to regions in the H&E?
What patterns do you see, for example between tumor and parenchymal regions?


### Spatially-aware clustering with BANKSY


[`BANKSY`](https://prabhakarlab.github.io/Banksy/) ([Singhal et al. 2024](https://doi.org/10.1038/s41588-024-01664-3)) is a method for spatially-aware dimensionality reduction and clustering.
These spatially-aware clusters can also serve as a first step towards tissue domain identification.

This method rests on the assumption that spots in the same tissue domain tend not only to share similar expression profiles, but tend also to be surrounded by spots with similar expression.
In other words, nearby spots are likely to be transcriptionally similar. 
`BANKSY` leverages this insight by incorporating neighborhood information into the PCA itself.
Performing clustering on this spatially-aware PCA allows `BANKSY` to identify clusters with both distinct expression and tissue domain structure.

The `BANKSY` algorithm works in three steps:

1. **Compute the BANKSY matrix** with `Banksy::computeBanksy()`.
For each spot, `BANKSY` averages expression across its `k_geom` nearest spatial neighbors to create a neighborhood expression profile, stored as the `H0` assay. 
Optionally, `BANKSY` can also compute an Azimuthal Gabor Filter (AGF), which captures directional expression gradients that can help resolve finer structure at domain boundaries. 
When AGF is turned on, a second `BANKSY` matrix is stored as the `H1` assay (the `0` and `1` in these assay names indicate if AGF is off or on, respectively). 


2. **Run BANKSY PCA** with `Banksy::runBanksyPCA()`.
`BANKSY` first creates single matrix that is a weighted combination of the normalized expression in `logcounts` and the `BANKSY` neighborhood matrix (`H0` or `H0` and `H1` if AGF is on). 
(Note that `BANKSY` will consider all genes present in the SPE when calculating the PCA; it does not subset to only the most variable genes.
We could perform this ourselves by subsetting the SPE before running `BANKSY`, but for this notebook we will proceed with the full object.)
To combine these matrices, `BANKSY` introduces the `lambda` parameter.
`lambda` ranges from 0 -- 1 and controls how much weight is given to the `BANKSY` matrix neighborhood information relative to `logcounts` expression, where a value of 0 considers only `logcounts` expression. 
The `BANKSY` authors recommend a value of `lambda = 0.2` for Visium spot data, as described in [this `BANKSY` vignette on parameter selection](https://prabhakarlab.github.io/Banksy/articles/parameter-selection.html#parameters).

`BANKSY` then performs PCA using this combined matrix as the input.

3. **Perform graph-based clustering**.
Once the PCA is calculated, it can be used for a variety of downstream tasks, including clustering and UMAP calculation.
In this notebook, we'll use `scran` and `scater` tools for these steps, but we will also point out that `BANKSY` includes wrapper functions for these calculations, which can be useful if you are evaluating multiple `BANKSY` parameters at once. 


In this notebook, we'll run `BANKSY` in two configurations — without AGF and with AGF — and compare the results.

But, before we get started, we have to modify our SPE object.
As `BANKSY` does not perform feature selection on its own, we'll perform that ourselves to ensure a fair comparison between clustering approaches.
We'll do this by subsetting the SPE to just the high-variance genes before running `BANKSY` and save it to a new variable `hvg_spe`.
We'll use this SPE object for the rest of the notebook.

```{r subset spe, live = TRUE}
# subset the SPE to contain only HVGs
hvg_spe <- hvg_spe[hvg_vector,]
```



#### Without AGF

We'll start by computing the BANKSY matrix without AGF.
As noted above, `lambda = 0.2` is a commonly used starting point for Visium spot data; this value means roughly 80% of the signal in the combined matrix comes from each spot's own expression and 20% comes from its neighborhood.
We'll use `k_geom = 18` spatial neighbors for the neighborhood averaging, which is also the recommended starting point for Visium spot data. 
 

```{r banksy matrix agf off, live = TRUE}
# define parameters recommended for Visium
lambda <- 0.2
k_geom <- 18

# calculate the BANKSY matrix without AGF
hvg_spe <- Banksy::computeBanksy(hvg_spe, assay_name = "logcounts", k_geom = k_geom)

# show the SPE
hvg_spe
```

Notice we now have only 2000 genes, which is because we subset the full object for the `BANKSY` calculations.

We also see that a new `H0` assay has been added to the SPE, representing the neighborhood-averaged spot expression. 
Each entry represents the average expression of a given gene across the spatial neighbors of a given spot, rather than the spot's own expression.

```{r show H0}
assay(hvg_spe, "H0")[1:5, 1:5]
```

`computeBanksy()` also stores the parameters it used in the SPE's `metadata` slot, which is useful for keeping track of settings across runs:

```{r banksy nonagf metadata}
# show the SPE metadata
metadata(hvg_spe)$BANKSY_params
```

Now, we can run `Banksy::runBanksyPCA()`, which combines `logcounts` and `H0` into a single matrix weighted by `lambda` and then runs PCA on it.

```{r banksy pca, live = TRUE}
# calculate PCA within BANKSY
hvg_spe <- Banksy::runBanksyPCA(hvg_spe, lambda = lambda)

# show the SPE
hvg_spe
```

The new PCA that `BANKSY` has calculated is named `PCA_M0_lam0.2`: `M0` indicates that AGF was not used, and `lam0.2` records the `lambda` value.
`BANKSY` generates these names automatically, making it easy to track different reduced dimension calculations when comparing multiple runs.

Here we'll use `scran` and `scater` directly to keep the workflow consistent with what we did above.
We'll also reuse the same `cluster_params` object defined earlier when we performed non-spatial clustering to ensure a direct comparison.

```{r cluster banksy nonagf}
# define PCA and UMAP names for bookkeeping
pca_name <- "PCA_M0_lam0.2"
umap_name <- "UMAP_banksy_nonagf"

# calculate UMAP
hvg_spe <- scater::runUMAP(
  hvg_spe,
  dimred = pca_name,
  name = umap_name
)

# perform clustering
colData(hvg_spe)$cluster_banksy_nonagf <- scran::clusterCells(
  hvg_spe,
  use.dimred = pca_name,
  BLUSPARAM = cluster_params
)
```

Let's visualize this initial set of `BANKSY` clusters alongside the H&E, just as we did for the expression-only results.

```{r plot banksy nonagf}
#| message: false
#| fig.width: 10
#| fig.height: 3

# define UMAP plot
umap_banksky_nonagf <- scater::plotReducedDim(
  hvg_spe,
  dimred = umap_name,
  color_by = "cluster_banksy_nonagf",
  point_size = 0.5
) +
  # set custom palette and legend title
  scale_color_brewer(
    name = "BANKSY cluster", 
    palette = "Set1"
  ) +
  # ensure more visible points in legend
  guides(
    color = guide_legend(override.aes = list(alpha = 1, size = 1.5))
  )


# define coordinates plot
coords_banksky_nonagf <- plotCoords(
  hvg_spe,
  annotate = "cluster_banksy_nonagf",
  point_size = 1.25
) +
  # set custom palette and legend title
  scale_color_brewer(
    name = "BANKSY cluster", 
    palette = "Set1"
  )

# combine plots using patchwork
umap_banksky_nonagf + coords_banksky_nonagf + he_plot
```

How do these clusters compare to the expression-only results?
What differences do you see between these results, both in terms of the UMAP and the coordinates plot?



#### With AGF

Let's do this again, but with AGF turned on. 
We'll change up the `k_geom` parameter for this to use two values, more explanation coming later.


```{r banksy matrix agf on, live = TRUE}
# 18 will be used for neighbor averaging, and 30 will be used for broader spatial effects with AGF
k_geom <- c(18, 30)

# turn on compute_agf
hvg_spe <- Banksy::computeBanksy(
  hvg_spe, 
  assay_name = "logcounts", 
  compute_agf = TRUE, 
  k_geom = k_geom
)

# show the SPE
hvg_spe
```

There's a new assay now, H1, which is the result of the AGF portion of this calculation.
BANKSY can only make assays named H0 and H1, so if you re-run in same mode but with a different `k_geom` your assays will get overridden.


```{r}
assay(hvg_spe, "H1")[1:5, 1:5]
```

Worth noting that BANKSY will overwrite metadata parameters.

```{r banksy agf metadata}
metadata(hvg_spe)$BANKSY_params
```


Now, we can calculate the PCA for this matrix.

```{r agf pca, live = TRUE}
# use_agf tells BANKSY to use the H1 asssay
hvg_spe <- Banksy::runBanksyPCA(hvg_spe, lambda = lambda, use_agf = TRUE)

# show the SPE
hvg_spe
```

This gave us a new PCA matrix, `PCA_M1_lam0.2` - where the `M1` signifies that AGF was used and `0.2` is the lambda parameter.
As before we'll do clustering and UMAP.

```{r agf umap}
# define reducedDim names
pca_name <- "PCA_M1_lam0.2"
umap_name <- "UMAP_banksy_agf"

# Calculate UMAP
hvg_spe <- scater::runUMAP(
  hvg_spe,
  dimred = pca_name,
  name = umap_name
)

# Perform clustering
colData(hvg_spe)$cluster_banksy_agf <- scran::clusterCells(
  hvg_spe, 
  use.dimred = pca_name,
  BLUSPARAM = cluster_params
)
```

And plot, this time showing a few plots for comparison!
We'll make use of patchwork for lots of colors.

```{r plot banksy agf}
#| message: false
#| fig.height: 6
#| fig.width: 14

umap_banksky_agf <- scater::plotReducedDim(
  hvg_spe, 
  dimred = umap_name,
  color_by = "cluster_banksy_agf", 
  point_size = 0.5
) + 
  # set custom palette and legend title
  scale_color_brewer(
    name = "BANKSY AGF cluster",
    palette = "Paired"
  ) +
  # ensure more visible points in legend
  guides(
    color = guide_legend(override.aes = list(alpha = 1, size = 1.5))
  )



coords_banksky_agf <- plotCoords(
  hvg_spe, 
  annotate = "cluster_banksy_agf", 
  point_size = 1.25
) + 
  # set custom palette and legend title
  scale_color_brewer(
    name = "BANKSY AGF cluster", 
    palette = "Paired"
  )


# Compile plots for display with patchwork
plot_list <- list(
  umap_nonspatial,
  umap_banksky_nonagf,
  umap_banksky_agf,
  coords_nonspatial,
  coords_banksky_nonagf, 
  coords_banksky_agf
)

wrap_plots(plot_list, nrow = 2)
```

What differences do you see with and without AGF?
Do you think AGF captured boundaries more cleanly as advertised?
How do these clusters overall compare to non-spatial?

TODO: Burn down the heatmap because it's just harder to look at now with more numbers?
We can see how clustering approaches compare a bit more precisely/plainly with a heatmap, specifically by comparing non-spatial to `BANKSY` with AGF.

```{r cluster banksy heatmap}
clusters_tab <- table(
  colData(hvg_spe)$cluster_nonspatial, 
  colData(hvg_spe)$cluster_banksy_agf
)

# normalize by the number of cells in each AGF cluster
clusters_tab <- apply(
  clusters_tab,
  2, # apply function to columns
  \(x) x/sum(x)
)

# plot heatmap:
# non-spatial clusters are rows, and BANKSY with AGF clusters are columns
pheatmap::pheatmap(
  clusters_tab,
  # turn off heatmap hierarchical cluster for easier viewing
  cluster_cols = FALSE,
  cluster_rows = FALSE
)
```

For example, we can see from the coordinate plots that non-spatial clusters 3 and 5 correspond to spots likely dominated by tumor cells, which has been split into several BANKSY AGF clusters including at least 3, 5, and 7.

Let's color by gene expression for a gene we expect to be highly expressed in the tumor, collagen (`Col1a1`).
Let's get the Ensembl id to plot it:

```{r find ensembl id}
ensembl_id <- rowData(hvg_spe) |>
  as.data.frame() |>
  dplyr::filter(Symbol == "Col1a1")|>
  rownames()

ensembl_id
```

```{r plot col1a1 expression}
#| message: false
#| fig.width: 12
#| fig.height: 4

nonspatial_expr_plot <- scater::plotExpression(
  hvg_spe,
  features = ensembl_id,
  x = "cluster_nonspatial",
  color_by = "cluster_nonspatial"
) + 
  scale_color_brewer(name = "non-spatial cluster", palette = "Dark2")
  
agf_expr_plot <- scater::plotExpression(
  hvg_spe,
  features = ensembl_id,
  x = "cluster_banksy_agf",
  color_by = "cluster_banksy_agf"
) + 
  scale_color_brewer(name = "BANKSY AGF cluster", palette = "Paired")

nonspatial_expr_plot + agf_expr_plot
```

Do these patterns match your expectations from the coordinate plots showing clusters on the H&E?


## Export

We'll conclude by exporting the `colData` slot of the `hvg_spe` as a TSV, which will include the clusters we have calculated.

```{r export, live = TRUE}
# convert colData to data frame and export as a TSV
colData(hvg_spe) |>
  as.data.frame() |>
  readr::write_tsv(cluster_tsv)
```


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

