---
title: "Clustering spatial transcriptomics data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Perform dimensionality reduction (PCA and UMAP) and graph-based clustering on Visium data
- Use BANKSY to perform spatially-aware dimensionality reduction and clustering
- Compare clusters calculated with and without spatial information

## Introduction

In this notebook, we'll cluster expression data from a mouse sample with osteosarcoma lung metastases from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049).
This sample was processed using second-generation Visium technology and quantified with `Space Ranger 2.1.1`.
We have already processed this dataset for you, including filtering out low quality spots (note that `SpotSweeper` removed 20 low-quality spots during this step) and normalization, which you can see in this notebook: <https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/osteo/prepare-osteo-data.Rmd>.

As we have discussed, Visium spatial transcriptomics data does not quantify expression at the single-cell level, but rather each Visium spot captures the combined expression of 5-20 cells.
This means that when we perform clustering on Visium data, we are actually clustering spots with similar transcriptional profiles, not individual cells.

Also unlike single-cell data, we now have an additional consideration of spatial information.
We can choose to incorporate each spot's physical position in the tissue alongside its expression profile during clustering, which may help to approximate spatial domains. 

In this notebook, we'll explore complementary approaches: first running expression-based dimensionality reduction and clustering just as we would for single-cell data, and then performing the same analyses using the [`BANKSY` package](https://bioconductor.org/packages/release/bioc/html/Banksy.html). 
`BANKSY` offers a spatially-aware dimensionality reduction and clustering approach that augments each spot's expression with information from its spatial neighborhood.

## Set up

To begin, we'll load the packages we'll use throughout this notebook.

```{r libraries}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
})


# To support plotting
library(ggspavis)
library(ggplot2)
library(patchwork)

# Set ggplot2 theme
theme_set(theme_bw())
```


We'll set a seed for reproducibility, since we'll be running UMAP and clustering.

```{r}
set.seed(2026)
```


### Directories and files

The pre-processed SPE object for this sample is stored in a sample-specific subdirectory of `data/osteo`, organized by GEO sample ID.

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "GSM8478586"

# data directory for this sample
data_dir <- file.path("data/osteo", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "osteo_normalized_spe.rds")
```

We'll also define the path for the output SPE with clustering results, which we'll write to at the end of the notebook:

```{r}
clustered_rds <- file.path(data_dir, "osteo_clustered_spe.rds")
```


## Import

Let's start by reading in the pre-processed SPE object:

```{r import-rds, live = TRUE}
spe <- readr::read_rds(normalized_rds)
```

Before working with this SPE, let's get familiar with it. 
We'll first print the SPE to review its contents, including the number of spots and genes and what assays are available.

```{r print-spe, live = TRUE}
# print the spe
spe
```

We'll also visualize the H&E image to orient ourselves to the tissue before diving into clustering, using some custom styling.
We'll refer back to it throughout the notebook to help interpret the spatial patterns we find.

```{r plot-visium}
#| fig.width: 5

# define H&E plot for general use in notebook
he_plot <- plotVisium(spe, spots = FALSE, zoom = TRUE) +
  ggtitle("H&E") +
  # add styling to match other panels
  theme(
    plot.title = ggplot2::element_text(
      hjust = 0.5,
      margin = ggplot2::margin(0, 0, 0.1, 0)
    ),
    panel.border = ggplot2::element_rect(color = "black", linewidth = 0.25)
  )

he_plot
```

This image clearly shows regions of tumor cells amid the lung parenchyma, which itself is more sparse because of the lung's alveolar spaces.

Let's also get a sense of the RNA counts here by plotting some of the QC statistics that were calculated for you, including the total UMIs (`sum`) and number of unique detected genes (`detected`).
Note that these plots use right-censoring and set an upper limit on the color scale to be able to see lower value spots more clearly.

```{r show qc}
#| warning: false
#| message: false

total_plot <- plotCoords(spe, annotate = "sum", point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 20,000; all value above this will be shown as 20,000
    limits = c(0, 20000), 
    oob = scales::oob_squish, 
    # fun with color scales!
    option = "plasma"
  )
detected_plot <- plotCoords(spe, annotate = "detected",  point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 10,000
    limits = c(0, 10000), 
    oob = scales::oob_squish, 
    option = "plasma"
  )
total_plot + detected_plot
```


There is clearly more RNA expression in the tumor regions than in the parenchyma, which makes sense biologically: tumor cells are highly proliferative and transcriptionally active, but parenchymal is comparatively sparse.

This plot also illustrates why local QC filtering thresholds are preferable to global ones for heterogeneous tissue samples like this.
Specifically, the parenchymal spots have lower UMI and gene counts not because they are low quality, but because of the alveolar architecture of lung tissue itself.
A global threshold would not distinguish between a genuinely low-quality spot and a spot that overlies sparse tissue, and would have ended up preferentially removing parenchymal spots.
For a Visium experiment, we want to retain much of the spatial context, including the contrast between tumor and parenchymal regions, and using local QC thresholds allows us to do so. 


## Calculate clusters

The first clustering analysis we'll perform will look analogous to single-cell post-processing. 
After that, we'll compare to clustering with the `BANKSY` approach to see how including spatial information influences cluster detection.

### Expression-based clustering

In this section, we'll apply the same dimensionality reduction and clustering workflow we would use for single-cell data, treating each spot's expression profile as the unit of analysis without any spatial information.
This serves as a useful baseline for comparison when we later apply the spatially-aware `BANKSY` approach.

Again, remember that even though we are using single-cell methods, we are still working at the spot level.
This means clusters will represent regions of similar transcriptional composition, rather that similar cells specifically.

We'll start by modeling per-gene expression variance with `scran::modelGeneVar()` and then selecting the top 2000 highly variable genes (HVGs) with `scran::getTopHVGs()` to use for PCA.
We should point out that `scran::modelGeneVar()` treats each spot independently and is only considering gene expression (not location) when identifying variable genes.
There is a related concept called _spatially variable genes_ (SVGs), which are genes whose expression varies across the tissue, but we won't be using those in this notebook.

We'll then calculate the PCA with `scater::runPCA()` .

```{r nonspatial pca, live = TRUE}
# model gene variance
gene_variance <- scran::modelGeneVar(spe)

# calculate the top 2000 HVGs
hvg_vector <- scran::getTopHVGs(spe, n = 2000)

# calculate PCE using the top HVGs, saving back to the SPE object
spe <- scater::runPCA(spe, subset_row = hvg_vector)
```

We can now perform graph-based clustering using `scran::clusterCells()`, which calculates clusters using the `bluster` package under the hood.

We'll use `k = 20` nearest neighbors and the Louvain algorithm.
We'll also save these clustering parameters as `cluster_params` so we can reuse the same settings when running spatially-aware `BANKSY` clustering later.

But first, one caveat: it's worth noting that the choice of `k` and clustering algorithm are important analytical decisions (as well as other parameters like resolution, which we are leaving at its default value of 1 here), but we are not going to evaluate those parameters here for time.
In your own research, we do recommend taking some time to explore parameter space when clustering, which you can learn more about in the [the `scRNA-seq-advanced` training module's clustering evaluation exercise](../scRNA-seq-advanced/exercise_05-cluster_evaluation.Rmd).

```{r nonspatial cluster, live = TRUE}
# define clustering parameters
cluster_params <- bluster::NNGraphParam(
  k = 20,
  cluster.fun = "louvain"
)

# perform clustering and save to SPE colData
colData(spe)$cluster_nonspatial <- scran::clusterCells(
  spe,
  use.dimred = "PCA",
  BLUSPARAM = cluster_params
)
```

We'll also calculate a UMAP for visualization using `scater::runUMAP()`.

```{r nonspatial umap}
spe <- scater::runUMAP(spe)
```

Let's visualize our results so far by plotting the UMAP colored by cluster alongside the spatial coordinates plot and H&E image.
We'll also specify a consistent color palette across both panels makes it easier to match each cluster to its location in the tissue.

```{r plot nonspatial}
#| message: false
#| fig.width: 10

# define UMAP plot
umap_nonspatial <- scater::plotUMAP(
  spe,
  color_by = "cluster_nonspatial",
  point_size = 0.5
) +
scale_color_brewer(name = "Non-spatial cluster", palette = "Dark2")

# define coordinates plot
coords_nonspatial <- plotCoords(
  spe,
  annotate = "cluster_nonspatial",
  point_size = 1.25
) +
  scale_color_brewer(name = "Non-spatial cluster", palette = "Dark2")

# combine plots with patchwork
umap_nonspatial + coords_nonspatial + he_plot
```

We see four clusters with some correspondence to the tissue structure visible in the H&E: the clusters appear to capture a gradient from the most tumor-dense spots to the sparser parenchymal regions.
Interestingly, spots that are spatially adjacent also tend to sit near each other in the UMAP, even though no spatial information was provided (and we want to be careful not to "over-interpret" the UMAP too!).

Do you see any other patterns here to point out?


### Spatially-aware clustering with BANKSY


[`BANKSY`](https://prabhakarlab.github.io/Banksy/) ([Singhal et al. 2024](https://doi.org/10.1038/s41588-024-01664-3)) is a method for spatially-aware dimensionality reduction and clustering.
These spatially-aware clusters can also serve as a first step towards tissue domain identification.

This method rests on the assumption that spots in the same tissue domain tend not only to share similar expression profiles, but tend also to be surrounded by spots with similar expression.
In other words, nearby spots are likely to be transcriptionally similar. 
`BANKSY` leverages this insight by incorporating neighborhood information into the PCA itself.
Performing clustering on this spatially-aware PCE allows `BANKSY` to identify clusters with both distinct expression and tissue domain structure.

The `BANKSY` algorithm works in three steps:

1. **Compute the BANKSY matrix** with `Banksy::computeBanksy()`.

For each spot, `BANKSY` averages expression across its `k_geom` nearest spatial neighbors to create a neighborhood expression profile, stored as the `H0` assay. 
Optionally, `BANKSY` can also compute an Azimuthal Gabor Filter (AGF), which captures directional expression gradients that can help resolve finer structure at domain boundaries. 
When AGF is turned on, a second `BANKSY` matrix is stored as the `H1` assay (the `0` and `1` in these assay names indicate if AGF is off or on, respectively). 


2. **Run BANKSY PCA** with `Banksy::runBanksyPCA()`.

`BANKSY` combines the spot-level normalized expression in `logcounts` and the `BANKSY` neighborhood matrix (`H0` or `H0` and `H1` if AGF is on) into a single weighted matrix that contains both expression and spatial information to calculate PCA on. 
(Note that `BANKSY` will consider all genes present in the SPE when calculating the PCA; it does not subset to only the most variable genes.) , although we could perform ths subsetting manualy
To combine these matrices, `BANKSY` introduces the `lambda` parameter.
`lambda` ranges from 0 -- 1 and controls how much weight is given to the `BANKSY` matrix neighborhood information relative to `logcounts` expression, where a value of 0 considers only `logcounts` expression. 
The `BANKSY` authors recommend a value of `lambda = 0.2` for Visium spot data, as described in [this `BANKSY` vignettes on parameter selection](https://prabhakarlab.github.io/Banksy/articles/parameter-selection.html#parameters).

3. **Cluster and visualize**.

Once the PCA is calculated, it can be used for a variety of downstream tasks including clustering and UMAP calculation.
In this notebook, we'll use `scran` and `scater` tools for these steps, but we will also point that `BANKSY` includes wrapper functions for these calculations, which can be useful if you are evaluating multiple `BANKSY` parameters at once. 


In this notebook, we'll run `BANKSY` in two configurations — without AGF and with AGF — and compare the results.

#### Without AGF

We'll start by computing the BANKSY matrix without AGF.
As noted above, `lambda = 0.2` is a commonly used starting point for Visium spot data; this value means roughly 80% of the signal in the combined matrix comes from each spot's own expression and 20% comes from its neighborhood.
We'll use `k_geom = 18` spatial neighbors for the neighborhood averaging, which is also the recommended starting point for Visium spot data. 

```{r banksy matrix agf off, live = TRUE}
# define parameters recommended for Visium
lambda <- 0.2
k_geom <- 18

# calculate the BANKSY matrix without AGF
spe <- Banksy::computeBanksy(spe, assay_name = "logcounts", k_geom = k_geom)

# show the spe
spe
```

A new `H0` assay has been added to the SPE, representing the neighborhood-averaged spot expression. 
Each entry represents the average expression of a given gene across the spatial neighbors of a given spot, rather than the spot's own expression.

```{r show H0}
assay(spe, "H0")[1:5, 1:5]
```

`computeBanksy()` also stores the parameters it used in the SPE's `metadata` slot, which is useful for keeping track of settings across runs:

```{r banksy nonagf metadata}
# show spe metadata
metadata(spe)$BANKSY_params
```

Now, we can run `Banksy::runBanksyPCA()`, which combines `logcounts` and `H0` into a single matrix weighted by `lambda` and then runs PCA on it.

```{r banksy pca, live = TRUE}
# calculate PCA within BANKSY
spe <- Banksy::runBanksyPCA(spe, lambda = lambda)

# show SPE
spe
```

The new PCA that `BANKSY` has calculated is named `PCA_M0_lam0.2`: `M0` indicates that AGF was not used, and `lam0.2` records the `lambda` value.
`BANKSY` generates these names automatically, making it easy to track different reduced dimension calculations when comparing multiple runs.

Here we'll use `scran` and `scater` directly to keep the workflow consistent with what we did above.
We'll also reuse the same `cluster_params` object defined earlier when we performed non-spatial clustering to ensure a direct comparison.

```{r cluster banksy nonagf}

# define PCA and UMAP names for bookkeeping
pca_name <- "PCA_M0_lam0.2"
umap_name <- "UMAP_banksy_nonagf"

# calculate UMAP
spe <- scater::runUMAP(
  spe,
  dimred = pca_name,
  name = umap_name
)

# perform clustering
colData(spe)$cluster_banksy_nonagf <- scran::clusterCells(
  spe,
  use.dimred = pca_name,
  BLUSPARAM = cluster_params
)
```

Let's visualize this initial set of `BANKSY` clusters alongside the H&E, just as we did for the expression-only results.

```{r plot banksy nonagf}
#| message: false
#| fig.width: 10

# define UMAP plot
umap_banksky_nonagf <- scater::plotReducedDim(
  spe,
  dimred = umap_name,
  color_by = "cluster_banksy_nonagf",
  point_size = 0.5
) +
scale_color_brewer(name = "BANKSY clusters", palette = "Set1")

# define coordinates plot
coords_banksky_nonagf <- plotCoords(
  spe,
  annotate = "cluster_banksy_nonagf",
  point_size = 1.25
) +
scale_color_brewer(name = "BANKSY clusters", palette = "Set1")

# combine plots using patchwork
umap_banksky_nonagf + coords_banksky_nonagf + he_plot
```

How do these clusters compare to the expression-only results?
What differences do you see between these results, both in terms of the UMAP and the coordinates plot?



#### With AGF

Let's do this again, but with AGF turned on. 
We'll change up the `k_geom` parameter for this to use two values, more explanation coming later.


```{r banksy matrix agf on, live = TRUE}
# 18 will be used for neighbor averaging, and 30 will be used for broader spatial effects with AGF
k_geom <- c(18, 30)

# turn on compute_agf
spe <- Banksy::computeBanksy(
  spe, 
  assay_name = "logcounts", 
  compute_agf = TRUE, 
  k_geom = k_geom
)

# show spe
spe
```

There's a new assay now, H1, which is the result of the AGF portion of this calculation.
BANKSY can only make assays named H0 and H1, so if you re-run in same mode but with a different `k_geom` your assays will get overridden.


```{r}
assay(spe, "H1")[1:5, 1:5]
```

Worth noting that BANKSY will overwrite metadata parameters.

```{r banksy agf metadata}
metadata(spe)$BANKSY_params
```


Now, we can calculate the PCA for this matrix.

```{r agf pca, live = TRUE}
# use_agf tells BANKSY to use the H1 asssay
spe <- Banksy::runBanksyPCA(spe, lambda = lambda, use_agf = TRUE)

# show the SPE
spe
```

This gave us a new PCA matrix, `PCA_M1_lam0.2` - where the `M1` signifies that AGF was used and `0.2` is the lambda parameter.
As before we'll do clustering and UMAP.

```{r agf umap}
# define reducedDim names
pca_name <- "PCA_M1_lam0.2"
umap_name <- "UMAP_banksy_agf"

# Calculate UMAP
spe <- scater::runUMAP(
  spe,
  dimred = pca_name,
  name = umap_name
)

# Perform clustering
colData(spe)$cluster_banksy_agf <- scran::clusterCells(
  spe, 
  use.dimred = pca_name,
  BLUSPARAM = cluster_params
)
```

And plot, this time showing a few plots for comparison!
We'll make use of patchwork for lots of colors.

```{r plot banksy agf}
#| message: false
#| fig.height: 6
#| fig.width: 14

umap_banksky_agf <- scater::plotReducedDim(
  spe, 
  dimred = umap_name,
  color_by = "cluster_banksy_agf", 
  point_size = 0.5
) + 
scale_color_brewer(palette = "Paired")

coords_banksky_agf <- plotCoords(
  spe, 
  annotate = "cluster_banksy_agf", 
  point_size = 1.25
) + 
scale_color_brewer(palette = "Paired")


# Compile plots for patchwork
# the legends will be a bit loud, but wrangling it is a pain!
plot_list <- list(
  umap_nonspatial,
  umap_banksky_nonagf,
  umap_banksky_agf,
  coords_nonspatial,
  coords_banksky_nonagf, 
  coords_banksky_agf
)


wrap_plots(plot_list, nrow = 2)
```

What differences do you see with and without AGF?
Did AGF capture boundaries more cleanly as advertised?
How do these clusters overall compare to non-spatial?

We can see how clusters compare a bit more precisely/plainly with a heatmap.
For this, we'll make one comparing non-spatial to AGF clusters.

```{r cluster banksy heatmap}
clusters_tab <- table(
  colData(spe)$cluster_banksy_agf, 
  colData(spe)$cluster_nonspatial
)

# normalize by the number of cells in each non-spatial cluster
clusters_tab <- apply(
  clusters_tab,
  2, # apply function to columns (cluster_nonspatial)
  \(x) x/sum(x)
)

pheatmap::pheatmap(clusters_tab)
```

For example, we can see from the coordinate plots that non-spatial cluster 3 is tumor cells, which has been split into three BANKSY clusters: 10, 6, & 8.
Notably, BANKSY cluster 6 also has spots in non-spatial cluster 1, which we can see from the coordinate plots too.

Let's color by gene expression for a gene we expect to be highly expressed in the tumor, collagen (`Col1a1`).
Let's get the Ensembl id to plot it:

```{r find ensembl id}
ensembl_id <- rowData(spe) |>
  as.data.frame() |>
  dplyr::filter(Symbol == "Col1a1")|>
  rownames()

ensembl_id
```

```{r plot col1a1 expression}
#| message: false
#| fig.width: 12
#| fig.height: 4


banksy_expr_plot <- scater::plotExpression(
  spe,
  features = ensembl_id,
  x = "cluster_banksy_agf",
  color_by = "cluster_banksy_agf"
) + 
  scale_color_brewer(palette = "Paired")
  
nonspatial_expr_plot <- scater::plotExpression(
  spe,
  features = ensembl_id,
  x = "cluster_nonspatial",
  color_by = "cluster_nonspatial"
) + 
  scale_color_brewer(palette = "Dark2")

banksy_expr_plot + nonspatial_expr_plot
```


We indeed see distinctions among BANKSY clusters 6, 8, & 10 - per the H&E, spots in 10 are most isolated from environment, then 8, and then 6 is bordering the parenchyma. 
6 being lowest expression makes sense given that it also corresponds, although less so, to non-spatial cluster 1.
We have therefore refined our non-spatial clusters with spatial information to give us clusters that consider spatial relationships as well.

Anything else you see to point out here?


## Export

We'll conclude by exporting our object with clusters, with compression.

```{r export, live = TRUE}
readr::write_rds(spe, clustered_rds, compress = "gz")
```


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

