---
title: "Spatially-aware clustering"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Perform PCA, clustering, UMAP sans spatial information
- Use BANKSY to perform spatially-aware PCA and clustering
- Compare clusters calculated with and without spatial information

## Introduction

We will be using a sample from a mouse with osteosarcoma lung metastases from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049).
This sample was processed using second-generation Visium technology, and it was quantified with `Space Ranger 2.1.1`.

We have already processed this dataset for you: filtered (note that `SpotSweeper` removed 20 low-quality spots during this step) and normalized it. 

## Set up


```{r libraries}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
})


# To support plotting
library(ggspavis)
library(ggplot2)
library(patchwork)

# Set ggplot2 theme
theme_set(theme_bw())
```


Set a seed for reproducibility:

```{r}
set.seed(2026)
```


### Directories and files

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "GSM8478586"

# data directory for this sample
data_dir <- file.path("data/osteo", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "osteo_normalized_spe.rds")
```

And the output file:

```{r}
clustered_rds <- file.path(data_dir, "osteo_clustered_spe.rds")
```


## Import

```{r import-rds, live = TRUE}
spe <- readr::read_rds(normalized_rds)
```

Look at it first:

```{r print-spe, live = TRUE}
# print the spe
spe
```


```{r plot-visium}
#| fig.width: 5

# define H&E plot for general use in notebook
he_plot <- plotVisium(spe, spots = FALSE, zoom = TRUE) +
  ggtitle("H&E") +
  # add styling to match other panels
  theme(
    plot.title = ggplot2::element_text(
      hjust = 0.5,
      margin = ggplot2::margin(0, 0, 0.1, 0)
    ),
    panel.border = ggplot2::element_rect(color = "black", linewidth = 0.25)
  )

he_plot
```

We can clearly see tumor cells amid the lung parenchyma, which itself is more sparse because of the alveolar spaces which is just how lungs work.
Let's get a sense of the RNA counts here by plotting some of the QC stats we calculated for you.
Note that these plots use right-censoring and set an upper limit on the color scale to be able to see lower value spots more clearly.

```{r show qc}
#| warning: false
#| message: false

total_plot <- plotCoords(spe, annotate = "sum", point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 20,000; all value above this will be shown as 20,000
    limits = c(0, 20000), 
    oob = scales::oob_squish, 
    # fun with color scales!
    option = "plasma"
  )
detected_plot <- plotCoords(spe, annotate = "detected",  point_size = 1) + 
  scale_color_viridis_c(
    # cap scale at 10,000
    limits = c(0, 10000), 
    oob = scales::oob_squish, 
    option = "plasma"
  )
total_plot + detected_plot
```

Takeaways:

- Clearly, more expression in the tumors in general vs parenchyma, which makes sense biologically!
- Because we have really different tissue compartments here, if we had used global thresholds we would have just ended up retaining almost exclusively tumor cells and removing their spatial context due to lower expression, but we would like to retain this spatial context because otherwise we've just wasted a ton of money on Visium


## Calculate clusters

We're going to see how standard single-cell post-processing would look and then compare to one approach for injecting spatial information into the clustering procedure.

### Spatial-free, aka expression only

First, we'll run PCA, clustering, UMAP as we might do for single-cell.
Therefore, in this section, clusters will be driven by expression only but keep in mind that we're working at the spot level with Visium data, not the cell level, and each spot contains 5-20 cells.
This means that we're clustering heterogeneous groups of cells as single unit regions of expression as each spot may not represent a single cell (if you'd like to know many & what cells each spot might have, please keep attending the workshop!). 

First, PCA:

- Model gene variance and pull out the 2000 HVGs for use in PCA
  - Note that this is not spatially aware and there is a distinct problem "spatially variable genes"
- Then we'll use these genes to run the PCA

```{r nonspatial pca, live = TRUE}
gene_variance <- scran::modelGeneVar(spe)
highvar_genes <- scran::getTopHVGs(spe, n = 2000) 

spe <- scater::runPCA(spe, subset_row = highvar_genes)
```

Now, we can cluster.
For this notebook, we'll generally use 20 nearest neighbors with the Louvain algorithm for graph-based clustering.
Worth noting that it's best practice to exploring parameter space though, see [the clustering exercise](../scRNA-seq-advanced/exercise_05-cluster_evaluation.Rmd) to learn more about this.

```{r nonspatial cluster, live = TRUE}
# define clustering params, and we can re-use this later too
cluster_params <- bluster::NNGraphParam(
  k = 20,
  cluster.fun = "louvain"
)

# perform clustering
colData(spe)$cluster_nonspatial <- scran::clusterCells(
  spe, 
  use.dimred = "PCA", 
  BLUSPARAM = cluster_params
)
```


We'll also calculate the UMAP for viz.

```{r nonspatial umap}
spe <- scater::runUMAP(spe)
```

  
Let's have a look, both at the UMAP and at the coordinates colored by cluster.
We'll use the same palette to assist in interpretation, and show the H&E here as well.


```{r plot nonspatial}
#| message: false
#| fig.width: 10

umap_nonspatial <- scater::plotUMAP(
  spe, 
  color_by = "cluster_nonspatial",
  point_size = 0.5
) + 
scale_color_brewer(palette = "Dark2")

coords_nonspatial <- plotCoords(
  spe, 
  annotate = "cluster_nonspatial", 
  point_size = 1.25
) + 
  scale_color_brewer(palette = "Dark2")

umap_nonspatial + coords_nonspatial + he_plot
```

We see four clusters which have some correspondence to patterns in the H&E for tumor vs parenchymal regions, and clusters seem to be based on expression patterns along a "gradient" from tumor -> parenchymal.
Interestingly we see in the UMAP that spatial regions tend to be adjacent; whether this is meaningful isn't clear but it's neat nonetheless.


### Spatially-aware clustering with BANKSY

The field is popping for approaches to incorporate spatial information, and we're going to go through one popular method used for clustering and/or domain identification in spatial data.

Text placeholder for BANKSY methods, briefly...
- calculates a matrix ("BANKSY matrix") that considers both expression and spatial information
  - there are two frameworks that BANKSY uses to consider spatial information: graph-based neighbor averaging, and optionally, AGF on top of that, where the neighborhood sizes are controlled by a `k_geom` parameter
  - Incorporating AGF may help to detect boundaries between domains and can resolve some finer structure
  - At some point we have to note a fun naming convention: H0 and M0 refer to matrices calculated sans AGF, and H1 and M1 with AGF. The H/M distinction is a bit of mathematics pedantry but just remember 0 --> no AGF and 1 --> with AGF.
- Then we can perform PCA on this matrix, using a parameter called `lambda` to tune how much the spatial information contributes
- Finally we can cluster/run UMAP on this PCA

See more details about parameters [from these docs](https://prabhakarlab.github.io/Banksy/articles/parameter-selection.html#parameters).

We're going to run this without and with AGF and then compare all around.

#### Without AGF

```{r banksy matrix agf off, live = TRUE}
# define parameters recommended for Visium
lambda <- 0.2
k_geom <- 18

# calculate the BANKSY matrix without AGF
spe <- Banksy::computeBanksy(spe, assay_name = "logcounts", k_geom = k_geom)

# show the spe
spe
```

Show and explain the H0 matrix (spatially-smoothed neighbor-averaged expression).

```{r show ho}
assay(spe, "H0")[1:5, 1:5]
```

Also note that parameters were added to the `metadata` slot:

```{r banksy nonagf metadata}
# show spe metadata
metadata(spe)$BANKSY_params
```


Now we can calculate PCA, specifying `lambda` for how to weight the original expression (`logcounts`) and the BANKSY matrix (here, H0, 

```{r banksy pca, live = TRUE}
# calculate PCA within BANKSY
spe <- Banksy::runBanksyPCA(spe, lambda = lambda)

# show SPE
spe
```

We have a new PCA called `PCA_M0_lam0.2`.
BANKSY named this for us: `M0` implies that AGF is off, and `lam0.2` is our lambda parameter.

We can proceed to calculate UMAP and clusters now, and we'll use `scran`/`scater` functions for that.
But, it's worth pointing out that BANKSY does allow you to specify multiple `lambda` parameters at once and then offers wrapper functions to cluster/UMAP across those parameters in one function call.

```{r cluster banksy nonagf}
pca_name <- "PCA_M0_lam0.2"
umap_name <- "UMAP_banksy_nonagf"

spe <- scater::runUMAP(
  spe,
  dimred = pca_name,
  name = umap_name
)

# perform clustering
colData(spe)$cluster_banksy_nonagf <- scran::clusterCells(
  spe, 
  use.dimred = pca_name,
  # previously defined as 20 nearest neighbors using the Louvain algorithm and default resolution 1 
  BLUSPARAM = cluster_params 
)
```

And plot:

```{r plot banksy nonagf}
#| message: false
#| fig.width: 10

umap_banksky_nonagf <- scater::plotReducedDim(
  spe, 
  dimred = umap_name,
  color_by = "cluster_banksy_nonagf", 
  point_size = 0.5
) + 
scale_color_brewer(palette = "Set1")


coords_banksky_nonagf <- plotCoords(
  spe, 
  annotate = "cluster_banksy_nonagf", 
  point_size = 1.25
) + 
scale_color_brewer(palette = "Set1")

umap_banksky_nonagf + coords_banksky_nonagf + he_plot
```

How does this compare to non-spatial?
Kind of neat to see that there is still spatial arrangement in the UMAP, although it's not clear what it means if anything!



#### With AGF

Let's do this again, but with AGF turned on. 
We'll change up the `k_geom` param for this to use two values, more explanation coming later.


```{r banksy matrix agf on, live = TRUE}
# 18 will be used for neighbor averaging, and 30 will be used for broader spatial effects with AGF
k_geom <- c(18, 30)

# turn on compute_agf
spe <- Banksy::computeBanksy(
  spe, 
  assay_name = "logcounts", 
  compute_agf = TRUE, 
  k_geom = k_geom
)

# show spe
spe
```

There's a new assay now, H1, which is the result of the AGF calculation.
BANKSY can only make assays named H0 and H1, so if you re-run in same mode but with a different `k_geom` your assays will get overridden.


```{r}
assay(spe, "H1")[1:5, 1:5]
```

Worth noting that Banksy will overwrite metadata parameters.

```{r banksy agf metadata}
metadata(spe)$BANKSY_params
```


Now, we can calculate the PCA for this matrix.

```{r agf pca, live = TRUE}
# use_agf tells BANKSY to use the H1 asssay
spe <- Banksy::runBanksyPCA(spe, lambda = lambda, use_agf = TRUE)

# show the SPE
spe
```

This gave us a new PCA matrix, `PCA_M1_lam0.2` - where the `M1` signifies that AGF was used and `0.2` is the lambda parameter.
As before we'll do clustering and UMAP.

```{r agf umap}
# define reducedDim names
pca_name <- "PCA_M1_lam0.2"
umap_name <- "UMAP_banksy_agf"

# Calculate UMAP
spe <- scater::runUMAP(
  spe,
  dimred = pca_name,
  name = umap_name
)

# Perform clustering
colData(spe)$cluster_banksy_agf <- scran::clusterCells(
  spe, 
  use.dimred = pca_name,
  BLUSPARAM = cluster_params
)
```

And plot, this time showing a few plots for comparison!
We'll make use of patchwork for lots of colors.

```{r plot banksy agf}
#| message: false
#| fig.height: 6
#| fig.width: 14

umap_banksky_agf <- scater::plotReducedDim(
  spe, 
  dimred = umap_name,
  color_by = "cluster_banksy_agf", 
  point_size = 0.5
) + 
scale_color_brewer(palette = "Paired")

coords_banksky_agf <- plotCoords(
  spe, 
  annotate = "cluster_banksy_agf", 
  point_size = 1.25
) + 
scale_color_brewer(palette = "Paired")


# Compile plots for patchwork
# the legends will be a bit loud, but wrangling it is a pain!
plot_list <- list(
  umap_nonspatial,
  umap_banksky_nonagf,
  umap_banksky_agf,
  coords_nonspatial,
  coords_banksky_nonagf, 
  coords_banksky_agf
)


wrap_plots(plot_list, nrow = 2)
```

What differences do you see with and without AGF?
Did AGF capture boundaries more cleanly as advertised?
How do these clusters overall compare to non-spatial?

We can see how clusters compare a bit more precisely/plainly with a heatmap.
For this, we'll make one comparing non-spatial to AGF clusters.

```{r cluster banksy heatmap}
clusters_tab <- table(
  colData(spe)$cluster_banksy_agf, 
  colData(spe)$cluster_nonspatial
)

# normalize by the number of cells in each non-spatial cluster
clusters_tab <- apply(
  clusters_tab,
  2, # apply function to columns (cluster_nonspatial)
  \(x) x/sum(x)
)

pheatmap::pheatmap(clusters_tab)
```

For example, we can see from coord plots that non-spatial cluster 3 is tumor cells, which has been split into three BANKSY clusters: 10, 6, & 8.
Notably, BANKSY cluster 6 also has spots in non-spatial cluster 1, which we can see from the coordinate plots too.

Let's color by gene expression for a gene we expect to be highly expressed in the tumor, collagen (`Col1a1`).
Let's get the Ensembl id to plot it:

```{r find ensembl id}
ensembl_id <- rowData(spe) |>
  as.data.frame() |>
  dplyr::filter(Symbol == "Col1a1")|>
  rownames()

ensembl_id
```

```{r plot col1a1 expression}
#| message: false
#| fig.width: 12
#| fig.height: 4


banksy_expr_plot <- scater::plotExpression(
  spe,
  features = ensembl_id,
  x = "cluster_banksy_agf",
  color_by = "cluster_banksy_agf"
) + 
  scale_color_brewer(palette = "Paired")
  
nonspatial_expr_plot <- scater::plotExpression(
  spe,
  features = ensembl_id,
  x = "cluster_nonspatial",
  color_by = "cluster_nonspatial"
) + 
  scale_color_brewer(palette = "Dark2")

banksy_expr_plot + nonspatial_expr_plot
```


We indeed see distinctions among BANKSY clusters 6, 8, & 10 - per the H&E, spots in 10 are most isolated from environment, then 8, and then 6 is bordering the parenchyma. 
6 being lowest expression makes sense given that it also corresponds, although less so, to non-spatial cluster 1.
We have therefore refined our non-spatial clusters with spatial information to give us clusters that consider spatial relationships as well.

Anything else you see to point out here?


## Export

We'll conclude by exporting our object with clusters, with compression.

```{r export, live = TRUE}
readr::write_rds(spe, clustered_rds, compress = "gz")
```


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

