---
title: "Deconvolution of spatial transcriptomics data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Use RCTD to perform reference-based deconvolution of Visium spots in doublet and full modes
- Visualize the spatial distribution of estimated cell type proportions across the osteosarcoma tissue
- Explore patterns of cell type co-occurrence across spots

## Introduction

In this notebook, we'll continue working with the mouse osteosarcoma lung metastases sample from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049) that we explored in spatial clustering notebook.

As we have discussed, each Visium spot captures the combined expression of approximately 5--20 cells, not a single cell.
While the clustering analysis from the previous notebook identified groups of spots with similar transcriptional profiles, clustering alone cannot tell us which specific cell types are contributing to each spot's expression.
To move from spot-level expression to cell-type-level information, we can apply **deconvolution**, which decomposes each spot's bulk gene expression into contributions from different cell types and estimates their relative proportions.

Specifically, we'll perform deconvolution using RCTD (Robust Cell Type Deconvolution), which is available in the `spacexr` package ([Cable et al. 2022](https://doi.org/10.1038/s41587-021-00830-w)).
RCTD uses maximum likelihood estimation to model each spot's expression as a weighted combination of reference cell type profiles, using a single-cell reference dataset.
RCTD also accounts for "platform effects" between the reference dataset, which was sequenced with single-cell technology, and the query spatial transcriptomics dataset, as the differences in library preparation between these platforms can introduce artifactual differences between gene expression measurements.


As with reference-based cell type annotation on single-cell data, it's critical for deconvolution to use a reference dataset that contains the cell types you expect to find in your data.
Here, we'll use the [OsteoCAR reference](https://www.biorxiv.org/content/10.64898/2026.01.13.698472v3), a single-cell atlas for osteosarcoma that has been divided into "sub-references" for human and mouse samples, as well as for primary and metastatic samples.
We'll specifically use the mouse metastasis-specific portion of this reference, which we expect will be a reasonably good match to our Visium sample which is also from a mouse osteosarcoma lung metastasis. 


## Set up

To begin, we'll load the packages we'll use throughout this notebook and setting the random seed.

```{r libraries}
suppressPackageStartupMessages({
  # The main class we use for spatial transcriptomic data
  library(SpatialExperiment)
  # Packages to support plotting
  library(ggspavis)
  library(ggplot2)
  library(patchwork)
})

set.seed(2026)
```

### Directories and files

Next, we'll define paths to the normalized SPE object prepared prior to this module, stored in a sample-specific subdirectory of `data/osteo`.

```{r inputs}
# define sample id we'll be analyzing 
sample_id <- "GSM8478586"

# top level data directory
data_dir <- file.path("data")

# data directory for this sample
sample_dir <- file.path(data_dir, "osteo", sample_id, "normalized")

normalized_rds <- file.path(sample_dir, "osteo_normalized_spe.rds")
```

Next, we'll define the file with the OsteoCAR mouse metastasis reference.
Note that we have prepared this object from its original form for use in this notebook by converting it from a Seurat to a `SingleCellExperment` object, converting row names from gene symbols to Ensembl ids (to match our object), and downsampling the number of cells for RCTD compatibility.
You can see the code we use for this pre-processing here: <https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/osteo/reformat-osteo-ref.R>.

```{r input reference}
ref_dir <- file.path(data_dir, "reference")
ref_rds <- file.path(ref_dir, "mm_mets_osteo_ref.rds")
```

We'll conclude this notebook by exporting a TSV with deconvolution results to an `analysis` directory, so let's define this here.

```{r analysis dir}
analysis_dir <- file.path("analysis/osteo")
fs::dir_create(analysis_dir)

rctd_tsv <- file.path(analysis_dir, "osteo_deconvolution.tsv")
```


## Import

### Import the reference SCE

We'll start by reading in the reference SCE object we'll use for deconvolution and define it as `osteo_ref`.

```{r import ref, live = TRUE}
# read in the reference object
osteo_ref <- readr::read_rds(ref_rds)

# show the object
osteo_ref
```

This object contains about 32,000 cells annotated with several cell type labels, which are stored in its `colData` slot:

```{r reference annotations}
# show the different annotation levels
colData(osteo_ref) |>
  as.data.frame() |>
  dplyr::select( starts_with("Ann_Level") )
```

For this notebook, we're going to focus on `Ann_Level1` which contains broad cell type categories, but a more fine-grained analysis might consider `Ann_Level2` or `Ann_Level3`.

Let's see the distribution of cell types in our chosen reference:

```{r count celltypes, live = TRUE}
colData(osteo_ref) |>
  as.data.frame() |>
  dplyr::count(Ann_Level1)
```
RCTD expects (although this can be customized) a maximum of 10,000 cells per cell type in the reference, so we downsampled cells to match this expectation during our pre-processing step.
The original version of this OsteoCAR reference dataset has even more cells. 

### Import the SPE object

Next, we'll read in the normalized SPE object we'll be running deconvolution on.

```{r import spe, live = TRUE}
spe <- readr::read_rds(normalized_rds)
```

Before diving into deconvolution, let's plot the H&E image to reorient ourselves to the tissue.
We'll refer back to this image throughout the notebook when interpreting the spatial cell type patterns we find.

```{r plot-visium}
#| fig.width: 5

# define H&E plot for general use in notebook
he_plot <- plotVisium(spe, spots = FALSE, zoom = TRUE) +
  ggtitle("H&E") +
  # add styling to match other panels
  theme(
    plot.title = element_text(
      hjust = 0.5,
      margin = margin(0, 0, 0.1, 0)
    ),
    panel.border = element_rect(color = "black", linewidth = 0.25)
  )

he_plot
```


Let's also look at some marker gene expression to develop a sense of what cell types we might expect in this data given this reference.
We'll show expression of select marker genes, obtained taken from the OsteoCAR paper, for cell types in the reference.


```{r plot mouse markers}
#| fig.width: 11

mouse_markers <- c(
  "Endothelial cell" = "Esam",
  "Epithelial cell" = "Muc1",
  "Mesenchymal cell" = "Vim",
  "Tumor cell" = "Sox9",
  # immune lymphoid markers
  "B cell" = "Ms4a1",
  "T cell" = "Cd3e",
  # immune myeloid markers
  "Monocyte" = "Cd68",
  "Macrophage" = "Adgre1"
)

mouse_markers |>
  purrr::imap(
    \(gene_symbol, gene_category) {
      
      # translate to ensembl id using SPE rownames
      ensembl_id <- rownames(spe)[rowData(spe)$Symbol == gene_symbol]
      
        plotCoords(
          spe, 
          annotate = ensembl_id, 
          assay = "logcounts", 
          pal = "viridis", 
          point_size = 1
        ) + 
          ggtitle( 
            glue::glue("{gene_category} ({gene_symbol})") 
          ) +
          theme(
            legend.key.width = unit(0.3, "cm"),
            legend.key.height = unit(0.4, "cm"),
            legend.text  = element_text(size = 8),
            title = element_text(size = 10),
            plot.title = element_text(hjust = 0)
          )
    }
  ) |> wrap_plots(ncol = 4)
```

Based on these gene expression plots, what cell types do you think are likely common in the dataset?
Uncommon?
Which cell types do you expect might co-occur?
Does this plot give you any insights about the osteosarcoma tumor microenvironment and the level of immune infiltration?

## Perform deconvolution
Next, we'll use this reference to attempt to determine which cell types are present in each spot and their relative abundances in the osteosarcoma sample.

RCTD models spot expression as a weighted combination of cell type-specific expression profiles from the single-cell reference.
RCTD offers three different modes you can use to specify how many cell types to model for each spot:

1. *Doublet mode:*
This mode constrains the RCTD model to infer a maximum of two cell types per spot, which may or may not be realistic for heterogeneous Visium datasets.
We have to note as well, this concept of "doublet" is not the same as doublets in single-cell analysis, in which multiple cells are sequenced together in a single droplet.
Instead, this means at most two _cell types_ can be present in a spot; the exact number of cells per spot is not known, but we assume it's likely between 5--20.
2. *Multi mode:*
This mode allows the RCTD model to infer multiple cell types per spot, but no more than a specified maximum number of cell types (`max_multi_types`).
This mode is an extension of doublet mode, where the maximum can be raised above 2.
3. *Full mode:*
This mode does not apply any constraints to the RCTD the model, which is allowed to infer an arbitrary number of cell types per spot

With each mode, we'll ultimately end up with a per-cell weight which we can interpret as a proportion.

We'll begin running deconvolution by running RCTD in doublet mode.

### Prepare the data for deconvolution

Next, we'll prepare the data for deconvolution by running the function `spacexr::createRctd()`.
This function sets various values and parameters that will be used for the deconvolution procedure, including:

* Identifying genes to use to correct for platform effects
* Identifying differentially expressed genes across cell types in the reference
* Using those differentially expressed genes, calculating a mean-normalized reference expression matrix that RCTD will use for the deconvolution process itself

We'll leave parameters are their defaults and prepare the data to run deconvolution with `Ann_Level1` labels.

```{r create_rctd, live = TRUE}
# prepare the data for deconvolution
rctd_data <- spacexr::createRctd(
  spe, 
  osteo_ref, 
  cell_type_col = "Ann_Level1"
)
```

We'll briefly look at the contents of `rctd_data`.
It contains a list of several items organized into a structure that RCTD expects.

```{r rctd_data summary, live = TRUE}
summary(rctd_data)
```

For example, here is the Visium data prepared for analysis:

```{r rctd_data SPE}
rctd_data$spatial_experiment
```

We can see that it now only contains a small subset of genes which the model identified for use out of its original 19465 genes.

Finally, we can also see the mean-normalized matrix that RCTD created for running deconvolution. 
It's stored inside `rctd_data$cell_type_info$info`, and we'll show the first few rows below just for context:

```{r show rctd matrix}
rctd_data$cell_type_info$info[[1]] |>
  head()
```

### Perform deconvolution: doublet mode

We're ready to run deconvolution with the `rctd_data` object as input.
We'll begin by running RCTD in doublet mode, which constrains the model to estimate at most two cell types per spot.
This will take 1-2 minutes to run and reports model progress to the Console as it proceeds.

```{r run doublet mode, live = TRUE}
rctd_doublet <- spacexr::runRctd(
  # prepared data for RCTD
  rctd_data, 
  # mode to run
  rctd_mode = "doublet",
  # number of cores to use
  max_cores = 4
) 
```

Now that results are ready, we'll have a look:

```{r show rctd_doublet, live = TRUE}
# show rctd_doublet
rctd_doublet
```

The resulting object is a new SPE object with only 5 features; these are not genes, but represent the cell type labels in the reference object. 
This object also has several assays with the main deconvolution results:

* `weights`: Per-spot cell type proportions 
* `weights_full`: Weights from the initial RCTD model fit, before constraining to two cell types
  * Note that these values are not what precisely we would get from running RCTD "full" mode, as the model parameters here were optimized for "doublet" mode
* `weighted_unconfident`: The same as `weights_full`, but including weights that the model is less confident in

The `weights` assay which contains the fitted proportions is of most interest to us:

```{r doublet weights}
# print all rows across the first four columns of the matrix
assay(rctd_doublet, "weights")[, 1:4]
```

This matrix tells us, for example, that the spot barcoded `AACACCTAAGCATTGC-1` is roughly split between epithelial/endothelial and mesenchymal cells.

This result SPE object also contains some useful information in the `colData` slot:

```{r rctd coldata, live = TRUE}
colData(rctd_doublet)
```

There's a few columns of particular interest here that we'll point out:

* `spot_class`: Whether each spot is more likely a doublet or singlet (i.e., containing a single cell type)
_Importantly_, RCTD in doublet mode does not evaluate whether doublet is an accurate characterization in the first place.
Doublet mode only assesses whether spots are more likely to be singlets or doublets but does not evaluate if enforcing a maximum of two cell types is a good fit to the data.
* `first_type`: The cell type with the higher proportion in the spot
* `second_type`: The cell type with the lower proportion in the spot

Let's see how many doublets and singlets we have according to this model fit:

```{r count spot class, live = TRUE}
colData(rctd_doublet) |>
  as.data.frame() |>
  dplyr::count(spot_class)
```

We see the following labels in this column:

* `reject`: Spots that were filtered out because of either low quality or the reference was a poor fit in the first place
* `singlet`: Under the constrained doublet model, spots that are more likely to have a single cell type
* `doublet_certain`: Under the constrained doublet model, spots that are more likely to have two cell types.
Again, this does not mean that RCTD is entirely certain that the spot is a doublet, only that it is more certain to be a doublet than a singlet.
* `doublet_uncertain`: Under the constrained doublet model, spots that are more likely to be doublets than singlets but with less confidence


We can also color the Visium coordinates by cell type proportions using the provided function `spacexr::plotCellTypeWeight()`.
We'll do this using default settings for the `Tumor` cell type first:

```{r plot tumor proportions, live = TRUE}
#| fig.width: 5

spacexr::plotCellTypeWeight(
  rctd_doublet, 
  cell_type = "Tumor",
  assay_name = "weights", 
  # minimally customize plot styling
  size = 2,
  stroke = 0.1
)
```

How does this plot compare to the regions we suspect are likely tumor-rich according to the H&E?

We'll proceed to plot a grid of coordinate plots, each showing the proportion for a given cell type, and we'll also flip the plot orientation to match the H&E for easier comparison.

To facilitate this, we've written a custom function; run this chunk to be able to use the function next.

```{r plot_all_props function}
# Function to plot a grid of coordinate plots, each showing
# a different cell type's per-spot proportion as present in the given
plot_all_props <- function(rctd_result, assay = "weights") {
  cell_types <- unique(rownames(rctd_result))
  
  cell_types |>
    purrr::map(
      \(cell_type) {
        spacexr::plotCellTypeWeight(
          rctd_result, 
          cell_type,
          assay_name = assay,
          size = 1.5, 
          stroke = 0.1,
          title = cell_type
        ) + 
          # add custom styling
          scale_fill_viridis_c(
            limits = c(0, 1), 
            option = "plasma"
          ) +
          # reverse y-axis orientation to match H&E orientation
          scale_y_reverse() + 
          theme(
            plot.title = element_text(size = 11),
            legend.title = element_text(size = 9),
            legend.text = element_text(size = 9)
          )
      } 
    ) |>
    wrap_plots() + 
    plot_layout(guides = "collect")
}

```

Now, we can use `plot_all_props()` to plot all the cell type proportions in `rctd_doublet`:

```{r plot doublet proportions, live = TRUE}
#| message: false
#| fig.width: 8

plot_all_props(rctd_doublet)
```

What overall patterns do you see?
Does the cell type distribution match your expectations from our earlier marker gene expression plots?

One specific observation here is this model inferred very few, if any, immune lymphoid cells in the data.
We did however observe limited expression of immune lymphoid cells, so it's possible that by using doublet mode, we have missed identifying rare cell type populations.

Therefore, we'll run RCTD once more but this time in "full" mode, which does not impose any constraints on the number of cell types per spot.

### Perform deconvolution: full mode

We can use the same `rctd_data` object as input to full mode:

```{r run full mode}
rctd_full <- spacexr::runRctd(
  rctd_data, 
  rctd_mode = "full",
  max_cores = 4
) 
```

We don't get the same `colData` columns out of this, just the weights as an assay. 
For these weights, we need to normalize to sum to 1 since the model does not constrain them as it does the doublet mode.
For explicitness we'll save this to a new assay called "proportions".

```{r norm full proportions}
assay(rctd_full, "proportions") <- assay(rctd_full, "weights") |>
  # convert to regular matrix for normalization
  as.matrix() |>
  # new in R 4.1, a function to normalize!
  # use 2 for columns
  proportions(margin = 2) |>
  # convert back to sparse matrix
  as("CsparseMatrix")
```


Now, let's plot, again using our custom `plot_all_props()` function, specifying to use the `proportions` assay:


```{r plot full proportions}
#| message: false
#| fig.width: 8

plot_all_props(rctd_full, assay = "proportions")
```

How does this result compare to doublet mode?
Does it change your interpretation of the overall landscape?

Interestingly, we still see that lymphoid immune is still roughly 0 (which again matches expectations from marker gene expression), but if proportions are low it may just be hard to see in the plot.
What exactly are those proportions, and do they represent real signal or noise?
We'll make a histogram of these values:

```{r lymphoid histogram}
# create a data frame of proportions
proportions_df <- assay(rctd_full, "proportions") |>
  # convert to wide data frame
  as.matrix() |>
  t() |>
  as.data.frame() |>
  # put barcodes from row names into a column 
  tibble::rownames_to_column("barcode")

# plot a histogram
ggplot(proportions_df) + 
  aes(x = Immune_Lymphoid) + 
  geom_histogram(fill = "seagreen", color = "gray20") +
  labs(x = "proportion")
```

We see that the values here are indeed quite low.
So, it is noise? Let's think through it:

* Bear in mind the marker gene expression we saw earlier - really not many lymphoid lineage markers. 
* Remember, we only have 5--20 cells per spot here
  * Given this, very low proportions are likely to be noise because there aren't enough cells e.g. for a 1-2\% proportion to be realistic.
  * In practice, people attribute proportions ~< 5-10% to be noise, and it looks like that's about where we are. 
* Both modes therefore show us that lymphoid are rare in this tissue but we can't necessarily rule out minor infiltration since we do see some defining markers. As this reference is pretty coarse-grained, it's possible if we looked more precisely for specific lymphoid cell types we could answer this question more reliably, but either way it seems clear that these cells (if present) are rare. 
  * Bear in mind caveats about how well the reference/data match!


If we assume a 0.1 threshold for calling presence, how many cell types are potentially in each spot?
We'll have to wrangle a bit to answer this.

```{r create proportions_long_df}
# create a long-format data frame of proportions
proportions_long_df <- proportions_df |>
  tidyr::pivot_longer(
    !barcode, 
    values_to = "proportion",
    names_to = "cell_type"
  ) 
```

Now, we can count up the number of cell types per spot, and then summarize this to get an overall distribution:

```{r count cell types per spot}
# first, make a data frame of confident cell types only
present_df <- proportions_long_df |>
  # consider only confident spots, conservatively
  dplyr::filter(proportion >= 0.1) 

# count 
present_df |>
  # count the number of cell types for each spot (how many times each barcode appears in the table)
  dplyr::count(barcode, name = "n_cell_types") |>
  # count the number of spots with each cell type count
  dplyr::count(n_cell_types, name = "n_spots")
```

Using this reference at this annotation level, we see that most spots have likely have 3 cell types.
This suggests to us that the doublet mode did not capture the full heterogeneity, so let's stick with the full mode results.

We'll interrogate further - what cell types do we actually have across spots?
Before we plot, let's define a vector of clearly visible colors for these cell types too:

```{r define cvd friendly pal}
celltype_pal <- c(
  "Epithelial_Endothelial" = "#E69F00",
  "Immune_Lymphoid" = "navy",
  "Immune_Myeloid" = "#009E73",
  "Mesenchymal" = "#F0E442",
  "Tumor" = "#56B4E9" 
)
```

We'll sort spots in this plot in descending order of their Tumor proportion, so we'll do some wrangling first


```{r barplot spot proportions}
#| fig.width: 12
#| fig.height: 4

barcode_order <- proportions_long_df |>
  dplyr::filter(cell_type == "Tumor") |>
  dplyr::arrange(desc(proportion)) |>
  dplyr::pull(barcode)

ggplot(proportions_long_df) + 
  aes(
    # sort barcodes in descending order of tumor weight
    x = forcats::fct_relevel(barcode, barcode_order), 
    y = proportion, 
    fill = cell_type) + 
  geom_col() + 
  scale_fill_manual(values = celltype_pal) +
  scale_y_continuous(expand = c(0,0)) +
  labs(
    y = "proportion of cells in the spot", 
    x = "spot",
    fill = "cell type"
  ) +
  # don't show the barcodes themselves
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

What patterns do you see?
Which cell types tend to co-occur in spots, vs not?

We can answer this question a bit more carefully with more exploratory analysis to ask whether pairs of cell types tend to co-occur in the same spot more than expected.
For this, we'll calculate the ratio of observed to expected co-occurrences of cell type pairs, where expected is like the chi-squared null, where cell types are distributed independently of one another.
We'll log this value to give us a symmetric scale around 0 that aids in interpretation: a value of 0 means they co-occur as often as expected by random chance, negative less than expected, and positive more than expected.
(PS: this metric actually is called the _pointwise mutual information_!)
For this, we'll consider again the 0.1 threshold for calling a cell type as present in each spot.

A few important caveats/pieces of context before we get going:

- this analysis is _unweighted_ - it does not consider the proportion of cells per spot, only presence/absence at the 0.1 threshold
- we don't consider things like spatial autocorrelation, where neighboring spots are more likely to have similar cell type profiles than more distant spots; we consider spots fully independent
- this isn't the same as a neighborhood analysis; we are querying spots, not neighborhoods, and as noted in the caveat above we are considering spots independently
  - a more rigorous analysis might therefore ask,"do tumor cells and epithelial cells tend to be physically close in space" which is a more involved analysis that actually considers neighborhoods and spot proximity


We'll do this in a few steps.
First, starting from our existing `present_df` we'll calculate observed values by join this data frame to itself to find and count all pairwise combinations and count them up.

```{r calculate observed}
# Count observed combinations of cell types
observed_df <- dplyr::inner_join(
  present_df,
  present_df, # join table to itself
  by = "barcode",
  relationship = "many-to-many",
  suffix = c("_a", "_b")
) |>
  # don't consider the same cell types, or duplicate based on order
  dplyr::filter(cell_type_a < cell_type_b) |>
  # count many spots pairs of cell types co-occur in
  dplyr::count(cell_type_a, cell_type_b, name = "observed")

# show data frame
observed_df
```

Next, we need to calculate the number of spots with each cell type which will feed into our observed calculation.


```{r calculate observed independent}
# count how many spots each individual cell type appears in
n_spots_df <- present_df |>
  dplyr::count(cell_type, name = "n_spots")

n_spots_df
```

Finally, we'll wrangle these all together to calculate the expected spots with the formula `(n_spots_with_cell_type_a * n_spots_with_cell_type_a)/number of total spots`, and then calculate the log ratio of observed/expected.


```{r calculate ratio}
# Calculate ratios
cooccur_df <- observed_df |>
  # combine data frames so we have a column for each of:
  # observed counts for a/b pairs
  # observed count for cell type a
  # observed count for cell type b
  dplyr::left_join(n_spots_df, by = c("cell_type_a" = "cell_type")) |>
  dplyr::rename(n_spots_cell_type_a = n_spots) |>
  dplyr::left_join(n_spots_df, by = c("cell_type_b" = "cell_type")) |>
  dplyr::rename(n_spots_cell_type_b = n_spots) |>
  # calculate expected count, and finally the log(observed/expected) ratio
  dplyr::mutate(
    expected = n_spots_cell_type_a * n_spots_cell_type_b / ncol(spe),
    ratio    = log(observed / expected)
  )
```

We'll display these results in a heatmap:

```{r coocurrence heatmap}
# plot a heatmap
ggplot(cooccur_df) + 
  aes(x = cell_type_a, y = cell_type_b, fill = ratio) +
  geom_tile(color = "grey90") +
  # label each tile with its ratio value
  geom_text(
    aes(label = round(ratio, 2)),
    color = "black", 
    fontface = "bold"
  ) +
  # add scale that diverges from 0 in the middle, using ALSF colors!
  scale_fill_gradient2(
    low = "#0053EA", mid = "white", high = "#FDEA91",
    midpoint = 0,
    name = "log ratio"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1)
  )
```
We'll point out a few items:

- Tumor and mesenchymal is the most negative, does this make sense given what we see in the barplot?
- Immune myeloid cells don't seem to co-occur with any other cell type in particular.

Anything else you see?


## Export

We'll conclude by exporting a data frame of the full RCTD results as we stored them in `proportions_long_df` to the `rctd_tsv` file we defined at the beginning.

```{r export tsv, live = TRUE}
readr::write_tsv(proportions_long_df, rctd_tsv)
```


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```

