---
title: "Identifying spatially variable genes"
author: Data Lab for ALSF
date: "2026"
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Introduce the `SpatialFeatureExperiment` and `Voyager` packages
- Demonstrate two methods for identifying spatially variable genes

## Introduction

Coming Soon

## Set up

For our setup, we will start by loading the `SpatialExperiment` package, as we have done before, as well as some familiar packages for plotting.
We'll also set our default theme for `ggplot2`, and set a seed for reproducibility.


### Libraries

```{r setup}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(ggplot2)
  library(patchwork)
})

# Set default ggplot2 theme
theme_set(theme_bw())

# Set seed
set.seed(2026)
```

### Files

The dataset we will be working with is the same pre-normalized osteosarcoma Visium dataset that we have used in previous notebooks.
If you recall, this sample is from a mouse sample with osteosarcoma lung metastases from [Reinecke et al. (2025)](https://doi.org/10.1158/1078-0432.CCR-24-0049).

We'll load in the pre-normalized data to save a number of starting steps.

For a later part of this analysis, we will also want to use some pre-computed results (some of the analyses are slow), so we will define those paths here as well.

```{r inputs}
# define sample id we'll be analyzing
sample_id <- "GSM8478586"

# data directory for this sample
data_dir <- file.path("data/osteo", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "osteo_normalized_spe.rds")

# we will also use some pre-computed results, which can be found in the `analysis` directory
analysis_dir <- file.path("analysis/osteo", sample_id)
moran_mc100k_tsv <- file.path(analysis_dir, "moran_mc100k.tsv")
```


### Import

With all of those paths defined, we will load in our normalized data, which is already formatted as a `SpatialExperiment` object.

```{r import-rds}
spe <- readr::read_rds(normalized_rds)
```

## Identifying Highly Variable Genes

In theory, we could proceed directly to identifying spatially variable genes, but in practice it is often useful to first filter the data to only include highly variable genes (HVGs).
This is because many of the calculations we will be doing for spatial analysis are not particularly fast, so this will help us speed up calculations.
We justify this in part by the fact that if a gene is not particularly variable overall, it is quite unlikely to be spatially variable, so we are not likely to miss many important genes.

We will identify the HVGs using the `scran` package, which is designed for working with single-cell RNA-seq data, but will work here just as well with the spot data.
We'll first model the gene variance to separate the biological component of the variance from the technical component, and then select the top 2000 HVGs by their biological variation. 

```{r find-hvg}
# model gene variance
gene_variance <- scran::modelGeneVar(spe)

# calculate the top 2000 HVGs
hvg_vector <- scran::getTopHVGs(gene_variance, n = 2000)
```

After selecting the HVGs, we can now filter the SPE to include only those genes.
We'll store that in a new variable called `hvg_spe`.

```{r filter-spe, live = TRUE}
# filter the SPE to only include the highly variable genes
hvg_spe <- spe[hvg_vector, ]
```

We can verify that the gene filtering happened by looking at the number of rows in the SPE before and after filtering:

```{r check-rows}
glue::glue("Before filtering: {nrow(spe)} rows.")
glue::glue("After filtering: {nrow(hvg_spe)} rows.")
```



## The `SpatialFeatureExperiment` object

For spatial analysis, it can be convenient to use a slightly different object type from the SpatialExperiment (SPE object that we have been working with so far.
For the rest of this notebook, we will be using `SpatialFeatureExperiment` (SFE) objects, which are an extension of the `SpatialExperiment` (SPE) objects that include additional slots for spatial features, most notably a way to represent a graph that defines how cells are connected to each other in space.

Almost all functions that work with SPEs will also work with SFEs (though not the reverse).
However, there are differences with respect to how images are stored and plotting defaults, so we will be using a different package for most of our spatial plots. Rather than `ggspavis`, we will be using the `Voyager` package, which is designed for working with SFE objects.
It works a bit differently, but still uses `ggplot2` as the underlying plotting engine, so we can still use the same kinds of `ggplot2` functions to modify the plots if we want to.


### Converting from SPE to SFE

The conversion from SPE to SFE is fairly straightforward, but there are a few things to be aware of.
The most important one right now is that  that the plotting functions we are going to use have different defaults:

`ggspavis` plots the origin of the spatial features in the upper left, while `Voyager` plots it in the lower left, so to match the way we have been looking at things, we will mirror the plots.
(If you were doing all your plotting using a `Voyager` function, you might not need to do this, as the orientation is arbitrary, but since we have been using `ggspavis` for most of our plots so far, we will mirror the coordinates to match what we have been looking at.)
To make things even more confusing `ggspavis` also automatically flips the image to match it's preferred orientation, but `Voyager` expects that this has been done when the images are imported, so we have to do another flip for just the image.
How did we find this out?
Trial and error...

To perform the initial conversion, we will use the `SpatialFeatureExperiment::toSpatialFeatureExperiment()` function, which takes an SPE object and converts it to an SFE object.
We will also specify the units for the spatial coordinates, which in this case are in pixels. 
This isn't really required, and we won't do anything with it except to show that the SFE object supports it.

Af the initial conversion function, we use `SpatialFeatureExperiment::mirror()` to flip the Y axis coordinates and then flip the image a second time using `SpatialFeatureExperiment::mirrorImg()`.
Just note that the exact series of flips you need are going to depend on how the data was read in the first time. 

```{r sfe-conversion, live = TRUE}
sfe <- SpatialFeatureExperiment::toSpatialFeatureExperiment(
  hvg_spe,
  unit = "full_res_image_pixel"
) |>
  # flip the coordinates to match how we have been looking at them
  SpatialFeatureExperiment::mirror() |>
  # we have to flip the image again to match: it was upside down
  SpatialFeatureExperiment::mirrorImg()
```

Now we have an SFE object, and we can look at the contents to see what is in there:

```{r SFE-contents, live = TRUE}
sfe
```

The first part looks pretty much the same as the SPE object that you should be familiar with at this point:
- `assays` with `counts` a`nd `logcounts`
- rows for genes, columns for spots
- reducedDims for PCA and UMAP (not present here yet)
- `spatialCoords` for the X and Y coordinates of the spots
- `imgData` for the image data

But you will notice a few extra slots in the summary:

- `unit` is where the unit we defined earlier is stored.
- `Geometries` stores the shapes and locations of features. In our case, we have only the `colGeometries`, which stores the location of each spot.
- `Graphs`: stores the graphs that define the relationships between spots or cells, though we do not have anything defined here yet.

One thing you may notice here is that the `SpatialFeatureExperiment` package is built from the start to accommodate multiple samples, so the graphs will be divided up by sample.

### Generating spatial graphs

The next step is to define the graph that defines how cells are connected to each other in space.
For Visium data, we can use `SpatialFeatureExperiment::findVisiumGraph()` which uses the known barcodes and their locations to define a graph that connects nearby spots.
This is a nice shortcut, but for other kinds of spatial data (such as image-based Xenium data), we would have to use a function like `SpatialFeatureExperiment::findSpatialNeighbors()` to find the nearest neighbors for each cell, and then define the graph based on that.

We will store that in the `colGraph` slot of the SFE object (this is one of the things we get from that package)

```{r define-graph, live = TRUE}
# use the Visium barcodes to define the graph
SpatialFeatureExperiment::colGraph(
  sfe
) <- SpatialFeatureExperiment::findVisiumGraph(sfe)

# view contents again
sfe
```

Now we can make our first plot using the `Voyager` package, which is designed for working with SFE objects.
We'll plot the graph we just made.


```{r plot-graph, live = TRUE}
Voyager::plotColGraph(sfe) +
  # the units are arbitrary, so let's leave them and the grid lines off
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```

This includes all of the spots where we have data.


## Identify Spatially Variable Genes

### Moran's I

The measure of spatial covariation we will use is Moran's *I*.
This is a measure of spatial autocorrelation, or how similar values are at nearby locations.

Comes from Patrick Moran, who I recognize from population genetics (Moran process is an overlapping generations model for evolution)

Once we have the graph defined, we can calculate Moran's *I* for each gene using the `Voyager::runMoransI()` function.

```{r moranI, live = TRUE}
sfe <- Voyager::runMoransI(sfe)
```

This gives us an I value for each gene as well as the kurtosis of the gene's expression (how fat the tails are) as a value `K`.
TODO: add better kurtosis explanation (what does high or low mean?)

```{r view_moran}
rowData(sfe)
```

You'll notice that the new column names include the `sample_id`, which is a bit annoying, but it is necessary to keep things separate if we have multiple samples.
We'll make sure to set that as a variable so we can reuse it as needed.

Let's first look at the distribution of Moran's I values across all (highly variable) genes:


```{r moran-hist}
rowData(sfe) |>
  as.data.frame() |>
  ggplot(aes(x = .data[[paste0("moran_", sample_id)]])) +
  geom_histogram(bins = 50) +
  labs(
    x = "Moran's I",
    y = "Number of genes"
  )
```


We get a value there, but no measure of significance.

We can calculate those using the more general `Voyager::runUnivariate()` function, which can do other types of tests as well
For that we can use either `moran.test()` or `moran.mc()` from the `spdep` package.
TODO: clarify where these functions are coming from

`moran.test` is relatively fast, so we can do that first (though it is not likely to be as accurate, as it does not account for the structure of the graph).


```{r moran-test, live = TRUE}
sfe <- Voyager::runUnivariate(
  sfe,
  type = "moran.test",
  BPPARAM = BiocParallel::MulticoreParam(4)
)

rowData(sfe)
```


We could also do a more careful test based on Monte Carlo simulation, but this will be slow, as we essentially we have to resample and recalculate for some number of replicates.
We'll start with 1000 replicates for time.

```{r moran-mc, live = TRUE}
reps <- 1000
sfe <- Voyager::runUnivariate(
  sfe,
  type = "moran.mc",
  nsim = reps,
  BPPARAM = BiocParallel::MulticoreParam(4)
)
```


```{r view-mc-result}
rowData(sfe)
```


Let's compare P values for the basic and Monte Carlo versions:

```{r moran-dataframe}
# define the sample_id so we can use it as needed
sample_id <- "GSM8478586"

moran_df <- rowData(sfe) |>
  as.data.frame() |>
  dplyr::select(
    ID,
    moran_i = paste0("moran_", sample_id),
    moran_test_p = paste0("moran.test_p.value_", sample_id),
    moran_mc_p = paste0("moran.mc_p.value_", sample_id)
  ) |>
  dplyr::mutate(
    moran_test_neglogp = -log10(moran_test_p),
    moran_mc_neglogp = -log10(moran_mc_p)
  )
```


```{r plot-p}
ggplot(moran_df, aes(x = moran_test_neglogp, y = moran_mc_neglogp)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I -log10(P value) (moran.test)",
    y = "Moran's I -log10(P value) (moran.mc)"
  ) +
  # restrict the values since some moran.test values are very large
  # which makes it hard to see the rest of the points
  xlim(0, 7) +
  # add a y=x line for comparison
  geom_abline(slope = 1, intercept = 0, color = "forestgreen")
```

Not too bad in the broad strokes, but definitely starts to disagree as we get to higher levels of significance.
This is largely driven by the fact that the smallest P value we can get with the Monte Carlo test is 1/(reps + 1), which is 0.001 in this case, while the parametric test can go much lower than that.

We could do better with more replicates, but that takes a very long time to run... so we pre-calculated a set with 100,000 replicates.

```{r read-moran_mc100k}
moran_mc100k_df <- readr::read_tsv(moran_mc100k_tsv) |>
  dplyr::select(ID, moran_mc100k_p = moran.mc_p.value_GSM8478586)

moran_df <- moran_df |>
  dplyr::left_join(moran_mc100k_df, by = "ID") |>
  dplyr::mutate(
    moran_mc100k_neglogp = -log10(moran_mc100k_p)
  )
```


```{r plot-moran-mc-100k}
ggplot(moran_df, aes(x = moran_test_neglogp, y = moran_mc100k_neglogp)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I -log10(P) (moran.test)",
    y = "Moran's I -log10(P) (moran.mc 100k)"
  ) +
  xlim(0, 7) +
  geom_abline(slope = 1, intercept = 0, color = "forestgreen")
```

It seems like the parametric approximation tends to overestimate significance at smaller P values.
(Though this does not say that either is accurate; just that they agree)
TODO: expand a bit on P value accuracy (bad model -> bad stats)

Let's apply a multiple testing correction to the P values, and assign significance for each at an FDR of 0.05.

```{r moran-significant}
moran_df <- moran_df |>
  dplyr::mutate(
    moran_test_p_adjust = p.adjust(moran_test_p, method = "BH"),
    moran_mc_p_adjust = p.adjust(moran_mc100k_p, method = "BH"),
    moran_test_sig = moran_test_p_adjust < 0.05,
    moran_mc_sig = moran_mc_p_adjust < 0.05
  )
```

```{r count-moran-significance, live = TRUE}
moran_df |>
  dplyr::count(moran_test_sig, moran_mc_sig)
```

Either way, we get the vast majority of genes as "significant" which is not too surprising.
Using significance as a filter here is not likely to be very useful.


Let's make a plot of the top few genes:

```{r get-top-moran, live = TRUE}
top_moran <- moran_df |>
  dplyr::slice_max(moran_i, n = 6) |>
  dplyr::pull(ID)
```


We'll use `Voyager` for the plot.

Much like `ggspavis` in results, but maybe some nicer defaults?

```{r plot-top-moran, live = TRUE}
Voyager::plotSpatialFeature(
  sfe,
  features = top_moran,
  ncol = 2,
  # use gene symbols instead of ensembl IDs for the plot
  swap_rownames = "Symbol"
)
```

### `nnSVG`

Let's try `nnSVG` now, since that is more specifically designed for this kind of data.

TODO: Add some more context about the method

The main advantage that it has is that it can model the spatial variation in a more flexible way, allowing for different scales of spatial variation.

We could perform this analysis on all genes, but again for now we will restrict ourselves to HVGs for time.

Note that this function works either on an SFE or an SPE object, but we will use the SFE object here since we already have it.

```{r run-nnsvg, live = TRUE}
sfe <- nnSVG::nnSVG(sfe, n_threads = 4)
```

Pause here for a break.
This is much slower, but we get significance calculations by default.


```{r view-nnsvg}
rowData(sfe)
```

Note that the added names are more generically named, in part because this package is designed to work with the simpler SPE objects, so there is less accommodation for multiple samples in a single object.


```{r count-nnsvg-significance, live = TRUE}
rowData(sfe) |>
  as.data.frame() |>
  dplyr::filter(padj < 0.05) |>
  nrow()
```

Again, most of them!
So most genes are significantly spatially variable, which is perhaps not a huge surprise.

Other stats we can look at include `prop_sv`, which is the proportion of variation that is attributable to spatial variation.

How do the P values compare across methods?

```{r collate-results}
svg_df <- rowData(sfe) |>
  as.data.frame() |>
  dplyr::select(
    ID,
    nnsvg_stat = LR_stat,
    nnsvg_propsv = prop_sv,
    nnsvg_p = pval,
    nnsvg_p_adjust = padj
  ) |>
  dplyr::mutate(
    nnsvg_neglogp = -log10(nnsvg_p)
  ) |>
  dplyr::left_join(
    moran_df,
    by = "ID"
  )
```


```{r plot-moran-nnsvg-p}
ggplot(svg_df, aes(x = moran_test_neglogp, y = nnsvg_neglogp)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I -log10(P) (moran.test)",
    y = "nnSVG -log10(P)"
  ) +
  xlim(0, 5)
```

Rather than comparing just the P values, we could also look at the correlation between I and `nnsvg`'s `prop_sv` statistic.

```{r plot-moran-nnsvg-stats}
ggplot(svg_df, aes(x = moran_i, y = nnsvg_propsv)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I",
    y = "Proportion of variance explained by spatial location"
  ) +
  annotate(
    geom = "label",
    x = 0.5,
    y = 0,
    label = paste0(
      "Spearman correlation: ",
      round(cor(svg_df$moran_i, svg_df$nnsvg_propsv, method = "spearman"), 2)
    )
  )
```

So for a lot of genes we get very similar scores, but `nnSVG` is uncovering some genes as spatially variable that Moran's I does not, presumably because it models the spatial variation in a more flexible way.

Let's look at the top few:

```{r}
top_nnsvg <- svg_df |>
  dplyr::slice_max(nnsvg_propsv, n = 6) |>
  dplyr::pull(ID)

Voyager::plotSpatialFeature(
  sfe,
  features = top_nnsvg,
  ncol = 2,
  swap_rownames = "Symbol"
)
```

This does seem to detect more punctate expression patterns, where expression of some genes is found in specific distinct spots rather than varying across larger scales.


### Comparing SVGs and HVGs

We can also look at how the variation measurements we get from the SVGs compares to the biological component of the variation in the HVGs from earlier.


```{r compare-svg-hvg, live = TRUE}
svg_hvg_df <- svg_df |>
  dplyr::inner_join(
    gene_variance |>
      as.data.frame() |>
      tibble::rownames_to_column("ID") |>
      dplyr::select(ID, bio_variance = bio),
    by = "ID"
  )
```


```{r plot-svg-hvg}
ggplot(svg_hvg_df, aes(x = bio_variance, y = nnsvg_propsv)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Biological variance",
    y = "Proportion of variance explained by spatial location"
  ) +
  annotate(
    geom = "label",
    x = 2,
    y = 0,
    label = paste0(
      "Spearman correlation: ",
      round(cor(svg_hvg_df$bio_variance, svg_hvg_df$nnsvg_propsv, method = "spearman"), 2)
    )
  )
```

Genes which have more biological variation do tend to have higher proportions of spatial variation, but the reverse is not always true!



## Session info

```{r}
sessionInfo()
```

