---
title: "Importing, processing, and exploring Xenium data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Read Xenium data into R
- Calculate quality control measures on imaging spatial transcriptomic data
- Remove likely low-quality spots with `SpaceTrooper()`
- Normalize spatial expression data

## Introduction

[Xenium](https://www.10xgenomics.com/platforms/xenium) is 10x Genomics' _imaging-based_ in situ platform.
Unlike sequencing-based methods such as single-cell RNA-seq or Visium, Xenium does not sequence anything.
Instead, it measures a targeted panel of a few hundred genes directly inside intact tissue by repeated rounds of fluorescence imaging, then computationally decodes those images into individual transcript molecules.

There are two big differences to be aware of when processing Xenium data:

- Resolution is subcellular.
Each detected transcript has an `(x, y, z)` location in the tissue, and transcripts are assigned to individual cells identified by image segmentation. 
So the columns of our data are _cells_, not the multi-cell _spots_ you get with Visium.

- The gene set is fixed and small.
We only measure the genes on the panel (here, 313 genes), so we can't discover expression of genes that weren't targeted, but we get very high sensitivity and spatial precision for the ones we do measure.

<!--TODO: Add in appropriate links-->
In this notebook we read in a Xenium human breast cancer dataset (the Janesick et al. dataset), compute quality-control metrics, remove low-quality cells, and normalize the expression values.

## Set up

To begin, we'll load `SpatialExperiment`, the core package for spatial transcriptomic objects in Bioconductor, as well as some packages for plotting.

```{r libraries}
# The main class we use for spatial transcriptomic data
library(SpatialExperiment)

# To support plotting
library(ggplot2)
library(patchwork)

# Set ggplot2 theme
theme_set(theme_bw())
```

### Directories and files

Xenium data comes as an output bundle produced by the Xenium Onboard Analysis (XOA) pipeline.
The specific files present will depend on the Xenium software version and assay, but they are all described in detail in the [10x Genomics documentation](https://www.10xgenomics.com/support/software/xenium-onboard-analysis/latest/analysis/xoa-output-understanding-outputs).

We've placed the Xenium output bundle for this breast cancer sample in the directory `data/brca-xenium/xenium`, organized as XOA returns it.

<!--TODO: Describe how cell typing was done-->
We also have an annotations file containing cell type annotations from the original publication. 

We'll define the path to both the Xenium output and the annotations file here. 

```{r inputs}
# main data directory for this sample
data_dir <- file.path("data/brca-xenium")

# directory with xenium output
xenium_outs_dir <- file.path(data_dir, "xenium")

# cell type annotations file
annotations_file <- file.path(data_dir, "Janesick_annotations.tsv")
```


And now, we'll define the output paths and files we'll use for exporting.

```{r outputs, live = TRUE}
# Directory and file to save output
normalized_dir <- file.path(data_dir, "normalized")

# create the directory if it does not exist
fs::dir_create(normalized_dir)

# output RDS file for normalized Spatial Experiment (SPE) object
output_spe_file <- file.path(
  normalized_dir,
  "brca_xenium_normalized.rds"
)

# output cell type color palette
output_color_palette_file <- file.path(
  data_dir, 
  "brca-xenium-cell-type-palette.tsv"
)
```


## Exploring Xenium outputs

With Visium you run Space Ranger yourself, starting from FASTQ files. 
Xenium is imaging-based, not sequencing-based, so there are no FASTQs. 
Instead, the XOA pipeline runs simultaneously on the instrument during data collection. 
So by the time you get data, the output is already analyzed. 
The instrument: 

- decoded transcripts from the images
- segmented cells 
- assigned transcripts to cells 
- built the count matrix

Let's look at what's in the bundle:


```{r dir-data, live = TRUE}
dir(xenium_outs_dir)
```

Below are brief descriptions of the core files in a Xenium output bundle.
Not every file is always present since it depends on the software version and the exact assay used.

<!--TODO: Fill in file specific information -->
##### `experiment.xenium`

A JSON file of run-level metadata: the gene panel, pixel size, software versions, and file paths.
This is the entry-point file that most loaders (including the one we use below) read first.

##### `cell_feature_matrix.h5` (and `cell_feature_matrix/`, `cell_feature_matrix.zarr.zip`)

Analogous to the feature barcode matrix output from Cell Ranger/Space Ranger. 

##### `cells.parquet` / `cells.csv.gz` (and `cells.zarr.zip`)

This is the file that provides our spatial coordinates and column (cell) metadata.

##### `transcripts.parquet` / `transcripts.csv.gz`

This is the subcellular data that distinguishes Xenium from spot-based methods.
This is analogous to the `tissue_positions_list.csv` for Visium.

##### `cell_boundaries.parquet` and `nucleus_boundaries.parquet` (plus `.csv.gz` versions)

The polygon vertices defining each cell's and each nucleus's segmentation outline.
<!--TODO: Clarify how these relate to the count matrix which has a cell (not nucleus) dimension -->

##### `analysis/` and `analysis.zarr.zip`

The instrument's automated secondary analysis.

##### `metrics_summary.csv` and `analysis_summary.html`

Summary information as CSV and HTML 

##### Morphology images (`morphology.ome.tif`, `morphology_mip.ome.tif`, `morphology_focus.ome.tif` / `morphology_focus/`)

We don't have these to save space (as of right now), but these are the tissue morphology/DAPI images

### Import with `SpatialExperimentIO`

We'll import the counts and spatial information into a `SpatialExperiment` (SPE) object using the `SpatialExperimentIO` package.
Note that here we are not using `VisiumIO` as our dataset is from Xenium, which is distinct from the previous Visium data sets we have been working with.

The only argument the function strictly requires is the path to the output bundle directory.
Below we also pass a few arguments explicitly (count matrix and per-cell metadata) just so we can see what files are required, but we are still using the defaults.

By default the function also moves the Xenium control features (negative control probes, negative control codewords, unassigned codewords, etc.) out of the main gene matrix and into `altExps`, and adds the `experiment.xenium` metadata and the transcript/boundary parquet paths to `metadata()`.

```{r import-xenium, live = TRUE}
spe <- SpatialExperimentIO::readXeniumSXE(
  xenium_outs_dir,                            # path to the Xenium bundle (required)
  countMatPattern = "cell_feature_matrix.h5", # default
  metaDataPattern = "cells.parquet",          # default
  coordNames = c("x_centroid", "y_centroid")  # default; columns used as spatialCoords
)
```

Let's have a look:

```{r print-spe, live = TRUE}
# print the spe
spe
```

A few things to notice about this object:

- The dimensions show only 313 genes but roughly 167,000 cells.
  The gene count is small because Xenium measures a targeted panel, and because the control features have been split off into `altExps` (so the main matrix is genes only).
- Each column corresponds to one cell.
There are no barcodes, or spot barcodes, just cell ids.
- The `spatialCoords` hold the `x_centroid`/`y_centroid` values in microns, while gene-level counts live in the `counts` assay.

Let's go ahead and plot the coordinates here. 
Unlike with Visium, Xenium does not natively have an H&E image. 
This is an image that can be grabbed separately and added to the object downstream, but we won't be working with one with this dataset. 

```{r plot-coords, live = TRUE}
#| fig.width: 7
# needed to be able to plot the coordinates with ggspavis
spe$in_tissue <- TRUE

# plot just the spots
ggspavis::plotCoords(spe) 
```

### A note on segmentation 

TODO: Fill in with text talking about segmentation 

### A note on control features and `altExps`

A Xenium panel does not only include genes.
<!--TODO: Explain what codebook and codewords are-->
The codebook also contains several categories of control features that let us estimate background and error rates:

- Negative control probes target sequences that are _not_ present in the transcriptome, so any counts reflect nonspecific probe binding (background).
- Negative control codewords are valid codewords deliberately left unassigned to any gene, so counts reflect optical/decoding errors.
- Other categories (unassigned/deprecated codewords, and, for some panels, genomic or antisense controls) exist too.

So we don't mix these with real gene counts, `readXeniumSXE()` moves them into alternative experiments stored alongside the main object.
These can be accessed with `altExp`. 

We can see which control experiments were separated out here:

```{r inspect-altexps, live = TRUE}
# control features are stored separately from the gene panel, as altExps
altExpNames(spe) |> 
  purrr::map(\(name) altExp(spe, name))
```

## Add in cell type annotations 

For this dataset, we have published cell type annotations, which we can join onto our object's cell metadata.

```{r read-annotations, live = TRUE}
# read the cell type annotations
cell_types_df <- readr::read_tsv(annotations_file)

head(cell_types_df)
```

```{r annotations-table}
# how many cells of each annotated type?
table(cell_types_df$cell_type_annotation)
```

The original annotations are fairly fine-grained.
To make plots easier to read, we'll collapse related labels into broader groups in a new `cell_types_lumped` column.

```{r lump-annotations}
cell_types_df <- cell_types_df |>
  dplyr::mutate(
    cell_types_lumped = dplyr::case_when(
      stringr::str_detect(cell_type_annotation, "Macrophages") ~ "Macrophage",
      stringr::str_detect(cell_type_annotation, "DCIS") ~ "DCIS",
      stringr::str_detect(cell_type_annotation, "Myoepi") ~ "Myoepithelial cell",
      stringr::str_detect(cell_type_annotation, "Hybrid") ~ "Hybrid",
      stringr::str_detect(cell_type_annotation, "Tumor") ~ "Tumor",
      stringr::str_detect(cell_type_annotation, "DCs") ~ "Dendritic cell",
      .default = cell_type_annotation
    ) |>
      stringr::str_replace_all("_", " ")
  )

# check the counts of the lumped categories
cell_types_df |>
  dplyr::count(cell_types_lumped) |> 
  dplyr::arrange(desc(n))
```


Now we join the annotations onto the object's `colData` by `cell_id`.
Cells that are not present in the annotation file will get `NA`.

```{r add-annotations}
coldata_df <- colData(spe) |>
  as.data.frame() |>
  dplyr::left_join(cell_types_df, by = "cell_id")

colData(spe) <- DataFrame(coldata_df)

head(colData(spe))
```

To keep colors consistent across plots, we'll build a single named palette mapping each lumped cell type to a color, and reuse it.

```{r celltype-palette}
# build a custom palette 
cell_type_palette <- list(
  "Stromal" = "#F6222E",
  "Tumor" = "#3283FE", 
  "DCIS" = "#325A9B",
  "Macrophage" = "#85660D",
  "Myoepithelial cell" = "#F7E1A0",
  "Endothelial" = "#AA0DFE",
  "CD4+ T Cells" = "#16FF32",
  "CD8+ T Cells" = "#1C8356",
  "B Cells" = "#FC1CBF",
  "Perivascular-Like" = "#FEAF16",
  "Dendritic cell" = "#1CBE4F",
  "Mast Cells" = "#565656",
  "Hybrid" = "#FFFF00",
  "Unlabeled" = "grey90"
)
```

```{r set-celltype-levels}
# re-order celltypes for plotting by setting them as a factor
spe$cell_types_lumped <- as.factor(spe$cell_types_lumped) |> 
  forcats::fct_relevel(names(cell_type_palette))
```


```{r plot-celltypes, live = TRUE}
#| fig.width: 8

# plot all spots colored by cell type 
ggspavis::plotCoords(spe, annotate = "cell_types_lumped") +
  scale_color_manual(values = cell_type_palette)
```


## Identifying low quality spots 

We use `SpaceTrooper` to compute per-cell quality-control metrics that are tailored to imaging-based spatial data (e.g., cell area, total counts, and control-feature counts).
`SpaceTrooper` can be used with many different imaging based spatial transcriptomics platforms (e.g., CosMx, MERFISH). 
Therefore we need to specify which technology produced the data, which we set in the object's `metadata()` first.

<!--TODO: Note about the warning being expected for Xenium data since that metric is not calculated upon input. 
Ultimately the metric is used for generating a QC score that is specifically useful for CosMx data and other data that have stitching artifacts within their FOV border-->

```{r spatial-qc}
# tell SpaceTrooper which platform this is
metadata(spe)$technology <- "10X_Xenium"

# compute spatial per-cell QC metrics (adds columns like cell area, and probe counts)
# note that some of this may have been pre-calculated for this experiment but we'll re-do that here
spe <- SpaceTrooper::spatialPerCellQC(
  spe,
  use_altexps = FALSE # ignore altexps as we won't use these
)

head(colData(spe))
```

<!--TODO: Add in some text about the new columns that are added-->

`spatialPerCellQC()` writes the centroid coordinates into `colData` as `x_centroid`/`y_centroid`.
Because `spatialCoords()` already holds columns with those exact names, `ggspavis` errors on the duplicated names when it assembles its plotting data frame.
We drop the `colData` copies (the coordinates still live safely in `spatialCoords`):

```{r dedupe-coords}
# remove the duplicated coordinate columns from colData to avoid a name
# collision with spatialCoords when plotting
colData(spe) <- colData(spe) |>
  as.data.frame() |>
  dplyr::select(-c("x_centroid", "y_centroid")) |>
  DataFrame()
```

Low-quality cells tend to be outliers in the relationship between total counts and cell area. 
Cells with very few counts for their area are often debris or segmentation artifacts. 
Below we can color the cells by these QC metrics. 

```{r plot-qc}
#| fig.width: 8

p1 <- ggspavis::plotCoords(spe, annotate = "total") +
  ggplot2::scale_color_distiller(palette = "Blues", direction = 1)

p2 <- ggspavis::plotCoords(spe, annotate = "Area_um") +
  scale_color_viridis_c() +
  ggplot2::scale_color_distiller(palette = "YlOrRd", direction = 1)

p1 + p2
```

We'll then calculate outliers by finding the ratio between total counts and area and getting the threshold for three median absolute deviations (MADs) below the median. 

```{r count-area-outliers}
# counts-per-area ratio for each cell
count_area_ratio <- spe$total / spe$Area_um

# flag cells that are outliers on the low side
# (3 median absolute deviations below the median, on the log2 scale since the metric is a ratio)
outliers <- scater::isOutlier(count_area_ratio, log = TRUE, type = "lower", nmads = 3)

# print the output
outliers
```


```{r define-threshold, live=TRUE}
# pull out the numeric threshold for plotting
threshold <- attr(outliers, "threshold")[1]
```

```{r plot-threshold}
# visualize the distribution and the chosen threshold
ggplot(mapping = aes(log(count_area_ratio))) +
  geom_histogram(bins = 100, fill = "grey90", color = "black") +
  theme_bw() +
  geom_vline(xintercept = log(threshold))
```



## Filtering low quality spots

Now we can filter. 

In addition to the counts-per-area outliers, we use the control features as a quality signal.
A high-quality cell should have essentially no control counts. 
Any counts in these categories indicate background or errors in that cell, so we exclude cells with a nonzero count in either control category.

```{r flag-exclude}
# a cell is excluded if it is a counts-per-area outlier OR has any control counts
exclude <- outliers |
  spe$control_probe_counts > 0 |
  spe$control_codeword_counts > 0

# fraction of cells flagged for removal
sum(exclude) / length(exclude)
```

```{r plot-exclude}
# store the flag and visualize which cells will be excluded
spe$exclude <- exclude

ggspavis::plotCoords(spe, annotate = "exclude") +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  # add custom colors so excluded cells are easy to see
  scale_color_manual(values = c("grey90", "#00274C"))
```

```{r filter-cells, live = TRUE}
# keep only the cells not flagged for removal
filtered_spe <- spe[, !spe$exclude]

dim(filtered_spe)
```


## Normalization 

Following the recommendations in [OSTA](https://bioconductor.org/books/release/OSTA/pages/img-intermediate-processing.html#normalization), we'll use a simple library-size normalization. 
This is the same normalization we would use for Visium data. 

```{r normalize, live = TRUE}
# calculate per-spot size factors
normalized_spe <- scuttle::computeLibraryFactors(filtered_spe)

# perform normalization
normalized_spe <- scuttle::logNormCounts(normalized_spe)
```


## Export

```{r export-spe}
readr::write_rds(normalized_spe, output_spe_file)
```

```{r export-palette}
cell_type_palette |> 
  tibble::enframe(name = "cell_type", values = "color") |> 
  readr::write_tsv(output_color_palette_file)
```


## Session info

To conclude, we'll run the `sessionInfo()` command to print out exactly what system setting and R package versions were used to run this code.

```{r}
sessionInfo()
```


