---
title: "Deconvolution of spatial transcriptomics data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Use `hoodscanR` to create a neighborhood profile for each cell
- Use the output from `hoodscanR` to quantify the complexity of spatial neighborhoods 
- Identify cell types that are commonly co-localized 
- Use `hoodscanR` for domain detection

## Introduction 

## Setup 

To begin, we'll load the packages we'll use throughout this notebook and set the random seed.

```{r libraries}
suppressPackageStartupMessages({
  # The main class we use for spatial transcriptomic data
  library(SpatialExperiment)
  # functions for plotting
  library(ggplot2)
})

# Set theme for ggplot2 plotting
theme_set(theme_bw())

set.seed(2026)
```

### Directories and files

Next, we'll define paths to the normalized SPE object that we prepared in the previous notebook. 
This object is stored in `data/brca-xenium/normalized`. 

```{r inputs}
# top level data directory for brca xenium data
data_dir <- file.path("data", "brca-xenium")

# data directory for the normalized data
normalized_dir <- file.path(data_dir, "normalized")

normalized_rds <- file.path(normalized_dir, "brca_xenium_normalized.rds")
```

```{r outputs}
# TODO 

```


Now we'll read in the normalized object. 

```{r}
spe <- readr::read_rds(normalized_rds)
```

As a reminder, let's look at the cell types that are present in this dataset. 
We'll use the `cell_types_lumped` column that we created in the previous notebook for the plot and for the neighborhood analysis.

Since this is a new notebook, we'll need to set the palette of colors to use for each cell type again. 

```{r celltype-palette}
# build a custom palette 
cell_type_palette <- list(
  "Stromal" = "#F6222E",
  "Tumor" = "#3283FE", 
  "DCIS" = "#325A9B",
  "Macrophage" = "#85660D",
  "Myoepithelial cell" = "#F7E1A0",
  "Endothelial" = "#AA0DFE",
  "CD4+ T Cells" = "#16FF32",
  "CD8+ T Cells" = "#1C8356",
  "B Cells" = "#FC1CBF",
  "Perivascular-Like" = "#FEAF16",
  "Dendritic cell" = "#1CBE4F",
  "Mast Cells" = "#565656",
  "Unlabeled" = "grey90"
)

# unlabeled should be greyed out
cell_type_palette$Unlabeled <- "grey90"
```

```{r set-celltype-levels}
# re-order celltypes for plotting by setting them as a factor
spe$cell_types_lumped <- as.factor(spe$cell_types_lumped) |> 
  forcats::fct_relevel(names(cell_type_palette))
```

```{r}
#| fig.width: 8

ggspavis::plotCoords(spe, annotate = "cell_types_lumped") +
  scale_color_manual(values = cell_type_palette)
```



## Identifying cell neighborhoods with `hoodscanR` 

TODO: Fill in with general explanation of `hoodscanR

### Setup the `hoodscanR` object

Also need text to explain creating the object used for `hoodscanR`

```{r setup-hood}
spe_hood <- hoodscanR::readHoodData(
  spe, 
  anno_col = "cell_types_lumped"
)

# print the modified object
spe_hood
```


```{r coldata-hood}
head(colData(spe_hood))
```

note that the cell_types_lumped column has been replaced with cell_annotation

TODO: Fill in with text explaining the new output and why this object is different from a regular SPE

### Generate the probability matrix

TODO: Add dicussion about identifying each "neighborhood"
include information about choosing k and tau 
 
```{r nearest-neighbors}
neighbors <- hoodscanR::findNearCells(
  spe_hood, 
  k = 100, # default
  anno_col = "cell_annotation" # default
)
```

What's inside this object? 
We have `cells` and `distance`.
Both are matrices. 

```{r}
neighbors$cells |> 
  head()
```


```{r}
neighbors$distance[1:10, 1:10]
```


Generate the actual probability matrix, discuss how this is done in practice
The distance matrix is used for this calculation. 

Here's actually where we need to talk about Tau in more detail. 
TODO: what are the different modes? We should mention what the other mode is and why we are using this one. 

```{r prob-mtx}
prob_mtx <- hoodscanR::scanHoods(neighbors$distance, mode = "proximityFocused")
```
```{r}
prob_mtx[1:10, 1:10]
```

TODO: Explain the numbers in this output and that each row is a cell and each column is neighbors 
The value corresponds to the probability of each cell associating with their nearest 100 neibhors

Now we need to merge the probabilities by cell types of the 100 neighbors 
TODO: How does this merging work? 
Requires the annotation matrix and the probability matrix

```{r merge-probabilities}
hoods <- hoodscanR::mergeByGroup(
  prob_mtx, 
  neighbors$cells
)
```


```{r}
hoods[1:10, ]
```

TODO: Below we can turn these probabilities into a heatmap
We aren't going to use the plotHoodMat since it's hard to make adjustments and its just ComplexHeatmap under the hood. 


```{r}
# plot the neighborhood profiles 

ComplexHeatmap::Heatmap(
  hoods[1:100,],
  col            = scico::scico(200, palette = "acton", direction = -1),
  rect_gp        = grid::gpar(col = "white", lwd = 1),
  name           = "Probability",
  column_title   = "Probability of cell types\n within 100 nearest neighbors",
  column_title_gp = grid::gpar(fontsize = 15),
  column_dend_height = unit(1, "cm"),
  row_dend_width = unit(1, "cm"),
  column_names_rot = 45,
  cluster_rows   = TRUE,
  cluster_columns = TRUE,
  show_row_names = FALSE # set to FALSE to remove them entirely
)
```



## Measuring neighborhood complexity 

Discuss what perplexity is and how it relates to entropy
Note how it's calculated and what the value tells us about each cell


```{r}
# merge neighborhood results withe the spe
spe <- hoodscanR::mergeHoodSpe(
  spe, 
  hoods
)

# show that one column per cell type has been added for each cell with the probability
colData(spe) |> 
  head()
```


```{r}
# calculate entropy and perplexity
spe <- hoodscanR::calcMetrics(
  spe, 
  pm_cols = colnames(hoods)
)

# now we can see both entropy and perplexity added
colData(spe) |> 
  head()
```

Since the columns are now in the SPE object, we can use the `ggspavis::plotCoords()` function we have used previously. 
We could use `hoodscanR::plotTissue()` but the plots are the same. 

```{r}
# plot entropy
ggspavis::plotCoords(spe, annotate = "perplexity") +
  scale_color_viridis_c()
```


## Co-localization 

Explain co-localization and what's the plot
Is there anything else we need here besides the correlation plot? 

Note that if we want to customize the plot we would first have to calculate the Pearson correlation matrix, which we could do? 

```{r}
#| fig.height: 5
# plot colocalization
hoodscanR::plotColocal(
  spe, 
  pm_cols = colnames(hoods),
  hm_width = 8,
  hm_height = 8
)
```

TODO: Some discussion about the cell types that are present in the same area
We see overlap in T cell populations, Endothelial with Perivascular-Like, and Myoepithelial with DCIS

## Domain detection 

TODO: Discuss how neighborhood profiles are used to cluster cells into domains
Comparable to Banksy in regards to domain detection, but here we are getting cells with neighborhood profiles that are similar to each other 
that means if there's areas of the slide where two cell types are always together and then other areas where one of those cell types is by itself, that cell type is going to show up in two different clusters. 

Also need to explain how clustering is implemented here and that we need to pick k so it can be useful to spend time figuring out the best k. 

```{r}
# calculate domains
spe <- hoodscanR::clustByHood(
  spe, 
  pm_cols = colnames(hoods),
  k = 8 # need to pick the appropriate k
)
```

```{r}
table(spe$clusters)
```


```{r}
# plot domains
ggspavis::plotCoords(spe, annotate = "clusters", point_size = 0.1) + 
  scale_color_brewer(palette = "Dark2") +
  guides(color = ggplot2::guide_legend(
      override.aes = list(size = 3),
      title = "domains"
  ))
```

For each domain, we can see what cell types belong.


```{r}
#| fig.height: 4
# look at individual cell types in each domain
hoodscanR::plotProbDist(
  spe, 
  pm_cols = colnames(hoods), 
  by_cluster = TRUE, 
  plot_all = TRUE, 
  show_clusters = seq(8), # default only includes 6 clusters
  outlier.size  = 0.1
) + 
  # default is 3 columns, but let's use 4
  facet_wrap(vars(clusters), ncol = 4) +
  # and remove the strip background
  theme(strip.background = element_rect(fill = "transparent"))
```

#TODO: Some text about domains dominated by one cell type (3 and 8 are DCIS and tumor) vs a mix of cell types (4 and 5)

The overall complexity of each domain should also be reflected in the perplexity measure we looked at earlier. 
We can plot the perplexity for all cells in each domain and see if that is true. 

```{r}
coldata_df <- colData(spe) |> 
  as.data.frame()

# using entropy to measure the complexity of domains
ggplot(coldata_df, aes(y = perplexity, x = clusters)) +
  ggforce::geom_sina() +
  theme_bw() +
  labs(
    x = "Domain",
    title = "Perplexity of cells in each domain"
  )
```

## Export 

Finally, we'll save the modified SPE or extract the specific columns from the colData and save as a TSV file? 

```{r}

```



## Session info

As usual, we'll end with `sessionInfo()` to record the package versions we used for this notebook.

```{r}
sessionInfo()
```




