---
title: "Spatial neighborhood analysis"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Use `hoodscanR` to create a neighborhood profile for each cell
- Use the output from `hoodscanR` to quantify the complexity of spatial neighborhoods 
- Identify cell types that are commonly co-localized 
- Use `hoodscanR` for domain detection

## Introduction 

Spatial neighborhood analysis is a way to summarize which cell types are near other cell types and describe the local tissue environment.

In this notebook we'll use the [`hoodscanR`](https://www.bioconductor.org/packages/release/bioc/html/hoodscanR.html) package to look at the following characteristics.

1. Neighborhood profiles, or the mix of cell types among each cells' nearest neighbors.
2. Neighborhood complexity, which defines if a cell sitting in a "pure" environment dominated by one cell type, or a highly mixed one.
3. Co-localization, which tells us which cell types tend to show up in the same neighborhoods across the whole slide.
4. Domain detection, which uses clustering of neighborhood profiles to identify spatial domains that share a similar neighborhood profile.

The general idea behind `hoodscanR` is to look at each cell's k nearest neighbors, convert the distances to those neighbors into probabilities, and then summarize those probabilities by cell type.
The result is a per-cell profile which is a set of probabilities describing how likely that cell type is to be found in the neighborhood of each cell. 

## Setup 

To begin, we'll load the packages we'll use throughout this notebook and set the random seed.
Setting a seed matters here because a few of the calculations in `hoodscanR` involve randomness, and we want the notebook to give the same results each time it is run.

```{r libraries}
suppressPackageStartupMessages({
  # The main class we use for spatial transcriptomic data
  library(SpatialExperiment)
  # functions for plotting
  library(ggplot2)
})

# Set theme for ggplot2 plotting
theme_set(theme_bw())

# Set a seed so the random steps (tau estimation, k-means) are reproducible
set.seed(2026)
```

### Directories and files

Next, we'll define paths to the normalized SPE object that we prepared in the previous notebook. 
This object is stored in `data/brca-xenium/normalized`. 

```{r inputs}
# top level data directory for brca xenium data
data_dir <- file.path("data", "brca-xenium")

# data directory for the normalized data
normalized_dir <- file.path(data_dir, "normalized")

# the normalized SPE object we will read in
normalized_rds <- file.path(normalized_dir, "brca_xenium_normalized.rds")

# the path to a tsv file containing a palette of colors for the cell types present in this dataset
cell_type_palette_file <- file.path("palettes", "brca-xenium-cell-type-palette.tsv")
```

We'll also set up the output paths we need at the end of the notebook.

```{r outputs}
# path to analysis directory
analysis_dir <- file.path("analysis", "brca-xenium")
fs::dir_create(analysis_dir)

# path to output spe file
hoodscanr_results_file <- file.path(analysis_dir, "hoodscanr_spe.rds")
```


Now we'll read in the normalized object. 

```{r read-normalized, live = TRUE}
# read the normalized SpatialExperiment object created in the previous notebook
spe <- readr::read_rds(normalized_rds)
```

<!--TODO: We don't really need this section since we just looked at in the previous notebook, but I thought it would be helpful to have this plot here? I'm okay to take it out or replace it with a table of cell types if that's more helpful.-->
As a reminder, let's look at the cell types that are present in this dataset. 
We'll use the `cell_types_lumped` column that we created in the previous notebook for the plot and for the neighborhood analysis.

Since this is a new notebook, we'll need to set the palette of colors to use for each cell type again. 

```{r celltype-palette}
# read in a custom palette mapping each cell type to a fixed color, so that
# colors stay consistent across every plot in the notebook
cell_type_palette <- readr::read_tsv(cell_type_palette_file, col_types = list(.default = "c")) |> 
  # turn the dataframe into a named list 
  # where the first column is names and the second column is values
  tibble::deframe()

cell_type_palette
```

Setting the cell type column as a factor lets us control the order the cell types appear in legends and plots.
We use the order of the palette we just defined so the two always match.

```{r set-celltype-levels}
# re-order celltypes for plotting by setting them as a factor,
# using the palette order as the factor level order
spe$cell_types_lumped <- factor(
  spe$cell_types_lumped, 
  levels = names(cell_type_palette)
)
```

With the palette and factor levels in place, we can plot the cells in their spatial coordinates, colored by cell type.

```{r plot-celltypes}
#| fig.width: 8

# plot the cells at their spatial coordinates, colored by lumped cell type
ggspavis::plotCoords(spe, annotate = "cell_types_lumped") +
  scale_color_manual(values = cell_type_palette)
```



## Identifying cell neighborhoods with `hoodscanR` 

The core output of `hoodscanR` is a neighborhood profile for every cell.
To build it, the package takes three steps that we'll walk through one at a time:

1. Find the k nearest neighboring cells for each cell (`findNearCells`).
2. Convert the distances to those neighbors into probabilities, so that closer cells contribute more (`scanHoods`).
3. Group those per-neighbor probabilities by cell type, giving each cell a probability of being surrounded by each cell type (`mergeByGroup`).

The end result is a matrix with one row per cell and one column per cell type, where each value is the probability that the cell's local neighborhood contains that cell type.

### Setup the `hoodscanR` object

`hoodscanR` expects the annotation it should use to live in a standardized column called `cell_annotation`, and it expects the spatial coordinates to be in the `spatialCoords` of the object.
Let's first add a new column to the `colData` named `cell_annotation` and fill it with the same contents as `cell_types_lumped`. 

```{r add-annotation-column, live=TRUE}
# reformat the SPE for hoodscanR 
# which expects a column named `cell_annotation` to use for neighborhood identification
spe$cell_annotation <- spe$cell_types_lumped
```

Now let's check to see if the spatial coordinates are indeed present in `spatialCoords`. 

```{r setup-hood, live=TRUE}
spatialCoords(spe)[1:10, ]
```

Great, it looks like our object is all set to run `hoodscanR`.  

### Generate the probability matrix

The first step is to find the nearest neighbors for each cell based on physical distance.
`findNearCells()` does this using an approximate nearest-neighbor search, returning the `k` closest cells for every cell in the dataset.
Note that distance here is the Euclidean distance between the x and y coordinates on the slide. 

The choice of `k` sets the size of the neighborhood we're describing.
<!--TODO: Fill in some more information about choosing k and why we're using the default-->
 
```{r nearest-neighbors, live=TRUE}
# for each cell, find its k nearest neighboring cells
neighbors <- hoodscanR::findNearCells(
  spe, 
  k = 100, # number of nearest neighbors to consider for each cell
  anno_col = "cell_annotation" # default 
)
```

What's inside this object? 
We have `cells` and `distance`.
Both are matrices. 

```{r inspect-structure}
str(neighbors)
```


The `cells` matrix has one row per cell and one column per neighbor (`nearest_cell_1` through `nearest_cell_100`).
Each entry is the cell type of that neighbor.

```{r inspect-cells, live=TRUE}
neighbors$cells[1:10, 1:10]
```

The `distance` matrix has the same shape, but each entry is the physical distance from the cell to that neighbor.

```{r inspect-distance, live=TRUE}
neighbors$distance[1:10, 1:5]
```

<!--TODO: Address the fact that there are missing numbers in the rows (due to cells that are filtered)
And note that the nearest_cell_1, etc corresponds to the 100 nearest cells and not the cell index-->

Next we convert those distances into probabilities with `scanHoods()`.
For each cell, the distances to its neighbors are passed through a softmax transformation, which transforms a vector into a probability distribution where all values sum to 1.
The effect is that nearer cells get more weight than farther ones.

<!--TODO: Add explanation of softmax here and/or in slides-->

The `tau` parameter controls how quickly that weight falls off with distance.
Small values of Tau capture local neighborhoods and large values of Tau capture global neighborhoods. 
We'll use the default value of Tau, which is 1/5 * median(D), where D is the distance matrix. 
This is recommended by the authors to capture local neighborhoods. 

```{r prob-mtx, live=TRUE}
# convert neighbor distances into probabilities
prob_mtx <- hoodscanR::scanHoods(neighbors$distance)
```

In this matrix each row is a cell and each column is one of its 100 neighbors.
The value is the probability weight assigned to that neighbor and every row sums to 1.

```{r inspect-prob-mtx, live=TRUE}
prob_mtx[1:10, 1:10]
```

Right now the columns are individual neighbors, which isn't very interpretable on its own.
`mergeByGroup()` fixes this by combining the neighbor probabilities according to the cell type of each neighbor by adding the probability weights of all neighbors that share a cell type.
It needs both the probability matrix (the weights) and the `cells` matrix (the cell type of each neighbor) to do this.

```{r merge-probabilities, live=TRUE}
# collapse the per-neighbor probabilities into one probability per cell type
hoods <- hoodscanR::mergeByGroup(
  prob_mtx, 
  neighbors$cells
)
```

Now each row is still a cell, but the columns are cell types, and each value is the probability that another cell in the focal cell's neighborhood is that cell type.
These per-cell probabilities are the neighborhood profiles that we will explore for the rest of the notebook.

```{r inspect-hoods}
hoods[1:10, ]
```

A nice way to get a feel for these profiles is to plot them as a heatmap, with cells as rows and cell types as columns.
`hoodscanR` provides `plotHoodMat()` for this, but that function is a thin wrapper around `ComplexHeatmap` and doesn't expose many options, so we'll call `ComplexHeatmap::Heatmap()` directly.
That gives us full control over the appearance and makes it easy to adjust later.
We plot the first 100 cells here just to keep the heatmap legible.

```{r neighborhood-heatmap}
# plot the neighborhood profiles for the first 100 cells

ComplexHeatmap::Heatmap(
  t(hoods[1:100,]), # transpose so cells are columns
  col            = scico::scico(200, palette = "acton", direction = -1),
  rect_gp        = grid::gpar(col = "white", lwd = 1),
  name           = "Probability",
  column_title   = "Probability of cell types\n within 100 nearest neighbors",
  column_title_gp = grid::gpar(fontsize = 15),
  column_dend_height = grid::unit(1, "cm"),
  row_dend_width = unit(1, "cm"),
  column_names_rot = 45,
  cluster_rows   = TRUE,     # group cells with similar profiles together
  cluster_columns = TRUE,    # group cell types that co-occur together
  show_column_names = FALSE     # hide per-cell labels; there are too many to read
)
```

TODO: interpretation of the heatmap and point out blocks of cells with similar neighborhood profiles and which cell types cluster together.


## Measuring neighborhood complexity 

A neighborhood profile can be dominated by a single cell type or a mix of many cell types.
`hoodscanR` quantifies this using perplexity, which is the effective number of cell types in the neighborhood. 
A perplexity near 1 means the neighborhood is essentially one cell type; a higher perplexity means a more mixed environment.

Before we can calculate these, we need to attach the neighborhood profiles back onto our SPE object with `mergeHoodSpe()`, which adds one column per cell type to the `colData`.

```{r merge-hood-spe, live=TRUE}
# merge the neighborhood profiles back into the spe object
spe <- hoodscanR::mergeHoodSpe(
  spe, 
  hoods
)

# one column per cell type has been added
# holding that probability that cell type is in the neighborhood of the specified cell in that row
colData(spe) |> 
  head()
```

Now we can calculate entropy and perplexity from those probability columns.

```{r calc-metrics}
# calculate entropy and perplexity from the neighborhood probability columns
spe <- hoodscanR::calcMetrics(
  spe, 
  pm_cols = colnames(hoods)
)

# now we can see both entropy and perplexity added to the colData
colData(spe) |> 
  head()
```

Since the columns are now in the SPE object, we can use the `ggspavis::plotCoords()` function we have used previously. 
We could use `hoodscanR::plotTissue()` but the plots are the same. 

```{r plot-perplexity}
# plot perplexity across the tissue to see where neighborhoods are more mixed
ggspavis::plotCoords(spe, annotate = "perplexity") +
  scale_color_viridis_c()
```

TODO: interpretation of the perplexity map and note where on the slide neighborhoods are more homogeneous vs. more mixed.


## Co-localization 

We can also use the neighborhood profiles to look at which cell types tend to be together, or co-localized.
`plotColocal()` answers this by first correlating the per-cell neighborhood probabilities across every pair of cell types and then plotting that correlation in a heatmap.

```{r plot-colocal}
#| fig.height: 5

# correlate neighborhood profiles between cell types to find co-localized pairs
hoodscanR::plotColocal(
  spe, 
  pm_cols = colnames(hoods),
  hm_width = 8,
  hm_height = 8
)
```

TODO: Some discussion about the cell types that are present in the same area
We see overlap in T cell populations, Endothelial with Perivascular-Like, and Myoepithelial with DCIS

## Domain detection 

The neighborhood profiles can also be used to partition the tissue into domains, or groups of cells with similar neighborhood profiles.
`clustByHood()` does this by running k-means clustering on the per-cell neighborhood profiles, so cells are grouped by the company they keep rather than by their own expression.

This is conceptually related to domain-detection tools like `Banksy`, but the emphasis is different. 
With `Banksy` we cluster cells based on both expression and local neighborhood.
Here we are clustering cells by their neighborhood composition and do not consider gene expression.

As with `k` in the nearest-neighbor step, the number of clusters `k` here is a choice we have to make, and it strongly shapes the result.
It's worth spending time exploring a few values of `k` to find one that gives interpretable, stable domains.

TODO: Do we want to keep this in instruction or move to exercise notebook? 
Note that this does take 2-3 minutes to run. 

```{r}
# candidate numbers of domains to try
k_range <- 2:12

# construct a dataframe with value of k and within-cluster sumo f squares
elbow_df <- tibble::tibble(k = k_range) |>
  dplyr::mutate(
    # fit k-means on the neighbourhood probability matrix for each k...
    fit  = purrr::map(k, \(x) kmeans(hoods, centers = x, nstart = 25, iter.max = 50)),
    # and pull out the total within-cluster sum of squares
    wcss = purrr::map_dbl(fit, "tot.withinss")
  )

# make the elbow plot
ggplot(elbow_df, aes(x = k, y = wcss)) +
  geom_line(colour = "grey60") +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = k_range) +
  labs(
    title    = "Elbow plot for domain identification",
    x = "Number of domains (k)",
    y = "Total within-cluster sum of squares (WCSS)"
  )
```


```{r clust-domains, live=TRUE}
# cluster cells into spatial domains based on their neighborhood profiles
spe <- hoodscanR::clustByHood(
  spe, 
  pm_cols = colnames(hoods),
  k = 8 # number of domains, chosen from the elbow plot above
)
```

We can see how many cells fall into each domain.

```{r domain-table, live=TRUE}
table(spe$clusters)
```


For each domain, we can look at which cell types it is made of by plotting the distribution of neighborhood probabilities within each domain.

```{r plot-probdist}
#| fig.height: 8
#| fig.width: 12

# for each domain, show the distribution of neighborhood probabilities per cell type
hoodscanR::plotProbDist(
  spe, 
  pm_cols = colnames(hoods), 
  by_cluster = TRUE, 
  plot_all = TRUE, 
  fill = hoods,
  show_clusters = seq(8), # show all 8 domains (default only shows 6)
  outlier.size  = 0.1     # shrink the boxplot outlier points
) + 
  # default is 3 columns, but let's use 4
  facet_wrap(vars(clusters), 
             ncol = 4,
             scales = "free_x") +
  # and remove the strip background
  theme(strip.background = element_rect(fill = "transparent")) +
  # add some colors to cell types
  scale_fill_manual(values = cell_type_palette)  +
  # change the default legend title
  labs(fill = "Cell types")
```

TODO: Some text about domains dominated by one cell type (3 and 8 are DCIS and tumor) vs a mix of cell types (4 and 5)

Now we can map the domains back onto the tissue.

```{r plot-domains}
# plot the domains across the tissue
ggspavis::plotCoords(spe, annotate = "clusters", point_size = 0.1) + 
  scale_color_brewer(palette = "Dark2") +
  guides(color = ggplot2::guide_legend(
      override.aes = list(size = 3), # enlarge the legend keys so colors are readable
      title = "domains"
  ))
```

TODO: interpretation of the domain map

The overall complexity of cell types present within a domain should also be reflected in the perplexity measure we looked at earlier. 
We can plot the perplexity for all cells in each domain and see if that is true. 

```{r plot-perplexity-domains}
# pull the colData into a data frame for plotting with ggplot2
coldata_df <- colData(spe) |> 
  as.data.frame()

# using perplexity to measure the complexity of each domain
ggplot(coldata_df, aes(y = perplexity, x = clusters)) +
  ggforce::geom_sina(size = 0.5) +
  theme_bw() +
  labs(
    x = "Domain",
    title = "Perplexity of cells in each domain"
  )
```

TODO: text around confirming that the single-cell-type domains have low perplexity and the mixed domains have high perplexity.


## Export 

Finally, we'll save our results.
We'll save the modified SPE objects with the results from `hoodscanR` to a new object. 

```{r export}
readr::write_rds(spe, hoodscanr_results_file)
```

## Session info

As usual, we'll end with `sessionInfo()` to record the package versions we used for this notebook.

```{r session-info}
sessionInfo()
```
