---
title: "Spatial Transcriptomics Exercise: Neighborhood analysis with Visium HD data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

In this exercise notebook, we'll be looking at a 10x Visium HD dataset from a colorectal cancer (CRC) sample.

---

In this exercise notebook you will,

- Part A: Read in and annotate a Visium HD dataset
- Part B: Perform neighborhood analysis with `hoodscanR`
- Part C: Perform domain detection with `hoodscanR`
- Part D: Explore parameters for `hoodscanR`

## Part A: Read in and annotate a Visium HD dataset

First, let's load the libraries we will use throughout this notebook.

```{r load-libraries}
library(SpatialExperiment)
library(ggplot2)
library(patchwork)
```

```{r theme-set}
# Set theme for ggplot2 plotting
theme_set(theme_bw())

# Set a seed so the random steps (tau estimation, k-means) are reproducible
set.seed(2026)
```


Next, we'll define the paths to the input files we will use in this notebook.

```{r input-paths}
# define path to data directory
data_dir <- file.path("data", "crc-hd")

# paths to input SPE and reference SCE which will be used for annotation
normalized_spe_file <- file.path(data_dir, "crc_hd_normalized_spe.rds")
ref_sce_file <- file.path(data_dir, "crc_single_cell_normalized_sce.rds")
```

We'll also define the path where we will save the final annotated `SpatialExperiment` object at the end of this notebook.

```{r output-paths}
# define path to results directory and create it if it doesn't exist
results_dir <- file.path("results")
fs::dir_create(results_dir)

# path to save the final SPE object with annotations and neighborhood results
hood_spe_file <- file.path(results_dir, "crc_hd_hood_spe.rds")
```

### About Visium HD outputs

Unlike regular Visium, where transcripts are captured in 55 µm spots, Visium HD slides contain a continuous grid of 2 µm barcoded squares.
Space Ranger provides several outputs from a Visium HD run:

- The raw 2 µm binned output
- The same data aggregated into 8 µm and 16 µm bins, which are commonly used for analysis since individual 2 µm bins are very sparse
- Segmented outputs, where nuclei are identified in the tissue image and nearby 2 µm bins are grouped together into what is _estimated_ to be a single cell

In this notebook we will be working with the **segmented outputs**.
The segmented outputs folder contains a filtered feature-cell matrix with counts summed across each segmented cell, the spatial coordinates for each cell, and the cell segmentation geometries derived from the tissue image.

We have already pre-processed this data for you, since filtering and normalization for Visium HD follow the same steps we used for regular Visium in the previous notebook.
This means the object you are reading in below has already been filtered to remove low-quality cells and contains normalized counts in the `logcounts` assay.

Read in both the pre-processed `SpatialExperiment` (SPE) object and the reference `SingleCellExperiment` (SCE) object.

```{r read-input-data}
# read in SPE and reference SCE
spe <- readr::read_rds(normalized_spe_file)
ref_sce <- readr::read_rds(ref_sce_file)
```

### Annotate with SingleR

The main difference here is that with segmented outputs we get information for what is estimated to be a single cell and not a spot.
We can therefore annotate these cells similar to how we would with single-cell RNA-seq.

Here we will use `SingleR`, a reference-based annotation method.
`SingleR` compares the expression profile of each cell in our dataset to a reference dataset that already has cell type labels, and assigns each of our cells the label it most closely resembles.
For our reference, we will use a previously annotated single-cell RNA-seq dataset from CRC samples, where cell type labels are stored in the `Level1` column of the `colData`.

If you would like a more detailed discussion of cell type annotation methods, including `SingleR`, we have a whole notebook on annotation in our [Introduction to single-cell RNA-seq training materials](https://github.com/AlexsLemonade/training-modules) that you can reference.

Let's first explore our reference object and the cell types it contains.
Use the `unique()` function to look at the cell type labels found in the `Level1` column of `ref_sce`.

Note that this reference has been downsampled to 1000 cells per cell type to keep the object size manageable; the original reference had > 250K cells.

```{r explore-ref, solution=TRUE}
# print the unique cell type labels in the Level1 column of ref_sce
unique(ref_sce$Level1)
```

Now let's go ahead and annotate.

### What does `SingleR` do?
<!--Note that this is mostly copied over from the cell type annotation instruction notebook-->

As mentioned earlier, `SingleR` builds a model from a set of training data, and then uses that model to classify cells (or groups of cells) in new datasets.

`SingleR` works by first identifying a set of marker genes that can be used to differentiate among the cell types in the reference dataset.
It does this by performing pairwise comparisons among all of the cell types, and retaining the top set of genes differentiating each pair.
The idea is that this set of genes will be the most informative for differentiating cell types.

Then, for each cell, `SingleR` calculates the Spearman correlation between expression of that cell and each cell type (using only the genes chosen earlier).
Notably, this is a non-parametric correlation, so the scaling and normalization that we apply (or don't) should not matter!
Note that if you used a single-cell technology that produces full-length transcripts (i.e., SMART-seq), you will probably want to convert your counts to Transcripts per Million (TPM), to allow more consistent ranking among transcripts of different lengths.

The reference cell type with the highest correlation is then chosen as the cell type assignment for that cell.
If there are multiple cell types with high scores, an optional fine-tuning step repeats the process using only the most relevant genes for those cell types.

### Running `SingleR`

Now we can run `SingleR` using the convenience function `SingleR::SingleR()`, which performs the marker gene selection (training) and classification in a single step.
We will save the output to an object named `singler_results`.

We will need to supply the following arguments:

- `test`: our SPE object, `spe`
- `ref`: the reference object, `ref_sce`
- `labels`: the vector of reference labels to use, here the `Level1` column of `ref_sce`
- `aggr.ref = TRUE`: this pseudobulks cells of the same cell type in the reference, which speeds things up considerably


```{r run-singler}
# annotate cells in the SPE using the Level1 labels from the reference SCE
singler_results <- SingleR::SingleR(
  test = spe,
  ref = ref_sce,
  labels = ref_sce$Level1,
  aggr.ref = TRUE # pseudobulk cells of the same cell type for speed
)
```

Let's explore the results from `SingleR`.
The output is a `DataFrame` with one row per cell in our SPE and the following components:

- `scores`: a matrix of correlation-based scores for each cell (rows) and each reference cell type (columns); higher scores indicate a closer match
- `labels`: the cell type with the highest score for each cell
- `delta.next`: the difference in score between the assigned label and the next-best label; small values indicate the assignment was ambiguous
- `pruned.labels`: the same as `labels`, but low-quality or ambiguous assignments have been replaced with `NA`

```{r singler-results}
singler_results
```

We can also pull out individual components of the results object for use in our analysis.
Here we will save the pruned labels (where low-quality assignments have been given an `NA` label), storing them back in our SPE object (specifically to a new column of the `colData` table).

Store the `pruned.labels` in a new `colData` column named `cell_annotation`, since ultimately that's the column name expected by `hoodscanR`.

```{r add-annotations}
# save the pruned labels to a new colData column named cell_annotation
spe$cell_annotation <- singler_results$pruned.labels
```

Now count how many of each cell type we see in our spatial data.
You can do this by using the `table()` function on the `cell_annotation` column.

```{r count-celltypes, solution=TRUE}
# count the number of cells assigned to each cell type
table(spe$cell_annotation)
```

What are the dominant cell types that you see?

Next we can visualize the cell types on a coordinate plot.
First, we will define a distinct color palette with one color per cell type, so cell types are easy to tell apart in this plot and in later plots.

```{r celltype-palette}
# get the cell type labels, dropping any NA values
celltypes <- na.omit(unique(spe$cell_annotation))

# create a named vector of distinct colors, one per cell type
cell_type_palette <- setNames(
  palette.colors(length(celltypes), palette = "Classic Tableau"),
  celltypes
)
```

Now use the `ggspavis::plotCoords()` function to plot the cells, coloring by the `cell_annotation` column (hint: use the `annotate` argument).
Add `scale_color_manual()` to apply the `cell_type_palette` we just created; you can use the `na.value` argument to give any unannotated cells a neutral color like `"grey80"`.

```{r plot-celltypes, solution=TRUE}
# plot cells on the tissue colored by cell type using our custom palette
ggspavis::plotCoords(spe, annotate = "cell_annotation") +
  scale_color_manual(values = cell_type_palette, na.value = "gray90")
```

<!--TODO: Add interpretation text for the cell type plot-->

## Part B: Perform neighborhood analysis with `hoodscanR`

Now we'll perform neighborhood analysis with `hoodscanR`.
Similar to the previous notebook, where we removed unlabeled spots before neighborhood analysis, we first need to remove any cells with an `NA` annotation.

Filter `spe` to keep only cells where `cell_annotation` is not `NA`.
Hint: remember that cells are columns of the SPE object, and you can use `!is.na()` to identify the cells to keep.

```{r remove-na, solution=TRUE}
# remove cells that did not receive a cell type annotation
spe <- spe[, !is.na(spe$cell_annotation)]
```

Neighborhood analysis with `hoodscanR` happens in three steps, each with its own function:

1. `hoodscanR::findNearCells()`: for each cell, find its k nearest neighboring cells
2. `hoodscanR::scanHoods()`: convert the distances to those neighbors into probabilities
3. `hoodscanR::mergeByGroup()`: collapse the per-neighbor probabilities into one probability per cell type

First, use `hoodscanR::findNearCells()` to find the 100 nearest neighbors for each cell, and save the output to an object named `neighbors`.
You will need to provide the `spe` object and set `k = 100`.
Note that the `anno_col` argument tells the function which `colData` column holds the cell type labels. 
Its default is `"cell_annotation"`, which is why we used that column name earlier.

```{r nearest-neighbors, solution=TRUE}
# for each cell, find its k nearest neighboring cells
neighbors <- hoodscanR::findNearCells(
  spe,
  k = 100, # number of nearest neighbors to consider for each cell
  anno_col = "cell_annotation" # default
)
```

Next, use `hoodscanR::scanHoods()` to convert the neighbor distances into probabilities, saving the output as `prob_mtx`.

```{r prob-mtx, solution=TRUE}
# convert neighbor distances into probabilities
prob_mtx <- hoodscanR::scanHoods(neighbors$distance)
```

Finally, use `hoodscanR::mergeByGroup()` to collapse the per-neighbor probabilities into one probability per cell type, saving the output as `hoods`.
This function takes two arguments: the probability matrix and the cell type annotations of each neighbor, stored in the `cells` element of the `neighbors` object.

```{r merge-probabilities, solution=TRUE}
# collapse the per-neighbor probabilities into one probability per cell type
hoods <- hoodscanR::mergeByGroup(
  prob_mtx,
  neighbors$cells
)
```

Let's visualize the neighborhood profiles we just created for a subset of cells using a heatmap.

```{r neighborhood-heatmap}
# plot the neighborhood profiles for the first 500 cells
ComplexHeatmap::Heatmap(
  t(hoods[1:500, ]), # transpose so cells are columns
  col            = scico::scico(200, palette = "acton", direction = -1),
  name           = "Probability",
  column_title   = "Probability of cell types\n within 100 nearest neighbors",
  column_title_gp = grid::gpar(fontsize = 15),
  column_dend_height = grid::unit(1, "cm"),
  row_dend_width = unit(1, "cm"),
  column_names_rot = 45,
  cluster_rows   = TRUE,     # group cells with similar profiles together
  cluster_columns = TRUE,    # group cell types that co-occur together
  show_column_names = FALSE  # hide per-cell labels; there are too many to read
)
```

<!--TODO: Add interpretation text for the neighborhood heatmap-->

### Measuring neighborhood complexity

Before we can calculate any metrics, we need to add the neighborhood probabilities back into our SPE object.
Use `hoodscanR::mergeHoodSpe()` to merge the `hoods` matrix into `spe`, saving the result back to `spe`.
This adds one new `colData` column per cell type, holding the neighborhood probabilities.

```{r merge-hood-spe}
# first we need to make sure that the rows in hoods have the same column names as in the original object
# this seems to be something specific to Visium data where generating the neighbor profiles doesn't keep the original cell id? 
rownames(hoods) <- colnames(spe)

# merge the neighborhood profiles back into the spe object
spe <- hoodscanR::mergeHoodSpe(
  spe,
  hoods
)
```

Now we can calculate entropy and perplexity from those probability columns.

Use `hoodscanR::calcMetrics()` to add these metrics to `spe`, saving the result back to `spe`.
Hint: the `pm_cols` argument tells the function which columns hold the probabilities; you can use `colnames(hoods)` to get those column names.

```{r calc-metrics, solution=TRUE}
# calculate entropy and perplexity from the neighborhood probability columns
spe <- hoodscanR::calcMetrics(
  spe,
  pm_cols = colnames(hoods)
)
```

We can now plot either entropy or perplexity across the tissue.
These two metrics measure the same underlying quantity but on different scales, so we expect the plots to show very similar patterns.

Make a coordinate plots with `ggspavis::plotCoords()`, using either `"entropy"` or `"perplexity"` for the `annotate` argument. 
The solution will show both plots.

```{r plot-entropy, solution=TRUE}
# plot entropy across the tissue
entropy_plot <- ggspavis::plotCoords(spe, annotate = "entropy") +
  scale_color_viridis_c()

# plot perplexity across the tissue
perplexity_plot <- ggspavis::plotCoords(spe, annotate = "perplexity") +
  scale_color_viridis_c()

# combine the plots side by side with patchwork
entropy_plot + perplexity_plot
```

Where in the tissue do you see the most mixed neighborhoods?

### Co-localization

We can also use the neighborhood profiles to look at which cell types tend to be found together, or co-localized.
Use `hoodscanR::plotColocal()` to create this heatmap.
You will need to provide the `spe` object and the `pm_cols` argument, just like you did for `calcMetrics()`.
We recommend using `hm_width = 8`, and `hm_height = 8` to keep the heatmap a readable size.

```{r plot-colocal, solution=TRUE}
#| fig.height: 5

# correlate neighborhood profiles between cell types to find co-localized pairs
hoodscanR::plotColocal(
  spe,
  pm_cols = colnames(hoods),
  hm_width = 8,
  hm_height = 8
)
```

Which cell types appear to be co-localized?

<!--TODO: Add interpretation text for the co-localization heatmap-->

## Part C: Perform domain detection with `hoodscanR`

`hoodscanR` identifies spatial domains by clustering cells with k-means based on their neighborhood profiles.
K-means requires us to choose the number of clusters (domains), so before clustering we will make an elbow plot to help pick a reasonable value of k.

For each candidate value of k, we fit k-means to the neighborhood probability matrix and record the total within-cluster sum of squares (WCSS), a measure of how tightly cells cluster together.

```{r wcss-calculation}
# candidate numbers of domains to try
k_range <- 2:12

# construct a data frame with each value of k and its within-cluster sum of squares
elbow_df <- k_range |>
  purrr::map(\(k){
    fit <- kmeans(hoods, centers = k, nstart = 25, iter.max = 50)
    data.frame(k = k, wcss = fit$tot.withinss)
  }) |>
  purrr::list_rbind()
```


```{r elbow-plot}
# make the elbow plot
ggplot(elbow_df, aes(x = k, y = wcss)) +
  geom_line(colour = "grey60") +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = k_range) +
  labs(
    title = "Elbow plot for domain identification",
    x = "Number of domains (k)",
    y = "Total within-cluster sum of squares (WCSS)"
  )
```

Use the elbow plot to pick the value of k you think is best.
WCSS always decreases as k increases, so we are not looking for the smallest value.
Instead, look for the "elbow" of the curve or the point where adding more domains stops giving a substantial drop in WCSS and the curve begins to flatten out.
Choosing a k near the elbow gives us domains that capture real structure in the data without splitting the tissue into more domains than we can meaningfully interpret.
There is often no single right answer, so it's okay if your choice differs slightly from ours!

Now cluster the cells into spatial domains using `hoodscanR::clustByHood()`, saving the result back to `spe`.
You will need to provide the `spe` object, the `pm_cols` argument (same as before), and your chosen value of `k`.
This will add a new `colData` column named `clusters` holding the domain assignment for each cell.

```{r cluster-domains, solution=TRUE}
# cluster cells into spatial domains based on their neighborhood profiles
spe <- hoodscanR::clustByHood(
  spe,
  pm_cols = colnames(hoods),
  k = 8 # number of domains, chosen from the elbow plot above
)
```

Use `table()` to see how many cells fall into each domain.

```{r domain-table, solution=TRUE}
# count the number of cells assigned to each domain
table(spe$clusters)
```

Let's also see where the domains fall in the tissue.
Use `hoodscanR::plotTissue()` to plot the cells colored by domain.

```{r plot-domains, solution=TRUE}
# plot the spatial domains on the tissue
hoodscanR::plotTissue(spe, color = clusters, size = 0.1) +
  scale_color_brewer(palette = "Paired") +
  # make the legend points larger and easier to see
  guides(color = ggplot2::guide_legend(override.aes = list(size = 2)))
```

For each domain, we can look at which cell types it is made of by plotting the distribution of neighborhood probabilities within each domain.

```{r plot-probdist}
#| fig.height: 8
#| fig.width: 12

# for each domain, show the distribution of neighborhood probabilities per cell type
hoodscanR::plotProbDist(
  spe,
  pm_cols = colnames(hoods),
  by_cluster = TRUE,
  plot_all = TRUE,
  fill = hoods,
  show_clusters = seq(8), # show all domains; change this if you chose a different k
  outlier.size  = 0.1     # shrink the boxplot outlier points
) +
  # default is 3 columns, but let's use 4
  facet_wrap(vars(clusters),
             ncol = 4,
             scales = "free_x") +
  # and remove the strip background
  theme(strip.background = element_rect(fill = "transparent")) +
  # add some colors to cell types
  scale_fill_manual(values = cell_type_palette) +
  # change the default legend title
  labs(fill = "Cell types")
```

Which cell types define each domain?

<!--TODO: Add interpretation text for the probability distribution plot-->

## Part D: Explore parameters for `hoodscanR`

So far we have used the default parameters for `hoodscanR::scanHoods()`.
One important parameter to consider is `tau`, which controls how quickly a neighbor's contribution to the probabilities decays with distance.
By default, `scanHoods()` calculates `tau` as the median squared neighbor distance divided by 5.

- With a **smaller `tau`**, only the closest neighbors contribute meaningfully, so the profiles capture very **local** neighborhoods.
- With a **larger `tau`**, more distant neighbors still contribute, so the profiles capture broader, more **global** neighborhoods.

Using a larger `tau` can therefore change the neighborhood profiles we create and, in turn, the neighborhood complexity (perplexity) and the spatial domains we detect.
<!--TODO: Add some text about how this could change biological interpretation of results-->

In this section, we will repeat the `hoodscanR` workflow from Part B, but this time with a larger, more "global" `tau`, and compare the results to our original "local" run.

One nice shortcut: finding the nearest neighbors does not depend on `tau` at all, since the cells and their distances haven't changed.
That means we can reuse the `neighbors` object from Part B and only need to repeat the steps that come after `findNearCells()`.

We also don't want to overwrite the results already stored in `spe`, since we'll need those for our comparisons.
Start by making a copy of `spe` named `global_spe`. 
We will store all of the global `tau` results in this copy while `spe` keeps the local results.

```{r copy-spe, solution=TRUE}
# make a copy of the SPE to hold the global tau results,
# so we don't overwrite the local results stored in spe
global_spe <- spe
```

Next, calculate the global value of `tau` and save it as `global_tau`.
Remember that the default `tau` is the median squared neighbor distance divided by 5.
Here, we'll calculate the median squared neighbor distance but *don't* divide it by anything which makes `tau` five times larger than the default, capturing more global neighborhoods.

```{r global-tau}
# calculate a global tau: the median squared neighbor distance,
# which is 5x larger than the default of median(distance^2) / 5
global_tau <- median(neighbors$distance^2)
```

Now repeat the neighborhood analysis steps from Part B, this time supplying your new `tau`:

Use `hoodscanR::scanHoods()` to convert the distances in `neighbors$distance` into probabilities, this time setting the `tau` argument to `global_tau`.
Save the output as `global_prob_mtx`.

```{r global-prob-mtx, solution=TRUE}
# convert neighbor distances into probabilities using the global tau
global_prob_mtx <- hoodscanR::scanHoods(
  neighbors$distance,
  tau = global_tau
)
```

Now use `hoodscanR::mergeByGroup()` with `global_prob_mtx` and `neighbors$cells` to collapse the probabilities to one per cell type.
Save the output as `global_hoods`.

```{r global-merge-probabilities, solution=TRUE}
# collapse the per-neighbor probabilities into one probability per cell type
global_hoods <- hoodscanR::mergeByGroup(
  global_prob_mtx,
  neighbors$cells
)
```

Just like in Part B, we now need to merge the new neighborhood probabilities into our SPE object.

```{r global-merge-spe}
# reset the column names to make sure they match
rownames(global_hoods) <- colnames(global_spe)

# merge the global neighborhood profiles into the copied spe object
global_spe <- hoodscanR::mergeHoodSpe(
  global_spe,
  global_hoods
)
```

### Compare domain assignments

Let's see how the choice of `tau` affects the spatial domains we detect.
Repeat the domain detection from Part C by running `hoodscanR::clustByHood()` on `global_spe`, saving the result back to `global_spe`.
Be sure to use `colnames(global_hoods)` for the `pm_cols` argument and the *same* value of `k` you chose from the elbow plot in Part C, so the two clustering results are comparable.

```{r global-cluster-domains, solution=TRUE}
# cluster cells into spatial domains using the global neighborhood profiles
global_spe <- hoodscanR::clustByHood(
  global_spe,
  pm_cols = colnames(global_hoods),
  k = 8 # use the same k you chose in Part C
)
```

Similar to when we compared clustering results in previous notebooks, we can make a heatmap comparing the cells assigned to each domain in the local run versus the global run.
If `tau` had no effect, we would expect each local domain to line up neatly with a single global domain.

First, we'll pull out the domain assignments from each run and add a prefix (e.g., `"local_"` and `"global_"`) so we can tell them apart in the heatmap. 
Then we'll use `table()` to build a matrix counting how many cells fall into each combination of local and global domain, and plot that matrix with `pheatmap::pheatmap()`.

```{r cluster-heatmap}
# pull out the domain assignments and add prefixes to the labels
local_clusters <- glue::glue("local_{spe$clusters}")
global_clusters <- glue::glue("global_{global_spe$clusters}")

# count cells in each combination of local and global domain
clusters_table <- table(local_clusters, global_clusters)

# normalize by the number of cells in each local tau cluster, which appear on rows
clusters_table <- prop.table(clusters_table, 1)

# plot the comparison as a heatmap
pheatmap::pheatmap(clusters_table)
```

Finally, we can quantify how similar the two sets of domain assignments are by calculating the Adjusted Rand Index (ARI).
The ARI ranges from 0 to 1, where 1 means the two clusterings are identical and values near 0 mean the assignments agree no more than expected by chance.

<!--TODO: Add some more text about ARI interpretation.
Add a link to the clustering exercise. 
Should we also show the diag here? 
-->

Use `bluster::pairwiseRand()` with `mode = "index"` to calculate the ARI between the local and global domain assignments.

```{r calculate-ari, solution=TRUE}
# calculate the ARI between local and global domain assignments
bluster::pairwiseRand(
  spe$clusters,
  global_spe$clusters,
  mode = "index"
)
```

Based on the heatmap and the ARI, how much does the choice of `tau` ultimately affect the resulting neighborhood profiles and domains?

<!--TODO: Add interpretation text for the domain comparison and ARI-->

## Save results

Before we finish, save the SPE object from our original analysis, which contains the cell type annotations, neighborhood probabilities, metrics, and domain assignments.
Use `readr::write_rds()` to save `spe` to the `hood_spe_file` path we defined at the top of the notebook.

```{r save-results, solution=TRUE}
# save the annotated SPE with neighborhood results
readr::write_rds(spe, hood_spe_file)
```

## Session Info

```{r session-info}
sessionInfo()
```
