---
title: "Spatial Transcriptomics Exercise: Clustering spatial data"
author: Data Lab for ALSF
date: 2026
output:
  html_notebook:
    toc: true
    toc_float: true
---

In this exercise notebook, we'll perform spatial clustering with `Banksy` on a 10x Genomics [colorectal cancer dataset](https://www.10xgenomics.com/datasets/human-colorectal-cancer-whole-transcriptome-analysis-1-standard-1-2-0), obtained from fresh frozen invasive adenocarcinoma of the large intestine tissue.
This dataset was processed with `Space Ranger 1.2.0` and used Visium v1 technology. 

As is best practice when clustering, we also took some time to identify optimal clustering parameters (in expression-only space) for this dataset, as you can see [in this notebook](https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/crc-v1/evaluate-cluster-parameters.Rmd).
Using the Leiden algorithm with Jaccard weighting, we explored performance of different values of the nearest neighbors and resolution parameters.
We assessed performance using three metrics which you can read more about at the provided links:

1. [Silhouette width](https://bioconductor.org/books/3.22/OSCA.advanced/clustering-redux.html#silhouette-width)
2. [Neighborhood purity](https://bioconductor.org/books/3.22/OSCA.advanced/clustering-redux.html#cluster-purity)
3. [Cluster stability](https://bioconductor.org/books/3.22/OSCA.advanced/clustering-redux.html#cluster-bootstrapping), using the [Adjusted Rand Index metric](https://bioconductor.org/books/3.22/OSCA.advanced/clustering-redux.html#adjusted-rand-index)

Note that there is an exercise notebook `exercise_05-cluster_evaluation.Rmd` in the `scRNA-seq-advanced` module which you can use to explore calculating and using these metrics more in depth, although this notebook will also include some cluster stability calculations.

In this exercise notebook, we use these previously identified parameters to perform non-spatial and `Banksy` clustering on this colorectal dataset and compare resulting clusters.
We'll then explore how varying `Banksy`'s `lambda` parameter affects clustering results.
As `lambda` increases, spatial information is increasingly weighted in cluster calculations.
This means that clusters produced from lower `lambda` values tend to reflect cell-type--related heterogeneity, and clusters produced from higher `lambda` values will increasingly consider spatial information more strongly.
Along the way, we'll learn methods for quantitatively comparing two sets of clusters and measuring cluster stability which helps to evaluate different clustering results.
The purpose of these explorations is to help you choose a value for the `lambda` parameter for your own analyses that is grounded in your specific scientific goals.

---

In this exercise notebook you will,

- Part A: Prepare data for clustering
- Part B: Perform clustering
- Part C: Compare clustering results
- Part D: Explore how the `lambda` parameter affects clustering
- Part E: Calculate cluster stability across `lambda` values

---

## Part A: Prepare data for clustering


### Setup and import

Let's begin by loading libraries we'll need for this notebook.


```{r load libraries}
suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(ggplot2)
  library(patchwork)
})

# set theme for ggplot2 plots
theme_set(theme_bw())
```

We also need to set the random seed for reproducibility, since clustering includes the use of random numbers.

```{r set seed}
set.seed(2026)
```


Next, we'll define input paths for the data to read in. 
We have prepared this dataset for use in this notebook by filtering and normalizing it.
For details, see the [`prepare-crc-v1-data.Rmd` notebook](https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/crc-v1/prepare-crc-v1-data.Rmd).

First, we need that prepared RDS file, which contains an SPE object.

```{r data paths}
data_dir <- file.path("data/crc-v1/normalized")
crc_rds <- file.path(data_dir, "crc_normalized_spe.rds")
```

Next, we'll define a path for an output RDS file which will contain the SPE object along with the clusters we'll calculate here.

```{r output paths}
crc_output <- file.path(data_dir, "crc_clustered_spe.rds")
```


Read and print in the SPE object in the chunk below.
Save the object to a variable named `spe`.

```{r import spe, solution = TRUE}
# read in the SPE

# print the SPE

```

How many spots are present? 
How many genes?

Next, plot the H&E with `ggspavis::plotVisium()` to see the associated tissue image.
You may wish to use the argument `spots = FALSE`.

```{r plot he, solution = TRUE}
# plot the H&E

```
The tissue shows classic characteristics of adenocarcinoma: disorganized glands of tumor epithelium surrounded by dense fibrous tissue known as desmoplastic stroma.

### Prepare for clustering

To begin, we'll perform expression-only (non-spatial) clustering as well as clustering with `Banksy`.
For both, we'll use the clustering parameters [that we identified in this notebook](https://github.com/AlexsLemonade/training-modules/blob/master/spatial/setup/crc-v1/evaluate-cluster-parameters.Rmd):

* The Leiden algorithm, a newer algorithm that has emerged as a popular alternative to Louvain.
It is a modified version of the Louvain algorithm and includes additional refinements to increase connectedness among cluster communities.
The Leiden algorithm optimizes clusters using what `bluster` (which uses [`igraph`](https://igraph.org/) under the hood) refers to as an "objective function."
We'll specify the "modularity" objective function, which tends to yield clusters somewhat more similar to Louvain compared to other objective functions.
* Jaccard weighting 
* Parameter values of 25 nearest neighbors, and a resolution of 0.75

Let's define those parameters first using `bluster::NNGraphParam()`:

```{r define cluster params}
# define clustering parameters
cluster_params <- bluster::NNGraphParam(
   k = 25,
   cluster.fun = "leiden",
   type = "jaccard", 
   cluster.args = list(
     objective_function = "modularity", 
     resolution = 0.75
  )
)
```


To ensure a fair comparison between non-spatial and `Banksy` clusters, we'll subset the SPE right off the bat to contain only highly variable genes. 
We'll then proceed to use that subsetted SPE throughout the notebook.

In the next chunk, create a vector of highly variable genes in two steps: first the gene variance with `scran::modelGeneVar()`, and then use `scran::getTopHVGs()` to get the top 2000 variable genes.
Save the vector of genes to a variable called `hv_genes`.

```{r get hvgs, solution = TRUE}
# model gene variance

# select the top 2000 HVGs

```

Next, create a new SPE object called `hvg_spe` which only contains these genes.

```{r subset spe, solution = TRUE}
# subset the SPE to contain only HVGs

```

## Part B: Perform clustering

### Calculate non-spatial clusters

Next, we'll perform non-spatial clustering, beginning by calculating the PCA matrix.
Even though we have already subsetted the object to highly-variable genes, we should still provide the 
`subset_row` argument when running `scater::runPCA()`.
This will ensure this full set of genes is used in the same order that `scran::modelGeneVar()` determined.
If we did not include this argument, `scater::runPCA()` would fall back to using the its default of the top 500 genes based on raw variance.  

In the chunk below, calculate the PCA matrix.

```{r calculate PCA, solution = TRUE}
# run PCA

```

Now, calculate non-spatial clusters using `scran::clusterCells()`.
You'll want to use the `cluster_params` object we defined earlier as part of your call.
Save the resulting cluster assignments to a `colData` column called `nonspatial_cluster`.

```{r calculate nonspatial clusters, solution = TRUE}
# Calculate graph-based clusters without spatial information

```

### Calculate `Banksy` clusters

Next, we'll calculate `Banksy` clusters.
For this, specify to turn the Azimuthal Gabor Filter (`AGF`) on.
Recall, the `AGF` matrix is the spatially-informed matrix which accounts for expression gradients over each neighborhood.
As we have done previously, we'll use the vector `c(18, 30)` for the `k_geom`: 18 neighbors for the neighbor averaging matrix, and 30 neighbors for the `AGF` matrix.

In the chunk below, calculate the `Banksy` matrix on `hvg_spe` using `Banksy::computeBanksy()` (be sure to use `compute_agf=TRUE`!).

```{r banksy matrix, solution = TRUE}
# define k_geom vector

# run computeBanksy with compute_agf = TRUE

```

Next, calculate the PCA from this `Banksy` matrix using `Banksy::runBanksyPCA()`.
Here, specify the arguments `use_agf = TRUE` to incorporate `AGF` and `lambda = 0.2`.

```{r banksy pca, solution = TRUE}
# calculate banksy PCA

```

You should now have a reduced dimension called `PCA_M1_lam0.2` present in the `hvg_spe` object, in addition to the previous one called `PCA` that doesn't consider spatial information.

```{r reduceddims}
reducedDimNames(hvg_spe)
```



Finally, calculate the `Banksy` clusters themselves using `scran::clusterCells()`, again using our existing `cluster_params` object.
Be sure to specify the correct PCA name for this calculation!
Save the resulting cluster assignments to a `colData` column called `banksy_cluster`.

```{r calculate banksy clusters, solution = TRUE}
# Calculate Banksy clusters

```


At this point, we should have both `nonspatial_cluster` and `banksy_cluster` columns in the `colData` slot. 
Lets confirm these are both present by printing the `colData`:

```{r show coldata}
head(colData(hvg_spe))
```

Let's quickly count how many clusters are present in each result, and how many spots are in each.

First, the non-spatial clusters:

```{r table nonspatial clusters}
# make a table of non-spatial cluster counts
table(hvg_spe$nonspatial_cluster)
```

Now, do the same for the `Banksy` clusters:

```{r table banksy clusters}
# make a table of banksy cluster counts
table(hvg_spe$banksy_cluster)
```

When we ran our version of this notebook, we got 9 non-spatial clusters and 12 `Banksy` clusters, but you may get different results depending on the exact sequences of code you ran.
This should be just about as high as we can go with built-in color palettes, although they are not necessarily color-blind friendly.
Along these lines, we should emphasize that it's quite challenging (if at all possible) to achieve a truly distinguishable and accessible color palette when there are this many colors.
But, for the purposes of this exploratory exercise notebook, making these plots should be fine.

In the chunk below, we'll plot the coordinates colored by each set of clusters, using different palettes for each.

```{r make coords plots}
#| fig.width: 8

nonspatial_plot <- ggspavis::plotCoords(
  hvg_spe, 
  annotate = "nonspatial_cluster", 
  point_size = 1
) + 
  labs(title = "Non-spatial clusters")  +
  # specify the Set1 palette which allows up to 9 colors
  scale_color_brewer(palette = "Set1") +
  # wrap legend to be in two columns with larger points
  guides(color = guide_legend(ncol = 2, override.aes = list(size = 2)))

banksy_plot <- ggspavis::plotCoords(
  hvg_spe, 
  annotate = "banksy_cluster", 
  point_size = 1
) + 
  labs(title = "Banksy clusters")  +
  # specify the Paired palette which allows up to 12 colors
  scale_color_brewer(palette = "Paired") +
  # wrap legend to be in two columns with larger points
  guides(color = guide_legend(ncol = 2, override.aes = list(size = 2)))

# add together with patchwork
nonspatial_plot + banksy_plot
```

## Part C: Compare clustering results

There's a few ways we can more precisely pin down how these clustering results compare.
First, we can make make a heatmap visualizing their overlap and how many spots clusters have in common between approaches.
We'll begin this by making a contingency table of the clustering results.
We'll set this up so non-spatial clusters appear in columns, and `Banksy` clusters appear in rows:

```{r make clusters_table}
# define contingency table for comparing clusters
clusters_table <- table(
  colData(hvg_spe)$banksy_cluster,
  colData(hvg_spe)$nonspatial_cluster
)

clusters_table
```

For a heatmap visualization, we would ideally not like to plot these raw counts because the total number of spots differs across clusters.
Therefore, we'll convert these counts into _proportions_, specifically so that proportion corresponds to the total number of spots in the given non-spatial cluster.
After this, each column in our heatmap will end up summing to 1.

```{r cluster proportions}
# normalize by the number of cells in each nonspatial cluster, which appear in columns
clusters_table <- prop.table(clusters_table, 2)
```

Now we can plot the heatmap:

```{r cluster heatmap}
# non-spatial clusters are columns, and Banksy clusters are rows
pheatmap::pheatmap(
  clusters_table,
  # turn off heatmap hierarchical cluster for easier viewing
  cluster_cols = FALSE,
  cluster_rows = FALSE, 
  # show cluster type with the labels
  labels_row = glue::glue("{levels(hvg_spe$banksy_cluster)}_banksy"),
  labels_col = glue::glue("{levels(hvg_spe$nonspatial_cluster)}_nonspatial")
)
```

Which clusters, if any, seem to be consistent between the two approaches?
Are there clusters that seem to be split into smaller "sub-clusters" in one approach versus the other?
What other patterns do you see here, and can you also see these trends reflected in the coordinate plots we made above?


In addition to visual comparisons, we can also compare results using a metric such as the [Adjusted Rand Index](https://bioconductor.org/books/3.22/OSCA.advanced/clustering-redux.html#adjusted-rand-index) (ARI).
ARI measures the similarity of two data clusterings by asking whether spots grouped together in one clustering (here, non-spatial clusters) are still grouped together in a different clustering (here, `Banksy` clusters).
ARI itself is the proportion of pairs of spots which have the same status (e.g., clustered together or separately) in both clusterings being compared.
ARI ranges from -1 to 1, where:

* A value of 1 indicates clusterings are completely overlapping
* A value of -1 indicates clusterings are completely distinct
* A value of 0 indicates a random relationship

We can calculate this value using the function `bluster::pairwiseRand()`, specifying the argument `mode = "index"`.

```{r compute pairwise ari}
# calculate a single ARI value
bluster::pairwiseRand(
  hvg_spe$nonspatial_cluster, 
  hvg_spe$banksy_cluster, 
  mode = "index"
)
```

When we ran this, we got a value of ~0.34, which represents weak-to-moderate agreement: There are definitely spots which are grouped together in both sets of clusters, but there are also a substantial number of differences.
In addition, we cannot expect a perfect value here since the non-spatial and `Banksy` results each have a different number of clusters.

A trend you may have noticed from visual inspection of the coordinate plots and the heatmap we made above is that `Banksy` seems to subdivide non-spatial clusters.
We can quantify this more rigorously by running `bluster::pairwiseRand()` with the argument `mode = "ratio"`. 
This will return a square matrix, with both rows and columns representing the non-spatial clusters (i.e., the first argument we provide to `bluster::pairwiseRand()`).

The values in this matrix represent deviation from what you'd expect under random chance, where 0 matches the expectation under random cluster assignment. 
We can think about random chance here like this: If we take all the spots and randomly shuffle which `Banksy` cluster they land in (preserving the same number of spots per cluster as the `Banksy` clusters actually have), how often would spots from each non-spatial cluster end up together just by luck?
This is what we would expect if `Banksy` clusters were randomly splitting up spots that used to be grouped together. 

This matrix can be quite complex to interpret, so we'll focus on one particular part of it: the diagonal.
Entries along the diagonal help us see if spots in the same non-spatial cluster remain grouped together in `Banksy` clusters:

* Large values indicate that spots in that non-spatial cluster tend to also be clustered together in the `Banksy` clusters
* Values near 0 indicate that spots in that non-spatial cluster tend to fall into different `Banksy` clusters

We therefore expect to see diagonal values near 0 if spots in non-spatial clusters are getting split up in `Banksy` clusters.

In the chunk below, we'll run the `bluster::pairwiseRand()` again but this time specifying `mode = "ratio"`.
Here, we intentionally specify the non-spatial clusters as the _first_ argument, and then the `Banksy` clusters.
This will ensure the output matrix tells if each non-spatial cluster's spots stay grouped together or split apart in the `Banksy` clustering, rather than the other way around.

To focus our interpretation on the diagonal, we'll also pipe the output into `diag()` which (you guessed it!) will just return the diagonal.
To help you interpret this output, we have created a brief [set of Google Slides](https://docs.google.com/presentation/d/1tOput9njIlKwfoyDhZcmsrz4LHIL4xaNxuniDPYaVHs/edit?usp=sharing) that walk through the interpretation of a similar heatmap and ARI matrix diagonal.

```{r compute ari ratio}
# calculate the matrix, but only show the diagonal.
bluster::pairwiseRand(
  hvg_spe$nonspatial_cluster, 
  hvg_spe$banksy_cluster, 
  mode = "ratio"
) |> diag()
```


From this output, identify a non-spatial cluster whose value is fairly close to 0, which implies that those spots are subdivided across `Banksy` clusters.
Do you see this reflected in the heatmap?
Next, consider a different cluster with a higher diagonal value, which suggests `Banksy` may not be splitting that cluster up as much.
Do you see a higher 1:1 correspondence in the heatmap for that cluster as compared to `Banksy` clusters?

Based on the diagonal values here overall, does `Banksy` tend to subdivide clusters?


## Part D: Explore how the `lambda` parameter affects clustering

So far for `Banksy` clustering, we've been using the parameter value of `lambda = 0.2`.
Recall that `lambda` controls the relative weighting given to spatial information versus expression information when constructing the `Banksy` matrix and be any value in the range of 0-1:

* A value of 0 means no spatial weighting is applied at all, so we end up with expression-only clusters
* A value of 1 means spatial information is given its maximum possible weight

Importantly, a `lambda` of 1 does not mean that expression information no longer contributes to clustering at all.
The neighborhood-averaged and `AGF` matrices that `Banksy` creates are themselves calculated using both spatial information and expression values. 
As such, expression always contributes to clustering, and the `lambda` parameter only controls how heavily `Banksy` weights its spatially-informed matrices.


We have so far been using `lambda = 0.2` for our Visium data, which the `Banksy` developers recommend for detecting spatially-aware clusters driven by cell-type identity.
However, choosing the right value for `lambda` should depend on your scientific goals; if, for example, you want your clusters to primarily reflect spatially-cohesive domains rather than cell types, using `lambda = 0.2` may not be suitable for your analysis.

In this section, we'll calculate `Banksy` clusters across a range of `lambda` values to see how our results change: 0 (no spatial weighting), 0.2, 0.4, and 0.8.
From this, we'll develop a sense of how `lambda` affects clustering results, which in turn may guide your choice of `lambda` value for your own analyses.

Note that we'll recalculate `lambda = 0.2` here again since it's quick and keeps our code simple, but bear in mind the result may differ slightly from what we saw in Part B.
This is not unexpected because clustering is stochastic and depends on the state of the random seed.

### Cluster across a range of lambda values

We'll begin by defining a named vector of `lambda` values to iterate over.
We'll name each value with the `colData` column name we'll use to store its resulting clusters, which will make it easy to refer back to specific results later on.

```{r lambdas}
lambdas <- c(0, 0.2, 0.4, 0.8)
names(lambdas) <- glue::glue("banksy_cluster_lam{lambdas}")
```

To calculate `Banksy` clusters for each `lambda` value, we need to first calculate a new `Banksy` PCA (since the PCA itself depends on `lambda`) and then cluster on that PCA, saving each result back to `hvg_spe` as we go.
Normally, we'd use a function like `purrr::map()` to iterate over the `lambdas` vector and create different sets of clusters for each.
However, this won't quite work for us here because we actually want to build up the `hvg_spe` object with both the PCA matrices (which we'll need for some calculations later in the notebook!) and cluster results in `colData` as we iterate. 
With `purrr::map()`, we'd be able to get a list of clusters, but not the `hvg_spe` object updated with each `lambda`'s results.

A different approach could be to use a `for` loop to iterate over `lambda` values and update `hvg_spe` as we go, but instead we're going to opt for something a little more fancy and use the [`purrr::reduce()` function](https://purrr.tidyverse.org/reference/reduce.html).

This function works by collapsing, or _reducing_, a list of values down to a single value.
To do this, `purrr::reduce()` starts with the first item in the provided list and updates it using the provided function. 
It then continues marching down the list items and iteratively applies the provided function.  

Here, we'll take advantage of a special usage of `purrr::reduce()`: rather than having it start with the first item we're iterating over, we can use the `.init` argument to specify a completely different starting object.
By doing this, the object we provide to `.init` gets updated in place at each iteration.
In our case, that starting object will be `hvg_spe` itself. 

Below, we'll use `purrr::reduce()` to iterate over the `lambda` values we provide it and reduce the `hvg_spe` object.
Upon each iteration, it will apply a custom function that we'll write to calculate `Banksy` clusters to the `hvg_spe` object itself.
The SPE object will therefore get iteratively updated with a new `Banksy` PCA and a new set of clusters gets added for each `lambda` value.

```{r cluster across lambda}
hvg_spe <- lambdas |>
  purrr::reduce(
    # Provide two arguments:
    # first, the item that should be reduced aka modified on each iteration - the SPE object itself
    # second, the item we are iterating over - the lambda
    \(spe, lam) {
      spe <- Banksy::runBanksyPCA(
        spe,
        lambda = lam,
        use_agf = TRUE
      )
      pc_name <- glue::glue("PCA_M1_lam{lam}")
      col_name <- glue::glue("banksy_cluster_lam{lam}")

      spe[[col_name]] <- scran::clusterCells(
        spe,
        use.dimred = pc_name,
        BLUSPARAM = cluster_params
      )

      # return the object we are updating in place
      spe
    },
    # initialize with this object so it gets updated on each iteration
    .init = hvg_spe
)
```



Let's use `purrr::map()` to see how many clusters we ended up with for each `lambda` value:

```{r count clusters}
names(lambdas) |>
  purrr::map(
    \(colname) levels(hvg_spe[[colname]])
  )
```

What do you notice about how the number of clusters changes as `lambda` increases?

Now, let's visualize these clusters directly on the tissue coordinates.
Since we'll be plotting four sets of clustering results that each contain quite a few clusters, we'll use the `pals` package to build ourselves a color palette large enough to comfortably distinguish that many groups.

Based on the cluster counts you just printed, define a variable `max_clusters` in the chunk below that stores the largest number of clusters observed across all four `lambda` values.
(We got 16, so the solutions will use that value.)

```{r max_clusters, solution = TRUE}
# define max_clusters

```

Using this variable, we'll define a color palette that we'll share across all four sets of clusters.
Note that reusing one palette across several plots conveying different values isn't generally something we'd recommend, but it's a reasonable tradeoff here given that this is an exploratory notebook and we have quite a few colors to keep track of!
To be able to use these custom colors, we need to assign them names matching the cluster names. 
In our case, that's just the numbers `1:max_clusters` itself. 
Below, we'll define the palette and set its names for plotting.

```{r define palette}
cluster_pal <- pals::alphabet(max_clusters) |>
  setNames(1:max_clusters)
```


Now we're ready to plot.
Conveniently, the names we assigned to `lambdas` earlier match the `colData` column names we want to plot, so we can map directly over them to make all of our plots.

```{r plot coords}
#| fig.width: 10
#| fig.height: 6

names(lambdas) |>
  purrr::map(
    \(colname) {
      ggspavis::plotCoords(
        hvg_spe,
        # color by the column name we're looping over
        annotate = colname,
        point_size = 1
      ) +
      scale_color_manual(values = cluster_pal) +
      guides(color = guide_legend(nrow = 3, override.aes = list(size = 2)))
  }
) |>
  patchwork::wrap_plots()
```

What observations can you make about the spatial cohesiveness of these clusters as `lambda` increases?
Do they start to look more like contiguous spatial regions, or do they become more scattered across the tissue?
What seems to be the overall effect of increasing `lambda` on the number of clusters and their layout?

### Compare clustering results at lambda 0 and lambda 0.8

Let's take a closer, more quantitative look at how `lambda` affects clustering by directly comparing two of these results, specifically clusters for `lambda = 0` and `lambda = 0.8`.
As we did in Part C, we'll compare these two sets of clusters with a heatmap, the overall ARI index, and the diagonal of the ARI ratio matrix.

In the chunk below, build a contingency table comparing these two sets of clusters, with `banksy_cluster_lam0.8` as the first argument (which will appear in rows) and `banksy_cluster_lam0` as the second argument (which will appear in columns).

```{r make heatmap table, solution = TRUE}
#| fig.width: 7
#| fig.height: 5

# Create a table comparing clusters

```

Next, normalize the table so that the columns (the `lambda = 0` clusters) sum to 1. 
In other words, each value will represent the proportion of spots in a given `lambda = 0.8` cluster (the rows) that came from each `lambda = 0` cluster (the columns).

```{r normalize heatmap table, solution = TRUE}
# Normalize by the number of cells in each non-spatial cluster

```

Finally, make the heatmap:

```{r}
#| fig.width: 7
#| fig.height: 5

# Plot heatmap
pheatmap::pheatmap(
  clusters_table,
  cluster_cols = FALSE,
  cluster_rows = FALSE,
  labels_col = glue::glue("{levels(hvg_spe$banksy_cluster_lam0)}_lam0"),
  labels_row = glue::glue("{levels(hvg_spe$banksy_cluster_lam0.8)}_lam0.8")
)
```

Can you find examples of clusters that have a close mapping between the two `lambda` values, i.e. are they mostly made up of spots from a single corresponding cluster?
Do you see any clusters that instead look like a blend of multiple `lambda = 0` clusters?

Next, calculate the overall ARI index comparing these two sets of clusters, just as we did in Part C.

```{r compute ari index lambdas, solution = TRUE}
# calculate the matrix, but only show the diagonal.

```

Where does this value fall within the -1 to 1 range?
How does it compare to the ARI value of ~0.34 we calculated between the non-spatial and `Banksy` (`lambda = 0.2`) clusters back in Part C?

Now, calculate the diagonal of the ARI ratio matrix below, specifying `hvg_spe$banksy_cluster_lam0` as the first argument.
Recall from Part C that this ordering determines what the diagonal actually tells us: here, it will show us whether spots grouped together in each `lambda = 0` cluster remained grouped together when we increased `lambda` to 0.8, or whether they were split apart into multiple clusters.

```{r compute ari ratio lambdas, solution = TRUE}
# calculate the matrix, but only show the diagonal.

```

How does this compare to the heatmap above?
Based on these diagonal values, is it more common for spots grouped together at `lambda = 0` to stay grouped together at `lambda = 0.8`, or to get split apart?


In the chunk below, you can feel free to repeat this comparison between two other `lambda` values to see how the results differ.
In the solution, we've compared `lambda = 0.4` and `lambda = 0.8`, which is a particularly interesting pair to look at because these two results actually contain the same number of clusters, but there may be some differences between them yet!

```{r optional diy heatmap, solution = TRUE}
#| fig.width: 7
#| fig.height: 5

# Create a table comparing clusters

# Normalize

# Plot heatmap

```


```{r optional diy ari index, solution = TRUE}
# calculate the ARI using mode = "index"

```


```{r optional diy ari matrix, solution = TRUE}
# now calculate the ARI using mode = "ratio"

```

## Part E: Calculate cluster stability across `lambda` values

We've now seen how `lambda` affects which spots end up grouped together within clusters, but what about the reliability of the clusters themselves?
Are the clusters we get with some `lambda` values more stable than others?

One way to explore this is to measure cluster _stability_ using a bootstrap approach.
For each set of clusters, we would:

* Sample PCA rows (i.e., spots) with replacement to create a bootstrapped PCA matrix
* Perform clustering on this bootstrapped PCA matrix
* Compare the bootstrapped clusters to the original clusters, using a metric such as the pairwise ARI index 

If clusters are stable, we expect spots that were clustered together in the original clustering to also end up clustered together across bootstrap replicates; in other words, we'd expect ARI values close to 1.
Repeating this across each of our `lambda` values tells us something specific about the underlying PCA matrices themselves: given the same clustering parameters, how well does each matrix's cluster structure hold up under resampling?

We'll use the function [`bluster::bootstrapStability()`](https://rdrr.io/github/LTLA/bluster/man/bootstrapStability.html) to calculate this for us.
This function takes the following arguments:

* The PCA matrix the clusters were built from
* A vector of the corresponding cluster assignments, specified with the argument `clusters`
* `BLUSPARAM`: A specification of clustering parameters to use for each bootstrap iteration.
These should be the same parameters we used to calculate the original clusters - and conveniently, we've already defined these as `cluster_params`!
* `mode`: Which ARI mode to calculate.
Since we want a single overall stability value for each `lambda`, we'll use `mode = "index"`, which reports the mean ARI across all bootstrap replicates.
* `iterations`: The number of bootstrap replicates to perform (by default, 20)

In the chunk below, we'll calculate stability for each of our four `lambda` values, setting `mode = "index"` and `iterations = 100`.
For these calculations, we need to provide both the PCA matrix and the corresponding clusters for each `lambda`.
For this, we'll use a special variant of `purrr::map()` called [`purrr::imap()`](https://purrr.tidyverse.org/reference/imap.html).
This function iterates simultaneously over both a vector/list and its names: here, we can iterate over the `lambda` vector and its column names, which match the `colData` column names that store those clusters.
This may take a minute or two to run!

```{r calculate stability}
lambdas |>
  purrr::imap(
    # iterate jointly over the lambda value and its name in the lambda vector, 
    # which conveniently matches the colData column name that stores those clusters
    \(lam, colname) {
      pc_name <- glue::glue("PCA_M1_lam{lam}")

      bluster::bootstrapStability(
        # specify the correct PCA matrix and clusters
        reducedDim(hvg_spe, pc_name),
        clusters = hvg_spe[[colname]],
        BLUSPARAM = cluster_params,
        # Perform 100 bootstrap iterations
        iterations = 100,
        # Calculate the mean ARI
        mode = "index"
      )
    }
  )
```

How does `lambda` affect cluster stability in this dataset?
In other words, what is the effect of increasing the weight given to spatial information?

## Conclusions

In this notebook, we've developed a sense of how `lambda` influences `Banksy` clusters on Visium data - both in terms of how spots get grouped together, and how stable those groupings are.
So, which `lambda` value is the "right" one?
...Trick question! 
Stability can help you decide which `lambda` values are more reproducible, but it isn't the only factor to consider. 
A key consideration is your scientific question and what you want to learn from your clusters:

* If you're more interested in comparing distinct spatial regions or domains within the tissue, you might lean toward a higher `lambda`.
* If you're more interested in comparing cell types, even when they're spread across different regions, you might lean toward a lower `lambda`, closer to (or even at) expression-only clustering.

Comparing results across a range of `lambda` values as we did here can also help you build intuition for how your clusters change as spatial information is incorporated in your own datasets.

## Export

To wrap up, we'll export the `hvg_spe` object, which now include all of the clusters we calculated here, to the `crc_output` path we defined back at the beginning.

```{r export, solution = TRUE}

```


## Session Info

```{r sessioninfo}
sessionInfo()
```

