---
title: "Identifying spatially variable genes"
author: Data Lab for ALSF
date: "2026"
output:
  html_notebook:
    toc: true
    toc_float: true
---

## Objectives

- Introduce the `SpatialFeatureExperiment` and `Voyager` packages
- Demonstrate two methods for identifying spatially variable genes

## Introduction

Coming Soon

## Set up

First we will load the `SpatialExperiment` package, as we have done before, as well as some familiar packages for plotting.

We'll also set a random seed for reproducibility.

### Libraries

```{r libraries}
# The main class we use for spatial transcriptomic data
suppressPackageStartupMessages({
  library(SpatialExperiment)
})

library(ggplot2)
library(patchwork)

# Set default ggplot2 theme
theme_set(theme_bw())
```


```{r seed}
set.seed(2026)
```

### Files

Wilms tumor data

```{r inputs}
# define sample id we'll be analyzing
sample_id <- "SCPCS000190"

# data directory for this sample
data_dir <- file.path("data/wilms-tumor", sample_id, "normalized")

normalized_rds <- file.path(data_dir, "SCPCS000190_normalized.rds")
```


### Import

Let's start by reading in the pre-processed SPE object:

```{r import-rds}
spe <- readr::read_rds(normalized_rds)
```

Let's make sure the sample id is set correctly for later import:
```{r}
spe$sample_id <- sample_id
imgData(spe)$sample_id <- sample_id
```



Last time we calculated hvgs first, but this time let's convert first, just to show that everything with spe works with sfe as well. 



```{r sfe-conversion, live = TRUE}
sfe <- SpatialFeatureExperiment::toSpatialFeatureExperiment(spe) |>
  # flip the coordinates to match how we have been looking at them
  SpatialFeatureExperiment::mirror() |>
  # we have to flip the image again to match: it was upside down
  SpatialFeatureExperiment::mirrorImg()
```


## Identifying Highly Variable Genes

For the first part of this notebook, we are going to use the scran package to identify highly variable genes, just as we have done before.
Many of the calculations we will be doing for spatial analysis are not particularly fast, so this will help us speed up calculations.
But more practically, if a gene is not particularly variable overall, it is quite unlikely to be spatially variable, and we are not likely to get much signal from it.

```{r find-hvg}
# model gene variance
gene_variance <- scran::modelGeneVar(sfe)

# calculate the top 2000 HVGs
hvg_vector <- scran::getTopHVGs(gene_variance, n = 2000)
```

```{r filter-spe, live = TRUE}
# filter the SPE to only include the highly variable genes
sfe <- sfe[hvg_vector, ]
```


Let's look at the contents of the SFE:

```{r SFE-contents, live = TRUE}
sfe
```

The first part looks pretty much the same as the SPE object, but you will notice a few extra slots in the summary:

- `unit`
- `Geometries`
- `colGeometries`
- `Graphs`

For now most of these are empty.

One thing you will see here is that the `SpatialFeatureExperiment` package is built from the start to accommodate multiple samples, so the graphs will be divided up by sample.
We only have one, but we will see some consequences of this as we move forward.

The first thing we will do is to define contents for the Graphs

The next step is to define the graph that defines how cells are connected to each other.
For Visium data, we can use `SpatialFeatureExperiment::findVisiumGraph()` which uses the known barcodes and their locations to define a graph that connects nearby spots.
This is a nice shortcut, but for other kinds of spatial data (such as image-based Xenium data), we would have to use a function like `SpatialFeatureExperiment::findKNNGraph()` to find the nearest neighbors for each cell, and then define the graph based on that.

We will store that in the `colGraph` slot of the SFE object (this is one of the things we get from that package)

```{r define-graph}
#| error: true

# use the Visium barcodes to define the graph
SpatialFeatureExperiment::colGraph(
  sfe
) <- SpatialFeatureExperiment::findVisiumGraph(sfe)
```

We get an error here because there are some isolated points in the graph, which is not allowed.
We could ignore it by setting `zero.policy = TRUE`, but then we would have to deal with those points downstream as well, so it is probably better to just dropt them.

Instead we will use a feature to find debris cells.
The distance cutoff here matters a lot, 300 seems to work well for this dataset with the resolution here, it does depend a lot on the units!

```{r}
sfe <- SpatialFeatureExperiment::findDebrisCells(sfe, distance_cutoff = 300)
```


```{r}
Voyager::plotSpatialFeature(
  sfe,
  size = 1.5,
  features = "is_debris",
  image_id = "lowres"
)
```


```{r}
sfe <- sfe[, !sfe$is_debris]
```


Now we will make the colGraph again: 
```{r}
SpatialFeatureExperiment::colGraph(
  sfe
) <- SpatialFeatureExperiment::findVisiumGraph(sfe)
```

Warnings about subgraphs are probably fine.



```{r plot-graph, live = TRUE}
Voyager::plotColGraph(sfe) +
  # the units are arbitrary, so let's leave them and the grid lines off
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    panel.grid = element_blank()
  )
```


Possibly: include SpatialFeatureExperiment::findSpatialNeighbors()?


## Identify Spatially Variable Genes

### Moran's I


```{r moranI, live = TRUE}
sfe <- Voyager::runMoransI(sfe)
```

This gives us an I value for each gene as well as the kurtosis of the gene's expression (how fat the tails are) as a value `K`.
TODO: add better kurtosis explanation (what does high or low mean?)

```{r view_moran}
rowData(sfe)
```

You'll notice that the new column names include the `sample_id`, which is a bit annoying, but it is necessary to keep things separate if we have multiple samples.
We'll make sure to set that as a variable so we can reuse it as needed.

Let's first look at the distribution of Moran's I values across all (highly variable) genes:


```{r moran-hist}
rowData(sfe) |>
  as.data.frame() |>
  ggplot(aes(x = .data[[paste0("moran_", sample_id)]])) +
  geom_histogram(bins = 50) +
  labs(
    x = "Moran's I",
    y = "Number of genes"
  )
```


We get a value there, but no measure of significance.

We can calculate those using the more general `Voyager::runUnivariate()` function, which can do other types of tests as well
For that we can use either `moran.test()` or `moran.mc()` from the `spdep` package.
TODO: clarify where these functions are coming from

`moran.test` is relatively fast, so we can do that first (though it is not likely to be as accurate, as it does not account for the structure of the graph).


```{r moran-test, live = TRUE}
sfe <- Voyager::runUnivariate(
  sfe,
  type = "moran.test",
  BPPARAM = BiocParallel::MulticoreParam(4)
)

rowData(sfe)
```


```{r view-mc-result}
rowData(sfe)
```


Let's apply a multiple testing correction to the P values, and assign significance for each at an FDR of 0.05.

```{r moran-significant}
moran_df <- rowData(sfe) |>
  as.data.frame() |>
  dplyr::select(
    ID,
    moran_i = paste0("moran_", sample_id),
    moran_test_p = paste0("moran.test_p.value_", sample_id)
    
  ) |>
  dplyr::mutate(
    moran_test_neglogp = -log10(moran_test_p),
    moran_test_p_adjust = p.adjust(moran_test_p, method = "BH"),
    moran_test_sig = moran_test_p_adjust < 0.05
  )
```

```{r count-moran-significance}
moran_df |>
  dplyr::count(moran_test_sig)
```

Either way, we get the vast majority of genes as "significant" which is not too surprising.
Using significance as a filter here is not likely to be very useful.


Let's make a plot of the top few genes:

```{r get-top-moran, live = TRUE}
top_moran <- moran_df |>
  dplyr::slice_max(moran_i, n = 6) |>
  dplyr::pull(ID)
```


```{r plot-top-moran, live = TRUE}
Voyager::plotSpatialFeature(
  sfe,
  features = top_moran,
  ncol = 2,
  # use gene symbols instead of ensembl IDs for the plot
  swap_rownames = "Symbol"
)
```

### `nnSVG`

Let's try `nnSVG` now, since that is more specifically designed for this kind of data.

TODO: Add some more context about the method

The main advantage that it has is that it can model the spatial variation in a more flexible way, allowing for different scales of spatial variation.

We could perform this analysis on all genes, but again for now we will restrict ourselves to HVGs for time.

Note that this function works either on an SFE or an SPE object, but we will use the SFE object here since we already have it.

```{r run-nnsvg, live = TRUE}
sfe <- nnSVG::nnSVG(sfe, n_threads = 12)
```

Pause here for a break.
This is much slower, but we get significance calculations by default.

```{r count-nnsvg-significance, live = TRUE}
rowData(sfe) |>
  as.data.frame() |>
  dplyr::filter(padj < 0.05) |>
  nrow()
```

Again, most of them!
So most genes are significantly spatially variable, which is perhaps not a huge surprise.

Other stats we can look at include `prop_sv`, which is the proportion of variation that is attributable to spatial variation.

How do the P values compare across methods?

```{r collate-results}
svg_df <- rowData(sfe) |>
  as.data.frame() |>
  dplyr::select(
    ID,
    nnsvg_stat = LR_stat,
    nnsvg_propsv = prop_sv,
    nnsvg_p = pval,
    nnsvg_p_adjust = padj
  ) |>
  dplyr::mutate(
    nnsvg_neglogp = -log10(nnsvg_p)
  ) |>
  dplyr::left_join(
    moran_df,
    by = "ID"
  )
```


```{r plot-moran-nnsvg-p}
ggplot(svg_df, aes(x = moran_test_neglogp, y = nnsvg_neglogp)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I -log10(P) (moran.test)",
    y = "nnSVG -log10(P)"
  ) +
  xlim(0, 5)
```

Rather than comparing just the P values, we could also look at the correlation between I and `nnsvg`'s `prop_sv` statistic.

```{r plot-moran-nnsvg-stats}
ggplot(svg_df, aes(x = moran_i, y = nnsvg_propsv)) +
  geom_point(alpha = 0.4) +
  labs(
    x = "Moran's I",
    y = "Proportion of variance explained by spatial location"
  ) +
  annotate(
    geom = "label",
    x = 0.5,
    y = 0,
    label = paste0(
      "Spearman correlation: ",
      round(cor(svg_df$moran_i, svg_df$nnsvg_propsv, method = "spearman"), 2)
    )
  )
```

So for a lot of genes we get very similar scores, but `nnSVG` is uncovering some genes as spatially variable that Moran's I does not, presumably because it models the spatial variation in a more flexible way.

Let's look at the top few:

```{r}
top_nnsvg <- svg_df |>
  dplyr::slice_max(nnsvg_propsv, n = 6) |>
  dplyr::pull(ID)

Voyager::plotSpatialFeature(
  sfe,
  features = top_nnsvg,
  ncol = 2,
  swap_rownames = "Symbol"
)
```

This does seem to detect more punctate expression patterns, where expression of some genes is found in specific distinct spots rather than varying across larger scales.

## Looking within Regions

You may have notice that there seems to be two main areas of this slide: A top region and a lower region.
As discussed in the first instruction, these may be the blastemal and stromal regions of a tumor. 

Let's do some quick clustering to see if we can detect that more broadly. We'll use BANKSY


```{r}

sfe <- Banksy::computeBanksy(sfe, assay_name = "logcounts", k_geom = 18)
```
```{r}
sfe <- Banksy::runBanksyPCA(sfe)
```

really simple clustering: only 2 groups

```{r}
sfe <- Banksy::clusterBanksy(sfe, algo = "kmeans", kmeans.centers = 2)
```

```{r}
Voyager::plotSpatialFeature(
  sfe,
  features = "clust_M0_lam0.2_kmeans2",
  image_id = "lowres",
  size = 1.5
)
```

The clustering here is very preliminary, but let's see if we can use it to look at the spatially variable genes within each of these regions.

Calculate number of spots per cluster

```{r}
table(sfe$clust_M0_lam0.2_kmeans2)
```


Let's take the smaller cluster and see if we can detect spatially variable genes within this group of cells.

```{r}
smallclust <- names(which.min(table(sfe$clust_M0_lam0.2_kmeans2)))
# get the list of spots in this cluster
smallclust_spots <- colnames(sfe)[sfe$clust_M0_lam0.2_kmeans2 == smallclust]
```


```{r}
small_spe <- spe[, smallclust_spots]
```

This fails because of the zero.policy issue, so we temporarilty convert to SPE to subset, then back to SFE:
```{r}
small_sfe <- SpatialFeatureExperiment::toSpatialFeatureExperiment(small_spe) |>
  SpatialFeatureExperiment::mirror() |>
  SpatialFeatureExperiment::mirrorImg()

```


re-find HVGs
```{r}
small_gene_variance <- scran::modelGeneVar(small_sfe)
small_hvg_vector <- scran::getTopHVGs(small_gene_variance, n = 2000)

small_sfe <- small_sfe[small_hvg_vector, ]

```

```{r}
SpatialFeatureExperiment::colGraph(
  small_sfe
) <- SpatialFeatureExperiment::findVisiumGraph(small_sfe)
```



```{r}
Voyager::plotColGraph(small_sfe)
```



Remove "debris", for which we will also include isolated spots that are a bit bigger.

There is some kind of bug if you do this directly on the SFE after subsetting, so we will pull out the coordinates directly for this calculation:


```{r}
small_debris  <- SpatialFeatureExperiment::findDebrisCells(
  SpatialFeatureExperiment::spatialCoords(small_sfe),
  distance_cutoff = 300,
  max_cells = 20
)

small_sfe <- small_sfe[, !small_debris]
```
 
Plot the graph for this smaller cluster

```{r}
Voyager::plotColGraph(small_sfe)
```

We'll run Moran's I and nnsvg on this SFE

```{r}
small_sfe <- Voyager::runMoransI(small_sfe)
small_sfe <- Voyager::runUnivariate(
  small_sfe,
  type = "moran.test",
  BPPARAM = BiocParallel::MulticoreParam(12)
)
```

```{r}
small_sfe <- nnSVG::nnSVG(small_sfe, n_threads = 12)
```

Pull out the results and compare to the full slide results:
```{r}
small_svg_df <- rowData(small_sfe) |>
  as.data.frame() |>
  dplyr::select(
    ID,
    moran_i = paste0("moran_", sample_id),
    moran_test_p = paste0("moran.test_p.value_", sample_id),
    nnsvg_stat = LR_stat,
    nnsvg_propsv = prop_sv,
    nnsvg_p = pval,
    nnsvg_p_adjust = padj
  ) |>
  dplyr::mutate(
    moran_test_neglogp = -log10(moran_test_p),
    moran_test_p_adjust = p.adjust(moran_test_p, method = "BH"),
    moran_test_sig = moran_test_p_adjust < 0.05,
    nnsvg_neglogp = -log10(nnsvg_p),
    nnsvg_sig = nnsvg_p_adjust < 0.05
  )
```

```{r}
small_svg_df |>
  dplyr::count(moran_test_sig, nnsvg_sig)
```
We still see a large number of significant genes, but the number is certainly smaller than for the full slide. 
Some of this is probably just power though.


```{r}
small_top_moran <- small_svg_df |>
  dplyr::slice_max(moran_test_neglogp, n = 6) |>
  dplyr::pull(ID)

plot <- Voyager::plotSpatialFeature(
  small_sfe,
  features = small_top_moran,
  ncol = 2,
  swap_rownames = "Symbol",
) 

```
  

```{r}
plot

```




## Session info

```{r}
sessionInfo()
```

