---
title: "Transform OpenPBTA data"
output:
  html_notebook:
    toc: true
    toc_float: true
author: CCDL for ALSF
date: 2020
---

Here, we'll transform the OpenPBTA stranded RNA-seq data for use in the machine learning module.

## Set up

### Library

```{r}
# For transformation
library(DESeq2)
```

### Directories and files

```{r}
# Set up the directories - we'll be processing OpenPBTA data, which we'll
# stick in a separate output directory
openpbta_dir <- file.path("..", "data", "open-pbta")
input_dir <- file.path(openpbta_dir, "download")
output_dir <- file.path(openpbta_dir, "processed")

# Create the output directory if it doesn't exist yet
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}
```

#### Input files

```{r}
# This file contains all the clinical information like disease labels
histologies_file <- file.path(input_dir,
                              "pbta-histologies.tsv")

# RSEM expected counts
stranded_counts_file <- file.path(
  input_dir,
  "pbta-gene-counts-rsem-expected_count.stranded.rds"
)
```

#### Output files

```{r}
# We're going to write out a histologies file that only includes data for the
# stranded RNA-seq samples
filtered_histologies_file <- file.path(output_dir,
                                      "pbta-histologies-stranded-rnaseq.tsv")
# VST transformed values (big file, so we will compress it)
rnaseq_transformed_file <- file.path(output_dir, "pbta-vst-stranded.tsv.gz")
```

## Data cleaning and transformation

### Clinical/metadata file

```{r}
histologies_df <- readr::read_tsv(histologies_file,
                                  guess_max = 10000) |>
  # We'll only be looking at the stranded RNA-seq dataset, so filter out
  # all other samples (e.g., WGS)
  dplyr::filter(experimental_strategy == "RNA-Seq",
                RNA_library == "stranded") |>
  # Remove columns that are all NA (typically pertain only to the DNA data)
  purrr::discard(~ all(is.na(.)))

# Write cleaned metadata file to the output dir
readr::write_tsv(histologies_df, filtered_histologies_file)
```

### RNA-seq data

```{r}
# Technically a data.frame, not a matrix
stranded_count_mat <- readr::read_rds(stranded_counts_file)
```

Reorder the stranded counts matrix such that it is in the same order as the histologies data frame.
The sample identifiers used in the counts matrix are called `Kids_First_Biospecimen_ID` in `histologies_df`.
We'll sometimes refer to them as biospecimen IDs or BSIDs for short!

```{r}
stranded_count_mat <- stranded_count_mat |>
  # We want to retain the Ensembl gene identifiers as the first column
  dplyr::select(gene_id, histologies_df$Kids_First_Biospecimen_ID) |>
  # Make the gene identifier column the rownames and make this into a matrix
  tibble::column_to_rownames("gene_id") |>
  as.matrix()

# This is the output of RSEM - some values will not be integers, rounding this
# makes it such that these values are integers
stranded_count_mat <- round(stranded_count_mat)
```

```{r}
# In the same order?
identical(colnames(stranded_count_mat),
          histologies_df$Kids_First_Biospecimen_ID)
```

We're going to use this for clustering, so we probably want to be blind to the experimental design.

```{r}
ddset <- DESeqDataSetFromMatrix(countData = stranded_count_mat,
                                colData = histologies_df,
                                design = ~ 1)
```

Remove genes with low total counts.

```{r}
genes_to_keep <- rowSums(counts(ddset)) >= 10
ddset <- ddset[genes_to_keep, ]
```

Variance stabilizing transformation and write to file.

```{r}
# Transformation itself
vst_data <- vst(ddset, blind = TRUE)

# Now create a data frame where the gene identifiers are in the first column
vst_df <- data.frame(assay(vst_data)) |>
  tibble::rownames_to_column("gene_id")

# Write to TSV!
readr::write_tsv(vst_df, file = rnaseq_transformed_file)
```

## Session Info

```{r}
sessionInfo()
```

