---
title: "Prepare CRC Visium v1 dataset"
author: Stephanie Spielman for CCDL
date: 2026
output: html_notebook
params:
  filtered_matrix: "../../data/crc_v1/spaceranger/filtered_feature_bc_matrix"
  spatial_dir: "../../data/crc_v1/spaceranger/spatial"
  output_rds: "../../data/crc_v1/normalized/crc_normalized_spe.rds"
  mito_genes_file: "../../data/reference/hs_mitochondrial_genes.tsv"
---

This notebook prepares this [CRC sample from 10x](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8478586), which was processed with `Space ranger v1` for analysis in the workshop by performing filtering and normalization.
It is meant to be run from the [snakemake workflow](Snakemake) in this directory.

Note that this notebook runs only deterministic algorithms, so no seed is set.

```{r}
suppressPackageStartupMessages({
  library(SpatialExperiment)
})
```

First, check paths:

```{r}
stopifnot(
  "filtered_feature_bc_matrix/ doesn't exist" = dir.exists(params$filtered_matrix),
  "spatial/ doesn't exist" = dir.exists(params$spatial_dir),
  "mito_genes_file doesn't exist" = file.exists(params$mito_genes_file),
  "output_rds must end in rds" = stringr::str_ends(params$output_rds, "\\.rds")
)
```

Begin by importing the SPE.
At this time of writing, `VisiumIO` does not support loading images into memory; it only stores the full path to the image resources.
Therefore, we'll construct the SPE from scratch.

```{r}
# first read in counts
# don't include column names; those will come in from the coords_table
sce <- DropletUtils::read10xCounts(
  params$filtered_matrix
)

# read in image data with load = TRUE, which is default but we'll specify to be safe
# this function can only accommodate: "lowres", "hires", "detected", "aligned"
img_hires <- SpatialExperiment::readImgData(
    path = params$spatial_dir,
    sample_id = "CRC",  # this argument is required
    imageSources = file.path(params$spatial_dir, "tissue_hires_image.png"),
    scaleFactors = file.path(params$spatial_dir, "scalefactors_json.json"),
    load = TRUE
)

# read in spatial coordinates
coords_table <- readr::read_csv(
  file.path(
    params$spatial_dir,
    "tissue_positions_list.csv"
  ), 
  # v1 spaceranger version of this file has no header, we need to add it
  col_names = c("barcode", "in_tissue", "array_row", "array_col", "pxl_row_in_fullres", "pxl_col_in_fullres")
) |>
  dplyr::filter(
    # only keep relevant spots in the tissue
    in_tissue == 1,
    barcode %in% sce$Barcode
  )
# confirm order matches
coords_table <- coords_table[match(sce$Barcode, coords_table$barcode), ]
stopifnot(
  "Barcode mismatch when making SPE" = all(sce$Barcode == coords_table$barcode)
)

# combine
spe <- SpatialExperiment(
  assays = list(counts = assay(sce)),
  rowData = rowData(sce),
  colData = DataFrame(coords_table, row.names = coords_table$barcode),
  spatialCoordsNames = c("pxl_col_in_fullres", "pxl_row_in_fullres"),
  imgData = img_hires,
  sample_id = "CRC" # match the image sample_id
)

spe
```

Now, perform filtering:

```{r}
mito_genes <- readr::read_tsv(params$mito_genes_file) |>
  dplyr::filter(gene_id %in% rownames(spe)) |>
  dplyr::pull(gene_id)
mito_genes
```

Confirm that the data has mito genes:

```{r}
all(mito_genes %in% rownames(spe)) # TRUE
```

Looks good, we can include this filtering.


```{r}
spe <- spe |>
  # calculate per-spot statistics
  scuttle::addPerCellQC(
    subsets = list(mito = mito_genes)
  ) |> 
  # Detect local outliers using defaults 
  SpotSweeper::localOutliers(
    metric = "sum",
    direction = "lower"
  ) |> 
  SpotSweeper::localOutliers(
    metric = "detected",
    direction = "lower"
  ) |>
  SpotSweeper::localOutliers(
    metric = "subsets_mito_percent",
    direction = "higher", 
    log = FALSE
  ) 
```

There are this many outlier spots to remove:

```{r}
# identify and remove local outliers
spe$local_outliers <- spe$sum_outliers | spe$detected_outliers | spe$subsets_mito_percent_outliers
print(sum(spe$local_outliers)) # 24 spots
```

```{r}
# keep only spots where `local_outliers` is FALSE
spe <- spe[, !spe$local_outliers]
```


Now, we can normalize the object.

```{r}
spe <- spe |>
  scuttle::computeLibraryFactors() |>
  scuttle::logNormCounts()
```

Finally, we can export with compression.

```{r}
readr::write_rds(spe, params$output_rds, compress = "gz")
```

## Session info

```{r}
sessionInfo()
```
