---
title: "Prepare medulloblastoma PDX DGE results"
output:
  html_notebook:
    toc: true
    toc_float: true
author: CCDL for ALSF
date: 2021
---

This notebook prepares DGE results for the ORA and GSEA pathway analysis exercises.

## Set up

### Libraries

```{r}
# DGE analysis and transformation
library(DESeq2)
```

### Directories and files

```{r}
input_dir <- "/shared/data/training-modules/RNA-seq/data/medulloblastoma"
txi_file <- file.path(input_dir, "txi", "medulloblastoma_txi.RDS")
```

```{r}
output_dir <- "/shared/data/training-modules/pathway-analysis/data/medulloblastoma"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dge_file <- file.path(output_dir, "pdx_medulloblastoma_treatment_dge.tsv")
```

## Read in data

```{r}
txi <- readr::read_rds(txi_file)
```

### Dataset creation

```{r ddset-creation}
# We have Group 3 and Group 4 MB PDX that were treated with digoxin (or control)
# We're going to use the treatment results for the exercise, so we put that
# *last*
ddset <- DESeq2::DESeqDataSet(txi,
                              design = ~ Group + Treatment)
```

We need to collapse technical replicates.

```{r}
ddset <- DESeq2::collapseReplicates(ddset,
                                    groupby = ddset$mouse_id)
```

We want the reference to be the untreated/control samples, check levels to make sure that's the case.

```{r}
levels(ddset$Treatment)
```

We need to relevel.

```{r}
ddset$Treatment <- relevel(ddset$Treatment, ref = "none")
```

## DGE

```{r}
ddset <- DESeq(ddset)
```
```{r}
resultsNames(ddset)
```

We used the design `~ Group + Treatment` when creating the `DESeqDataSet`.
The results for `Treatment` in our example should be the differences between digoxin vs. none, controlling for differences due to molecular subtype (`Group`).

```{r}
results_lfc <- lfcShrink(ddset,
                         coef = "Treatment_digoxin_vs_none",
                         type = "ashr")
```

```{r}
results_df <- results_lfc |>
  data.frame() |>
  tibble::rownames_to_column("Gene") |>
  # We didn't do any prefiltering to this data set, so there are genes
  # with 0 counts
  dplyr::filter(!is.na(log2FoldChange))

readr::write_tsv(results_df, file = dge_file)
```

## Session Info

```{r}
sessionInfo()
```

