---
title: "Prepare osteosarcoma dataset"
author: Stephanie Spielman for CCDL
date: 2026
output: html_notebook
params:
  filtered_matrix: "/shared/data/training-modules/spatial/data/osteo/GSM8478586/outs/filtered_feature_bc_matrix"
  spatial_dir: "/shared/data/training-modules/spatial/data/osteo/GSM8478586/outs/spatial"
  output_rds: "/shared/data/training-modules/spatial/data/osteo/GSM8478586/normalized/osteo_normalized_spe.rds"
  mito_genes_file: "/shared/data/training-modules/spatial/data/reference/hs_mitochondrial_genes.tsv"
---

This notebook prepares the osteosarcoma sample from [`GSM8478586`](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8478586) for analysis in the workshop by performing filtering and normalization.
It is meant to be run from the [snakemake workflow](Snakemake) in this directory.

Note that this notebook runs only deterministic algorithms.

First, check paths:

```{r}
stopifnot(
  "filtered_feature_bc_matrix/ doesn't exist" = dir.exists(params$filtered_matrix),
  "spatial/ doesn't exist" = dir.exists(params$spatial_dir),
  "mito_genes_file doesn't exist" = file.exists(params$mito_genes_file),
  "output_rds must end in rds" = stringr::str_ends(params$output_rds, "\\.rds")
)
```

Begin by importing the SPE:

```{r}
spe <- VisiumIO::import(
  VisiumIO::TENxVisium(
    resources = params$filtered_matrix,
    spatialResource = params$spatial_dir,
    images = c("hires", "cytassist")
  )
)
spe
```

Now, perform filtering:

```{r}
mito_genes <- readr::read_tsv(params$mito_genes_file) |>
  dplyr::filter(gene_id %in% rownames(spe)) |>
  dplyr::pull(gene_id)
```


```{r}
spe <- spe |>
  # calculate per-spot statistics
  scuttle::addPerCellQC(
    subsets = list(mito = mito_genes)
  ) |> 
  # Detect local outliers using defaults 
  SpotSweeper::localOutliers(
    metric = "sum",
    direction = "lower"
  ) |> 
  SpotSweeper::localOutliers(
    metric = "detected",
    direction = "lower"
  ) |> 
  SpotSweeper::localOutliers(
    metric = "subsets_mito_percent",
    direction = "higher", # we want to remove HIGHER values
    log = FALSE # don't log
)
```

There are this many outlier spots to remove:

```{r}
# identify and remove local outliers
spe$local_outliers <- spe$sum_outliers |
  spe$detected_outliers |
  spe$subsets_mito_percent_outliers
print(sum(spe$local_outliers))
```

```{r}
# keep only spots where `local_outliers` is FALSE
spe <- spe[, !spe$local_outliers]
```


Now, we can normalize the object.

```{r}
spe <- spe |>
  scuttle::computeLibraryFactors() |>
  scuttle::logNormCounts()
```

Finally, we can export with compression.

```{r}
readr::write_rds(spe, params$output_rds, compress = "gz")

```

## Session info

```{r}
sessionInfo()
```